### 2016-06-22 2016_G3_PackageVQTL_Corty.tex
just getting started
able to compile

\documentclass[9pt,twocolumn,twoside]{gsag3jnl}

\articletype{inv} % article type

\usepackage{bm}
\usepackage{tabu}
\usepackage{subfigmat}
\usepackage{graphicx}

\input{styles/Corty2015commands}

\title{\texttt{vqtl}: An \texttt{R} package for Efficient QTL Mapping on Phenotypes with Heterogeneous Variance}

\author[$\ast$]{Robert W. Corty}
\author[$\ast, 1$]{William Valdar}
\affil[$\ast$]{Department of Genetics, University of North Carolina at Chapel Hill}

\keywords{QTL mapping, variance heterogeneity}
\runningtitle{R package vqtl} % For use in the footer
\correspondingauthor{Robert Corty}
\setboolean{displaycopyright}{true}

\begin{abstract}
Existing methods for QTL mapping in experimental crosses assume that the extent of residual variation is constant across all individuals.
When this assumption is not met, the simultaneous mean-variance modeling approach described in the companion article can yield higher power, better protection against false positives, and detection of the rarely-considered class of QTL that influence phenotype variance, vQTL.
The \texttt{R} package \texttt{vqtl} implements this statistical model, a significance testing procedure, and facilitates visualization and interpretation of results.
Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for many geneticists to analyze the results of their experimental crosses with \texttt{vqtl}, possibly discovering new QTL.
Here, we demonstrate a typical usage case.
\end{abstract}

\begin{document}

\maketitle
\thispagestyle{firststyle}
\logomark
\articletypemark
\marginmark
\firstpagefootnote
\correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
\vspace{-11pt}%

\noindent Introductory material, motivation, rationale, etc.  Introductory material, motivation, rationale, etc.  Introductory material, motivation, rationale, etc.  Introduce mQTL, vQTL, and mvQTL nomenclature.

\section*{Simulation}

We will illustrate the usage and interpretation of the \texttt{vqtl} package through the analysis of an experimental cross simulated with \texttt{R/qtl}.
The simulated population consists of 100 male and 100 female F2 offspring, with 5 chromosomes of length 100 cM tagged by 30 equally-spaced markers each.
We simulate four phenotypes

\begin{enumerate}
	\item phenotype1 consists only of random noise
	\item phenotype2 has one large-effect mQTL at the 15th marker on chromosome one
	\item phenotype3 has one large-effect vQTL at the 15th marker on chromosome two
	\item phenotype2 has one large-effect mvQTL at the 15th marker on chromosome three
\end{enumerate}

\section*{Conducting a genome scan}

Analogous to the central \texttt{scanone} function in package \texttt{qtl}, package \texttt{vqtl} provides function \texttt{scanonevar}.
Like \texttt{scanone}, \texttt{scanonevar} takes as input an object of class \texttt{cross}, and the specification of a a single phenotype contained in the \texttt{cross} and returns as output a statistic for each genetic locus in the \texttt{cross} that summarizes the association between that locus and the phenotype.
However, \texttt{scanone} only tests for association between each locus and the phenotype mean, while \texttt{scanonevar} tests for associations with mean and variance.
Thus \texttt{scanonevar} returns not one, but three, statistics for each genetic locus: one that summarizes the association of the locus with phenotype mean, one that summarizes the association of the locus with phenotype variance, and one that jointly summarizes the association between the locus, phenotype mean, and phenotype variance.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{images/LOD_scans.pdf}
	\caption{For each of the four simulated phenotypes, we have the three new tests in black, blue, and red.  The traditional test is in green and clearly similar to the blue test in the vast manority of loci. \label{fig:lod_score_scans}}
\end{figure}

\section{Assessing the Statistical Significance of Findings}

Figure~\ref{fig:lod_score_scans} shows the log of the odds at each locus for each of the four tests as a measure of association.
This measure is traditional in QTL mapping, but in this scenario where multiple different tests are conducted at each locus, it suggesets a misleading interpretation, that points of different colors with the same LOD score represent an equal quantity of evidence against the null (no genetic effect).
In fact, as is visible in the top panel of figure~\ref{fig:lod_score_scans}, the LOD score that results from differnet tests is different under the null.
For example, the LOD score of the DGLM-joint test is always higher than the other three tests because all the other tests are ``nested'' inside the DGLM joint test.
% To put the four tests on a level playing field, we must conduct randomized genome scans and assess the observed LOD scores in the context of the per-scan maxima from the randomized genome scans, as illustrated in figure~\ref{fig:FWER}.

% \begin{figure}
% 	\includegraphics[width=0.5\textwidth]{images/FWER.pdf}
% 	\caption{Illustration of the method for controlling family-wide error rate in the genome scan. \label{fig:FWER}}
% \end{figure}

\section{Interpreting Significant Results}

% \section*{Guide to using this template in Overleaf}

% This template is provided to help you prepare your article for submission to the \textit{Genetics}.

% \section*{Author Affiliations}

% For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.

% \section*{Your Abstract}

% In addition to the guidelines provided in the example abstract above, your abstract should:

% \begin{itemize}
% \item provide a synopsis of the entire article;
% \item begin with the broad context of the study, followed by specific background for the study;
% \item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
% \item emphasize new or important aspects of the research;
% \item engage the broad readership of G3 and be understandable to a diverse audience (avoid using jargon);
% \item be a single paragraph of less than 250 words;
% \item contain the full name of the organism studied;
% \item NOT contain citations or abbreviations.
% \end{itemize}

% \section*{Introduction}

% For the introduction, authors should be mindful of the broad readership of the journal. The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study. The scope and impact of the work should be clearly stated.

% In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism. Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone. General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.

% Authors are encouraged to:

% \begin{itemize}
% \item cite the supporting literature completely rather than select a subset of citations;
% \item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
% \item to cite similar work in other organisms.
% \end{itemize}

% \section*{Materials and Methods}

% Manuscripts submitted to G3 should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist. If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.  For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis. If working with mutants indicate how many independent mutants were isolated. If working with populations indicate how samples were collected and whether they were random with respect to the target population.

% \subsection*{Statistical Analysis}

% Indicate which statistical analysis has been performed and describe the method and model applied. If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided. The type of correction applied should be clearly stated. It should also be clear whether the p-values reported are raw, or after correction. Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.

% \subsection*{Data Availability}

% At the end of the Materials and Methods section, include a statement on reagent and data availability. Please read the Data and Reagent Policy before writing the statement. Make sure to list the accession numbers or DOIs of any data you have placed in public repositories. List the file names and descriptions of any data you will upload as supplemental information. The statement should also include any applicable IRB numbers. You may include specifications for how to properly acknowledge or cite the data.

% For example: Strains are available upon request. File S1 contains detailed descriptions of all supplemental files. File S2 contains SNP ID numbers and locations. File S3 contains genotypes for each individual. Sequence data are available at GenBank and the accession numbers are listed in File S3. Gene expression data are available at GEO with the accession number: GDS1234. Code used to generate the simulated data is provided in file S4.

% \section*{Results and Discussion}

% The results and discussion should not be repetitive. The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context. Authors are encouraged to cite recent work relevant to their interpretations. Present and discuss results only once, not in both the Results and Discussion sections. It is sometimes acceptable to combine results and discussion. The text should be as succinct as possible. Heed Strunk and White's dictum: "Omit needless words!"

% \section*{Additional guidelines}

% \subsection*{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement. Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.

% \subsection*{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes". Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).

% \subsection*{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.  Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII. Write the names of strains in roman except when incorporating specific genotypic designations. Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme. Do not use "+" to indicate wild type. Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.

% \section*{In-text Citations}

% Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}.

% For examples of different references, please see the example bibliography file (accessible via the Project menu in the Overleaf editor). This contains examples of articles \citep{neher2013genealogies, rodelsperger2014characterization}, a book \citep{Sturtevent2001}, a book chapter \citep{Sturtevent2001chp7}, ahead-of-print work \citep{Starita2015} and software \citep{Kruijer2015}.

% \section*{Examples of Article Components}
% \label{sec:examples}

% The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.

% \section*{First level section header}

% Use this level to group two or more closely related headings in a long article.

% \subsection*{Second level section header}

% Second level section text.

% \subsubsection*{Third level section header:}

% Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.

% \section*{Figures and Tables}

% Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.

% \subsection*{Sample Figure}

% Figure \ref{fig:spectrum} shows an example figure.

% \begin{figure}[htbp]
% \renewcommand{\familydefault}{\sfdefault}\normalfont
% \centering
% \includegraphics[width=\linewidth]{example-figure-g3}
% \caption{Example figure from \url{http://dx.doi.org/10.1534/g3.115.017509}. Please include your figures in the manuscript for the review process. You can upload figures to Overleaf via the Project menu. Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).  Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.  RGB, CMYK, and Grayscale are all acceptable. Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process. Figures should be 10-20 cm in width and 1-25 cm in height. Graph axes must be exactly perpendicular and all lines of equal density.
% Label multiple figure parts with A, B, etc. in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight. Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented. All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used. Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.  
% }%
% \label{fig:spectrum}
% \end{figure}

% \subsection*{Sample Video}

% Figure \ref{video:spectrum} shows how to include a video in your manuscript.

% \begin{figure}[htbp]
% \renewcommand{\familydefault}{\sfdefault}\normalfont
% \centering
% \includegraphics[width=\linewidth]{example-figure-g3}
% \caption{Example movie (the figure file above is used as a placeholder for this example). G3 supports video and movie files that can be linked from any portion of the article - including the abstract. Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.   
% }%
% \label{video:spectrum}
% \end{figure}

% \subsection*{Sample Table}

% Table \ref{tab:shape-functions} shows an example table. Avoid shading, color type, line drawings, graphics, or other illustrations within tables. Use tables for data only; present drawings, graphics, and illustrations as separate figures. Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.

% Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.

% \begin{table*}[htbp]
% \renewcommand{\familydefault}{\sfdefault}\normalfont
% \centering
% \caption{\bf Students and their grades}
% \begin{tableminipage}{\textwidth}
% \begin{tabularx}{\textwidth}{XXXX}
% \hline
% \header Student & Grade\footnote{This is an example of a footnote in a table. Lowercase, superscript italic letters (a, b, c, etc.) are used by default. You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
% \hline
% Alice & 82\% & 1 & Performed very well.\\
% Bob & 65\% & 3 & Not up to his usual standard.\\
% Charlie & 73\% & 2 & A good attempt.\\
% \hline
% \end{tabularx}
%   \label{tab:shape-functions}
% \end{tableminipage}
% \end{table*}

% \section*{Sample Equation}

% Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
% \begin{equation}
% S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
%       = \frac{1}{n}\sum_{i}^{n} X_i
% \label{eq:refname1}
% \end{equation}
% denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.

% \bibliography{example-bibliography}

\end{document}

### 2016-06-22 scripts/make_LOD_score_scans_figure.R

@@ -1,0 +1,52 @@
+library(vqtl)
+
+set.seed(27599)
+
+my.cross <- sim.cross(map = sim.map(len = rep(100, 4), n.mar = 30, eq.spacing = TRUE, include.x = FALSE),
+											n.ind = 200,
+											type = 'f2')
+my.cross$pheno$sex <- rep(x = c(0, 1), each = 100)
+my.cross <- calc.genoprob(my.cross)
+
+my.cross$pheno$phenotype1 <- rnorm(n = nind(my.cross))
+my.cross$pheno$phenotype2 <- rnorm(n = nind(my.cross), mean = 0.8*my.cross$geno$`1`$data[,15])
+my.cross$pheno$phenotype3 <- rnorm(n = nind(my.cross), sd = my.cross$geno$`2`$data[,15])
+my.cross$pheno$phenotype4 <- rnorm(n = nind(my.cross), mean = my.cross$geno$`3`$data[,15], sd = my.cross$geno$`3`$data[,15])
+
+
+a1 <- scanonevar(cross = my.cross,
+								 mean.formula = 'phenotype1 ~ sex + mean.QTL.add + mean.QTL.dom',
+								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
+								 return.models = TRUE)$varscan
+a2 <- scanone(cross = my.cross,
+							pheno.col = 'phenotype1')
+
+b1 <- scanonevar(cross = my.cross,
+								 mean.formula = 'phenotype2 ~ sex + mean.QTL.add + mean.QTL.dom',
+								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
+								 return.models = TRUE)$varscan
+b2 <- scanone(cross = my.cross,
+							pheno.col = 'phenotype2')
+
+c1 <- scanonevar(cross = my.cross,
+								 mean.formula = 'phenotype3 ~ sex + mean.QTL.add + mean.QTL.dom',
+								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
+								 return.models = TRUE)$varscan
+c2 <- scanone(cross = my.cross,
+							pheno.col = 'phenotype3')
+
+d1 <- scanonevar(cross = my.cross,
+								 mean.formula = 'phenotype4 ~ sex + mean.QTL.add + mean.QTL.dom',
+								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
+								 return.models = TRUE)$varscan
[…]

### 2016-08-31 2016_G3_PackageVQTL_Corty.tex
continued progress
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 \usepackage{tabu}
-\usepackage{subfigmat}
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 
@@
 
-\title{\texttt{vqtl}: An \texttt{R} package for Efficient QTL Mapping on Phenotypes with Heterogeneous Variance}
+\title{\texttt{vqtl}: An \texttt{R} package for QTL Mapping on Phenotypes with Heterogeneous Variance}
 
@@
 \begin{abstract}
-Existing methods for QTL mapping in experimental crosses assume that the extent of residual variation is constant across all individuals.
-When this assumption is not met, the simultaneous mean-variance modeling approach described in the companion article can yield higher power, better protection against false positives, and detection of the rarely-considered class of QTL that influence phenotype variance, vQTL.
-The \texttt{R} package \texttt{vqtl} implements this statistical model, a significance testing procedure, and facilitates visualization and interpretation of results.
-Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for many geneticists to analyze the results of their experimental crosses with \texttt{vqtl}, possibly discovering new QTL.
-Here, we demonstrate a typical usage case.
+Existing methods for QTL mapping in experimental crosses assume that the amount of residual variation is constant across all individuals.
+Many common situations can violate this assumption.
+For example, female mice may have more variable phenotypes than males, some experimenters may make more precise measurements than others, and specific genetic loci may influence the extent of variation between individuals.
+In all such cases, the heterogeneous variance modeling approach demonstrated here provides higher power, better protection against false positives, and allows for detection of QTL that influence phenotype variance, termed vQTL.
+The \texttt{R} package \texttt{vqtl} makes it easy for geneticists to apply the heterogeneous variance model, control family-wide error rate (FWER), and visualize and interpret their results.
+Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze the results of their experimental crosses with \texttt{vqtl}, possibly discovering new QTL.
+Here, we demonstrate typical usage.
 \end{abstract}
@@
 
-\noindent Introductory material, motivation, rationale, etc.
-Introductory material, motivation, rationale, etc.
-Introductory material, motivation, rationale, etc.
-Introduce mQTL, vQTL, and mvQTL nomenclature.
+\noindent Experimental crosses of inbred organisms have been a cornerstone of forward genetics for over a century \citep{mendel1866}.
+They have provided important insights on nearly every trait of interest in human disease, agriculture, and livestock production.
+Studies across all these diverse fields were enabled by advances in methods for efficient breeding, phenotyping \citep{Yang2014a}, genotyping \citep{Williams1990}, statistical methods \citep{Lander1989a}, and software tools \citep{Broman2003}.
 
-\section*{Simulation}
+One assumption that has been constant throughout all these advances is that the extent of residual variation is constant across all organisms in a study population.
+Said another way, it has always been assumed that no environmental factors and no genetic factors influence the extent of residual phenotype variation.
+In the companion piece of this article, we introduced a statistical modeling approach that accommodates heterogeneity in residual variation within a study population.
+We further demonstrated two critical benefits of using this ``simultaneous mean-variance'' modeling approach.
 
-We will illustrate the usage and interpretation of the \texttt{vqtl} package through the analysis of an experimental cross simulated with \texttt{R/qtl}.
-The simulated population consists of 100 male and 100 female F2 offspring, with 5 chromosomes of length 100 cM tagged by 30 equally-spaced markers each.
+\begin{enumerate}
+	\item It allows for the detection of genetic loci that influence residual phenotype variation, which are likely to play central roles in the network of molecular interactions that gives rise to a complex trait.
+	\item It allows for accurate consideration of the quantity of information provided by each individual.
+Individuals with less residual phenotype variance provide more information about the mean of the groups to which they belong, as is evident in the standard equation for the standard error of the mean: $\text{SE}_\mu = \sigma/\sqrt{n}$.
+\end{enumerate}
+
+The companion piece of this article is based on our reanalysis of an F2 intercross of two mouse strains carried out and published in 2008, but some of the results are completely novel.
+In support of other researchers who may be interested in reanalysing previously-conducted mapping studies using the simultaneous mean-variance mapping approach, we have composed an R package that provides functions for conducting mean-variance genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
+R package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
+
+Here, we demonstrate typical usage of the \texttt{vqtl} package.
+The code used to generate all statistics and figures in this paper is available at \texttt{github.com/rcorty}.
+
+\section*{Simulated Experimental Cross}
+
+We used \texttt{R/qtl} to simulate the experimental cross to be analyzed.
+The simulated population consists of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 30 equally-spaced markers and genotype probabilities estimated by HMM to 2 cM separation.
 We simulate four phenotypes
@@
 \begin{enumerate}
-	\item phenotype1 consists only of random noise
-	\item phenotype2 has one large-effect mQTL at the 15th marker on chromosome one
-	\item phenotype3 has one large-effect vQTL at the 15th marker on chromosome two
-	\item phenotype2 has one large-effect mvQTL at the 15th marker on chromosome three
+	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests
+	\item \texttt{phenotype2} is influenced by the 15th marker on chromosome one.
+The marker influences the mean of the phenotype, but not the variance, so it will serve as an example of a pure ``mQTL''.
+	\item \texttt{phenotype3} is influenced by the 15th marker on chromosome two.
+The marker influences the variance of the phenotype, but not the mean, so it will serve as an example of a pure ``vQTL''.
+	\item \texttt{phenotype4} is influenced by the 15th marker on chromosome three.
+The marker influences both the mean and the variance of the phenotype, so it will serve as an example of a joint ``mvQTL''.
 \end{enumerate}
+
+We additionally consider \texttt{phenotype1'} through \texttt{phenotype4'}, which have the same genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, but additionally have covariate effects on phenotype variance.
+All the same analyses and plots that are shown for \texttt{phenotype1} through \texttt{phenotype4} are shown for \texttt{phenotype1'} through \texttt{phenotype4'} in the supplementary materials.
 
@@
 
-Analogous to the central \texttt{scanone} function in package \texttt{qtl}, package \texttt{vqtl} provides function \texttt{scanonevar}.
-Like \texttt{scanone}, \texttt{scanonevar} takes as input an object of class \texttt{cross}, and the specification of a a single phenotype contained in the \texttt{cross} and returns as output a statistic for each genetic locus in the \texttt{cross} that summarizes the association between that locus and the phenotype.
-However, \texttt{scanone} only tests for association between each locus and the phenotype mean, while \texttt{scanonevar} tests for associations with mean and variance.
-Thus \texttt{scanonevar} returns not one, but three, statistics for each genetic locus: one that summarizes the association of the locus with phenotype mean, one that summarizes the association of the locus with phenotype variance, and one that jointly summarizes the association between the locus, phenotype mean, and phenotype variance.
+The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone}.
+Analogously, the central function for genetic mapping in package \texttt{vqtl} is \texttt{scanonevar}.
+
+\texttt{scanonevar} takes three required inputs:
+
+\begin{enumerate}
+    \item \texttt{cross} contains the genetic and phenotypic information from an exerimental cross.  This object can be the same \texttt{cross} object used in package \texttt{qtl}.
+    \item \texttt{mean.formula} specifies the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted (\texttt{additive} and \texttt{dominance} components by default).
+The \texttt{mean.formula} uses the standard \texttt{R} formula notation.
+    \item \texttt{var.formula} specifies the covariates to be corrected for as well as the QTL terms to be fitted (\texttt{additive} and \texttt{dominance} components by default) in modeling the residual variance.
+The \texttt{var.formula} also uses the standard \texttt{R} formula notation.
+\end{enumerate}
+
+Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
+Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
+
+Unlike \texttt{scanone}, which only tests for association between each locus and the phenotype mean, \texttt{scanonevar} computes three tests for each locus -- association with phenotype mean, association with phenotype variance, and joint association with phenotype mean and variance.
+The statistic for each of these associations is a LOD score, the log of the ratio of the likelihood of the alternative model to the null model.
+The details of the null and alternative models used in each of the three tests can be found in the companion article.
+LOD scores are hard to compare across autosomal and sex chromosomes due to the difference in number of parameters.
+The $p$-value of each LOD score is also calculated, based on the asymptotic $\chi^2$ distribution with the appropriate degrees of freedom of each test, but the interpretation of these $p$-values is clouded by the multiplicity of tests that are conducted in each scan.
+Assessing the significance of the LOD scores in a manner that controls family-wide error rate (FWER) to the desired level is decribed below and is the recommened method for assessing the significance of QTL mapping results.
+
+The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
+Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
+Calling \texttt{summary} on this object produces a summary of how the scan was conducted and what the results were.
 
 \begin{figure}
-	\includegraphics[width=0.5\textwidth]{images/LOD_scans.pdf}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phen1.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phen2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phen3.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phen4.pdf}
+    \end{subfigure}
+
     \caption{For each of the four simulated phenotypes, we have the three new tests in black, blue, and red.
@@
 
-\section{Assessing the Statistical Significance of Findings}
+\section*{Assessing the Significance of Results}
 
-Figure~\ref{fig:lod_score_scans} shows the log of the odds at each locus for each of the four tests as a measure of association.
-This measure is traditional in QTL mapping, but in this scenario where multiple different tests are conducted at each locus, it suggesets a misleading interpretation, that points of different colors with the same LOD score represent an equal quantity of evidence against the null (no genetic effect).
-In fact, as is visible in the top panel of figure~\ref{fig:lod_score_scans}, the LOD score that results from differnet tests is different under the null.
-For example, the LOD score of the DGLM-joint test is always higher than the other three tests because all the other tests are ``nested'' inside the DGLM joint test.
+The effective number of statistical tests conducted in a family of tests (a genome scan) is typically must be estimated to control family-wide error rate (FWER).
+The effective number of tests in a genome scan, however, is difficult to estimate.
+One lower bound is the number of chromosomes.
+Due to the randomization in meiosis no two non-syntenic loci are correlated in an experimental cross and therefore tests on different chromosomes are always independent.
+But there are many tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
+One upper bound on the effective number of tests is the total number of loci.
+But, loci on the same chromosome are often in linkage disequilibrium and so the total number of loci is an over-estimate.
+
+Our empirical approach avoids the need to estimate the effective number of tests.
+We conduct many genomes scans, each with its own permutation of the genotype probabilities and estimate an extreme value distribution for the genome-wide maximum LOD score of each test.
+This approach is implemented in \texttt{scanonevar.perm}.
+
+\texttt{scanonevar.perm} takes two required inputs:
+
+\begin{enumerate}
+	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed.
+	\item \texttt{n.perms} is the number of permutations to conduct.
+\end{enumerate}
+
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with one important additional piece of information.
+An empirical $p$-value for each test at each locus is included.
+These $p$-values are FWER-corrected, so a value of $0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe \textit{some} locus this significant or more significant.
+A list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
+
+Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
+We recommend at least 100, and rarely more than 1000.
+These permutation scans can be broken into groups, run on separate computers, and combined with the generic \texttt{c} function.
+
+\begin{figure}
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phen1.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phen2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phen3.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phen4.pdf}
+    \end{subfigure}
+
+    \caption{For each of the four simulated phenotypes, we have the three new tests in black, blue, and red.
+The traditional test is in green and clearly similar to the blue test in the vast manority of loci.
+\label{fig:empir_p_scans}}
+\end{figure}
+
+% Figure~\ref{fig:lod_score_scans} shows the log of the odds at each locus for each of the four tests as a measure of association.
+% This measure is traditional in QTL mapping, but in this scenario where multiple different tests are conducted at each locus, it suggesets a misleading interpretation, that points of different colors with the same LOD score represent an equal quantity of evidence against the null (no genetic effect).
+% In fact, as is visible in the top panel of figure~\ref{fig:lod_score_scans}, the LOD score that results from differnet tests is different under the null.
+% For example, the LOD score of the DGLM-joint test is always higher than the other three tests because all the other tests are ``nested'' inside the DGLM joint test.
 % To put the four tests on a level playing field, we must conduct randomized genome scans and assess the observed LOD scores in the context of the per-scan maxima from the randomized genome scans, as illustrated in figure~\ref{fig:FWER}.
@@
 
-\section{Interpreting Significant Results}
+\section*{Investigate Significant Findings}
 
-% \section*{Guide to using this template in Overleaf}
-
-% This template is provided to help you prepare your article for submission to the \textit{Genetics}.
-
-% \section*{Author Affiliations}
-
-% For the authors' names, indicate different affiliations with the symbols: $\ast$, $\dagger$, $\ddagger$, $\S$. After four authors, the symbols double, triple, quadruple, and so forth as required.
-
-% \section*{Your Abstract}
-
-% In addition to the guidelines provided in the example abstract above, your abstract should:
-
-% \begin{itemize}
-% \item provide a synopsis of the entire article;
-% \item begin with the broad context of the study, followed by specific background for the study;
-% \item describe the purpose, methods and procedures, core findings and results, and conclusions of the study;
-% \item emphasize new or important aspects of the research;
-% \item engage the broad readership of G3 and be understandable to a diverse audience (avoid using jargon);
-% \item be a single paragraph of less than 250 words;
-% \item contain the full name of the organism studied;
-% \item NOT contain citations or abbreviations.
-% \end{itemize}
-
-% \section*{Introduction}
-
-% For the introduction, authors should be mindful of the broad readership of the journal.
-The introduction should set the stage for the importance of the work to a generalist reader and draw the reader in to the specific study.
-The scope and impact of the work should be clearly stated.
-
-% In individual organisms where a mutant is being studied, the rationale for the study of that mutant must be clear to a geneticist not studying that particular organism.
-Similarly, study of particular phenotypes should be justified broadly and not on the basis of interest for that organism alone.
-General background on the importance of the genetic pathway and/or phenotype should be provided in a single, well-reasoned paragraph near the beginning of the introduction.
-
-% Authors are encouraged to:
-
-% \begin{itemize}
-% \item cite the supporting literature completely rather than select a subset of citations;
-% \item provide important background citations, including relevant review papers (to help orient the non-specialist reader);
-% \item to cite similar work in other organisms.
-% \end{itemize}
-
-% \section*{Materials and Methods}
-
-% Manuscripts submitted to G3 should contain a clear description of the experimental design in sufficient detail so that the experimental analysis could be repeated by another scientist.
-If the level of detail necessary to explain the protocol goes beyond two paragraphs, give a short description in the main body of the paper and prepare a detailed description for supporting information.
-For example, details would include indicating how many individuals were used, and if applicable how individuals or groups were combined for analysis.
-If working with mutants indicate how many independent mutants were isolated.
-If working with populations indicate how samples were collected and whether they were random with respect to the target population.
-
-% \subsection*{Statistical Analysis}
-
-% Indicate which statistical analysis has been performed and describe the method and model applied.
-If many genes were examined simultaneously, or many phenotypes, a multiple comparison correction should be used to control the type I error rate, or a rationale for not applying a correction must be provided.
-The type of correction applied should be clearly stated.
-It should also be clear whether the p-values reported are raw, or after correction.
-Corrected p-values are often appropriate, but raw p-values should be available in the supporting materials so that others may perform their own corrections.
-
-% \subsection*{Data Availability}
-
-% At the end of the Materials and Methods section, include a statement on reagent and data availability.
-Please read the Data and Reagent Policy before writing the statement.
-Make sure to list the accession numbers or DOIs of any data you have placed in public repositories.
-List the file names and descriptions of any data you will upload as supplemental information.
-The statement should also include any applicable IRB numbers.
-You may include specifications for how to properly acknowledge or cite the data.
-
-% For example: Strains are available upon request.
-File S1 contains detailed descriptions of all supplemental files.
-File S2 contains SNP ID numbers and locations.
-File S3 contains genotypes for each individual.
-Sequence data are available at GenBank and the accession numbers are listed in File S3.
-Gene expression data are available at GEO with the accession number: GDS1234.
-Code used to generate the simulated data is provided in file S4.
-
-% \section*{Results and Discussion}
-
-% The results and discussion should not be repetitive.
-The results section should give a factual presentation of the data and all tables and figures should be referenced; the discussion should not summarize the results but provide an interpretation of the results, and should clearly delineate between the findings of the particular study and the possible impact of those findings in a larger context.
-Authors are encouraged to cite recent work relevant to their interpretations.
-Present and discuss results only once, not in both the Results and Discussion sections.
-It is sometimes acceptable to combine results and discussion.
-The text should be as succinct as possible.
-Heed Strunk and White's dictum: "Omit needless words!"
-
-% \section*{Additional guidelines}
-
-% \subsection*{Numbers} In the text, write out numbers nine or less except as part of a date, a fraction or decimal, a percentage, or a unit of measurement.
-Use Arabic numbers for those larger than nine, except as the first word of a sentence; however, try to avoid starting a sentence with such a number.
-
-% \subsection*{Units} Use abbreviations of the customary units of measurement only when they are preceded by a number: "3 min" but "several minutes".
-Write "percent" as one word, except when used with a number: "several percent" but "75\%." To indicate temperature in centigrade, use ° (for example, 37°); include a letter after the degree symbol only when some other scale is intended (for example, 45°K).
-
-% \subsection*{Nomenclature and Italicization} Italicize names of organisms even when  when the species is not indicated.
-Italicize the first three letters of the names of restriction enzyme cleavage sites, as in HindIII.
-Write the names of strains in roman except when incorporating specific genotypic designations.
-Italicize genotype names and symbols, including all components of alleles, but not when the name of a gene is the same as the name of an enzyme.
-Do not use "+" to indicate wild type.
-Carefully distinguish between genotype (italicized) and phenotype (not italicized) in both the writing and the symbolism.
-
-% \section*{In-text Citations}
-
-% Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}.
-
-% For examples of different references, please see the example bibliography file (accessible via the Project menu in the Overleaf editor).
-This contains examples of articles \citep{neher2013genealogies, rodelsperger2014characterization}, a book \citep{Sturtevent2001}, a book chapter \citep{Sturtevent2001chp7}, ahead-of-print work \citep{Starita2015} and software \citep{Kruijer2015}.
-
-% \section*{Examples of Article Components}
-% \label{sec:examples}
-
-% The sections below show examples of different header levels, which you can use in the primary sections of the manuscript (Results, Discussion, etc.) to organize your content.
-
-% \section*{First level section header}
-
-% Use this level to group two or more closely related headings in a long article.
-
-% \subsection*{Second level section header}
-
-% Second level section text.
-
-% \subsubsection*{Third level section header:}
-
-% Third level section text. These headings may be numbered, but only when the numbers must be cited in the text.
-
-% \section*{Figures and Tables}
-
-% Figures and Tables should be labelled and referenced in the standard way using the \verb|\label{}| and \verb|\ref{}| commands.
-
-% \subsection*{Sample Figure}
-
-% Figure \ref{fig:spectrum} shows an example figure.
-
-% \begin{figure}[htbp]
-% \renewcommand{\familydefault}{\sfdefault}\normalfont
-% \centering
-% \includegraphics[width=\linewidth]{example-figure-g3}
-% \caption{Example figure from \url{http://dx.doi.org/10.1534/g3.115.017509}.
-Please include your figures in the manuscript for the review process.
-You can upload figures to Overleaf via the Project menu.
-Upon acceptance, we'll ask for your figure files to be uploaded in any of the following formats: TIFF (.tiff), JPEG (.jpg), Microsoft PowerPoint (.ppt), EPS (.eps), or Adobe Illustrator (.ai).
-Images should be a minimum of 300 dpi in resolution and 500 dpi minimum if line art images.
-RGB, CMYK, and Grayscale are all acceptable.
-Halftones should be high contrast with sharp detail, because some loss of detail and contrast is inevitable in the production process.
-Figures should be 10-20 cm in width and 1-25 cm in height.
-Graph axes must be exactly perpendicular and all lines of equal density.
-% Label multiple figure parts with A, B, etc.
-in bolded type, and use Arrows and numbers to draw attention to areas you want to highlight.
-Legends should start with a brief title and should be a self-contained description of the content of the figure that provides enough detail to fully understand the data presented.
-All conventional symbols used to indicate figure data points are available for typesetting; unconventional symbols should not be used.
-Italicize all mathematical variables (both in the figure legend and figure) , genotypes, and additional symbols that are normally italicized.
-% }%
-% \label{fig:spectrum}
-% \end{figure}
-
-% \subsection*{Sample Video}
-
-% Figure \ref{video:spectrum} shows how to include a video in your manuscript.
-
-% \begin{figure}[htbp]
-% \renewcommand{\familydefault}{\sfdefault}\normalfont
-% \centering
-% \includegraphics[width=\linewidth]{example-figure-g3}
-% \caption{Example movie (the figure file above is used as a placeholder for this example).
-G3 supports video and movie files that can be linked from any portion of the article - including the abstract.
-Acceptable formats include .asf, avi, .wav, and all types of Windows Media files.
-% }%
-% \label{video:spectrum}
-% \end{figure}
-
-% \subsection*{Sample Table}
-
-% Table \ref{tab:shape-functions} shows an example table.
-Avoid shading, color type, line drawings, graphics, or other illustrations within tables.
-Use tables for data only; present drawings, graphics, and illustrations as separate figures.
-Histograms should not be used to present data that can be captured easily in text or small tables, as they take up much more space.
-
-% Tables numbers are given in Arabic numerals. Tables should not be numbered 1A, 1B, etc., but if necessary, interior parts of the table can be labeled A, B, etc. for easy reference in the text.
-
-% \begin{table*}[htbp]
-% \renewcommand{\familydefault}{\sfdefault}\normalfont
-% \centering
-% \caption{\bf Students and their grades}
-% \begin{tableminipage}{\textwidth}
-% \begin{tabularx}{\textwidth}{XXXX}
-% \hline
-% \header Student & Grade\footnote{This is an example of a footnote in a table.
-Lowercase, superscript italic letters (a, b, c, etc.) are used by default.
-You can also use *, **, and *** to indicate conventional levels of statistical significance, explained below the table.} & Rank & Notes \\
-% \hline
-% Alice & 82\% & 1 & Performed very well.\\
-% Bob & 65\% & 3 & Not up to his usual standard.\\
-% Charlie & 73\% & 2 & A good attempt.\\
-% \hline
-% \end{tabularx}
-%   \label{tab:shape-functions}
-% \end{tableminipage}
-% \end{table*}
-
-% \section*{Sample Equation}
-
-% Let $X_1, X_2, \ldots, X_n$ be a sequence of independent and identically distributed random variables with $\text{E}[X_i] = \mu$ and $\text{Var}[X_i] = \sigma^2 < \infty$, and let
-% \begin{equation}
-% S_n = \frac{X_1 + X_2 + \cdots + X_n}{n}
-%       = \frac{1}{n}\sum_{i}^{n} X_i
-% \label{eq:refname1}
-% \end{equation}
-% denote their mean. Then as $n$ approaches infinity, the random variables $\sqrt{n}(S_n - \mu)$ converge in distribution to a normal $\mathcal{N}(0, \sigma^2)$.
-
-% \bibliography{example-bibliography}
+\bibliography{10_Aim1-2016_G3_PackageVQTL}
 

### 2016-08-31 FWER_correct_scans.R

@@ -1,0 +1,131 @@
+library(tidyr); library(dplyr); library(vqtl)
+
+set.seed(27599)
+
+my.cross <- qtl::sim.cross(map = qtl::sim.map(len = rep(100, 3), n.mar = 20, eq.spacing = TRUE, include.x = FALSE),
+													 n.ind = 400,
+													 type = 'f2')
+my.cross$pheno$sex <- rep(x = c(0, 1), each = 200)
+my.cross <- qtl::calc.genoprob(my.cross)
+
+my.cross$pheno$phenotype1 <- rnorm(n = qtl::nind(my.cross))
+my.cross$pheno$phenotype2 <- rnorm(n = qtl::nind(my.cross), mean = 0.6*my.cross$geno$`1`$data[,10])
+my.cross$pheno$phenotype3 <- rnorm(n = qtl::nind(my.cross), sd = exp(0.35*(my.cross$geno$`2`$data[,10] - 2)))
+my.cross$pheno$phenotype4 <- rnorm(n = qtl::nind(my.cross), mean = 0.3*my.cross$geno$`3`$data[,10], sd = exp(0.25*(my.cross$geno$`3`$data[,10] - 2)))
+
+
+# make fig 1 -- LOD score scans
+ymax <- 30
+
+a0 <- qtl::scanone(cross = my.cross,
+									 pheno.col = 'phenotype1')
+a1 <- scanonevar(cross = my.cross,
+								 mean.formula = phenotype1 ~ sex + mean.QTL.add + mean.QTL.dom,
+								 var.formula = ~sex + var.QTL.add + var.QTL.dom)
+p <- plot(x = a1, y = a0, ylim = c(0, ymax))
+p
+ggplot2::ggsave(filename = 'images/LOD_scan_phen1.pdf', height = 3, width = 9)
+
+
+b0 <- qtl::scanone(cross = my.cross,
+									 pheno.col = 'phenotype2')
+b1 <- scanonevar(cross = my.cross,
+								 mean.formula = phenotype2 ~ sex + mean.QTL.add + mean.QTL.dom,
+								 var.formula = ~sex + var.QTL.add + var.QTL.dom)
+p <- plot(x = b1, y = b0, ylim = c(0, ymax))
+p
+ggplot2::ggsave(filename = 'images/LOD_scan_phen2.pdf', height = 3, width = 9)
+
+
+c0 <- qtl::scanone(cross = my.cross,
+									 pheno.col = 'phenotype3')
[…]

### 2016-08-31 make_predictive_plots.R

@@ -1,0 +1,15 @@
+library(tidyr); library(dplyr); library(vqtl)
+
+set.seed(27599)
+
+my.cross <- sim.cross(map = sim.map(len = rep(100, 4), n.mar = 30, eq.spacing = TRUE, include.x = FALSE),
+											n.ind = 200,
+											type = 'f2')
+my.cross$pheno$sex <- rep(x = c(0, 1), each = 100)
+my.cross <- calc.genoprob(my.cross)
+
+my.cross$pheno$phenotype1 <- rnorm(n = nind(my.cross))
+my.cross$pheno$phenotype2 <- rnorm(n = nind(my.cross), mean = 0.8*my.cross$geno$`1`$data[,15])
+my.cross$pheno$phenotype3 <- rnorm(n = nind(my.cross), sd = my.cross$geno$`2`$data[,15])
+my.cross$pheno$phenotype4 <- rnorm(n = nind(my.cross), mean = my.cross$geno$`3`$data[,15], sd = my.cross$geno$`3`$data[,15])
+

### 2016-08-31 scripts/copy_over_bib.sh

@@ -1,0 +1,1 @@
+cp ~/Dropbox\ \(ValdarLab\)/bibtex/10_Aim1-2016_G3_PackageVQTL.bib .

### 2016-09-15 2016_G3_PackageVQTL_Corty.tex
wrote section on mean_var plots and appendix
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 Existing methods for QTL mapping in experimental crosses assume that the amount of residual variation is constant across all individuals.
-Many common situations can violate this assumption.
-For example, female mice may have more variable phenotypes than males, some experimenters may make more precise measurements than others, and specific genetic loci may influence the extent of variation between individuals.
+Many common situations violate this assumption.
+For example, female mice often have more variable phenotypes than males, some experimenters make more precise measurements than others, and specific genetic loci may influence the extent of variation between individuals.
 In all such cases, the heterogeneous variance modeling approach demonstrated here provides higher power, better protection against false positives, and allows for detection of QTL that influence phenotype variance, termed vQTL.
@@
 
-\section*{Simulated Experimental Cross}
+\section*{Simulate an Experimental Cross}
 
 We used \texttt{R/qtl} to simulate the experimental cross to be analyzed.
-The simulated population consists of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 30 equally-spaced markers and genotype probabilities estimated by HMM to 2 cM separation.
-We simulate four phenotypes
+We simulated a population of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden markov model.
+We simulated four phenotypes:
 
 \begin{enumerate}
-	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests
-	\item \texttt{phenotype2} is influenced by the 15th marker on chromosome one.
+	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
+	\item \texttt{phenotype2} is influenced by the 6th (middle) marker on chromosome one.
 The marker influences the mean of the phenotype, but not the variance, so it will serve as an example of a pure ``mQTL''.
-	\item \texttt{phenotype3} is influenced by the 15th marker on chromosome two.
+	\item \texttt{phenotype3} is influenced by the 6th (middle) marker on chromosome two.
 The marker influences the variance of the phenotype, but not the mean, so it will serve as an example of a pure ``vQTL''.
-	\item \texttt{phenotype4} is influenced by the 15th marker on chromosome three.
+	\item \texttt{phenotype4} is influenced by the 6th (middle) marker on chromosome three.
 The marker influences both the mean and the variance of the phenotype, so it will serve as an example of a joint ``mvQTL''.
@@
 
-We additionally consider \texttt{phenotype1'} through \texttt{phenotype4'}, which have the same genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, but additionally have covariate effects on phenotype variance.
-All the same analyses and plots that are shown for \texttt{phenotype1} through \texttt{phenotype4} are shown for \texttt{phenotype1'} through \texttt{phenotype4'} in the supplementary materials.
+We additionally consider \texttt{phenotype1x} through \texttt{phenotype4x}, which have the same type of genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, but additionally have covariate effects on phenotype variance.
+All the same analyses and plots that are shown for \texttt{phenotype1} through \texttt{phenotype4} are shown for \texttt{phenotype1x} through \texttt{phenotype4x} in the appendix.
 
-\section*{Conducting a genome scan}
+\section*{Conduct a Genome Scans}
 
@@
 Analogously, the central function for genetic mapping in package \texttt{vqtl} is \texttt{scanonevar}.
-
 \texttt{scanonevar} takes three required inputs:
@@
 The statistic for each of these associations is a LOD score, the log of the ratio of the likelihood of the alternative model to the null model.
+The results of \texttt{scanonevar} on each of the four described phenotypes is shown in figure~\ref{fig:lod_score_scans}.
 The details of the null and alternative models used in each of the three tests can be found in the companion article.
-LOD scores are hard to compare across autosomal and sex chromosomes due to the difference in number of parameters.
-The $p$-value of each LOD score is also calculated, based on the asymptotic $\chi^2$ distribution with the appropriate degrees of freedom of each test, but the interpretation of these $p$-values is clouded by the multiplicity of tests that are conducted in each scan.
-Assessing the significance of the LOD scores in a manner that controls family-wide error rate (FWER) to the desired level is decribed below and is the recommened method for assessing the significance of QTL mapping results.
+
+\begin{figure}[ht!]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
+    \end{subfigure}
+
+    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+The traditional test is in green and globally similar to the mean test.
+\label{fig:lod_score_scans}}
+\end{figure}
+
+The LOD score is the traditional test statistic in QTL mapping, but its interpretation can be difficult for two reasons:
+1) LOD scores must be interpreted differently in autosomes and sex chromosomes.  The sex chromosomes are typically fit with fewer parameters, so the LOD scores are typically lower.
+2) LOD scores from different tests must be interpreted differently.
+For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests, because they are nested tests.
+
+These difficulties in interpreting the LOD score are borne out in figure~\ref{fig:lod_score_scans}.
+Given the scale of the vertical axis, it seems that there are no important signicals in the genome scan of \texttt{phenotype1}.
+It is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
+But it is not evident how to compare the results of the mvQTL test to the results of the other tests.
+Nor would it be clear how to compare the results of tests from sex chromosomes, if they were present, to those from autosomes.
+Nor is is possible to make an appropriate multple testing correction to account for the many tests that are conducted in a genome scan.
+
+To put all genetic loci and all three tests on a level playing field, we consider two types of $p$-values:
+1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom.
+Though these $p$-values overcome the two problems of working with LOD scores described above, they leave unresolved the problem of interpreting $p$-values when many tests were conducted.
+2) Empirical, family-wide error rate (FWER) -corrected $p$-values resolve the multiple-testing issue, thus they are the recommended means of assessing the significance of QTL mapping results.
+Methods for calculating them are described in the next section.
 
@@
 
-\begin{figure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phen1.pdf}
-    \end{subfigure}
+\section*{Assess the Significance of Results}
 
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phen2.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phen3.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phen4.pdf}
-    \end{subfigure}
-
-    \caption{For each of the four simulated phenotypes, we have the three new tests in black, blue, and red.
-The traditional test is in green and clearly similar to the blue test in the vast manority of loci.
-\label{fig:lod_score_scans}}
-\end{figure}
-
-\section*{Assessing the Significance of Results}
-
-The effective number of statistical tests conducted in a family of tests (a genome scan) is typically must be estimated to control family-wide error rate (FWER).
+The effective number of statistical tests conducted in a family of tests (a genome scan) typically must be estimated to control family-wide error rate (FWER).
 The effective number of tests in a genome scan, however, is difficult to estimate.
@@
 One upper bound on the effective number of tests is the total number of loci.
-But, loci on the same chromosome are often in linkage disequilibrium and so the total number of loci is an over-estimate.
+But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate.
 
-Our empirical approach avoids the need to estimate the effective number of tests.
-We conduct many genomes scans, each with its own permutation of the genotype probabilities and estimate an extreme value distribution for the genome-wide maximum LOD score of each test.
+Our empirical approach sidesteps estimation of the effective number of tests.
+We conduct many genomes scans, each executing a carefully-constructed model comparison to maintain all mean and variance effects of covariates and any non-focal genetic effects on mean and variance.
+The details of these model comparisons are provided in the companion article.
+For each test, we extract the highest observed value of the test statistic from each permutation scan and use those to model a generalized extreme value distribution.
+The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value distribution to estimate the FWER-controlling $p$-values.
 This approach is implemented in \texttt{scanonevar.perm}.
-
 \texttt{scanonevar.perm} takes two required inputs:
@@
 \begin{enumerate}
-	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed.
+	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed through permutation.
 	\item \texttt{n.perms} is the number of permutations to conduct.
@@
 
-The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with one important additional piece of information.
-An empirical $p$-value for each test at each locus is included.
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with one important additional piece of information, an empirical $p$-value for each test at each locus.
 These $p$-values are FWER-corrected, so a value of $0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe \textit{some} locus this significant or more significant.
@@
 
-Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
-We recommend at least 100, and rarely more than 1000.
-These permutation scans can be broken into groups, run on separate computers, and combined with the generic \texttt{c} function.
-
-\begin{figure}
+\begin{figure}[ht!]
     \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phen1.pdf}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phen2.pdf}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phen3.pdf}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
     \end{subfigure}
@@
     \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phen4.pdf}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
     \end{subfigure}
 
-    \caption{For each of the four simulated phenotypes, we have the three new tests in black, blue, and red.
-The traditional test is in green and clearly similar to the blue test in the vast manority of loci.
+    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
 \label{fig:empir_p_scans}}
@@
 
-% Figure~\ref{fig:lod_score_scans} shows the log of the odds at each locus for each of the four tests as a measure of association.
-% This measure is traditional in QTL mapping, but in this scenario where multiple different tests are conducted at each locus, it suggesets a misleading interpretation, that points of different colors with the same LOD score represent an equal quantity of evidence against the null (no genetic effect).
-% In fact, as is visible in the top panel of figure~\ref{fig:lod_score_scans}, the LOD score that results from differnet tests is different under the null.
-% For example, the LOD score of the DGLM-joint test is always higher than the other three tests because all the other tests are ``nested'' inside the DGLM joint test.
-% To put the four tests on a level playing field, we must conduct randomized genome scans and assess the observed LOD scores in the context of the per-scan maxima from the randomized genome scans, as illustrated in figure~\ref{fig:FWER}.
-
-% \begin{figure}
-% 	\includegraphics[width=0.5\textwidth]{images/FWER.pdf}
-% 	\caption{Illustration of the method for controlling family-wide error rate in the genome scan. \label{fig:FWER}}
-% \end{figure}
+Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
+We recommend at least 100, and rarely more than 1000.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument.
+On an Intel Core i5, running 100 permutations takes about five minutes.
+When many phenotypes are studied, or the 5-50 minute runtime is problematic, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
+This function combines the permutations from all the inputted scans, re-evaluates the observed LOD scores in light of all available permutations, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values in about one second.
 
@@
 
+\begin{figure}[htbp]
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
+    \end{subfigure}  
+    \hfill
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
+    \end{subfigure}
+
+    \caption{mean\_var plots show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
+More information on the interpretation of these plots is provided in the `Investigate Significant Findings' section.
+\label{fig:mean_var_plots}}
+\end{figure}
+
+Having identified some QTL, we want to visualize the allele effects at those loci.
+Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
+To investigate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
+This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
+There are both ``model-free'' and ``model-based'' versions of this plotting utility, but here we show only the ``model-based'' version for conciseness.
+
+Figure~\ref{fig:mean_var_plots} shows a model-based \texttt{mean\_var\_plot} for each of the four phenotypes.
+In each plot, the location of the dot shows the estimated mean and variance effect of each genotype, with the mean effect indicated by the horizontal position and the variance effect indicated by the vertical position.
+The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the variance estimate.
+Additional grouping factors, such as sex, treatment group, or batch can also be specified and all combinations of effects for which an observation was made are plotted, but we here we show only the allele effects for clarity.
+
+For phenotype1, the \texttt{mean\_var\_plot} is shown at the first marker of the first chromosome.
+The estimates of the genotype effects on phenotype mean and variance are within one standard error of each other.
+This pattern is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+
+For phenotype2, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the first chromosome.
+The estimates of the genotype effects differ in the horizontal axis, but not the vertical axis.
+This pattern is consistent with the fact that there is a genetic effect on phenotype mean but none on phenotype variance and with the highly significant $p$-value for the mQTL test, but non-significant $p$-value for the vQTL test.
+
+For phenotype3, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the second chromosome.
+The estimates of the genotype effects differ in the vertical axis, but on on the horizontal axis.
+This pattern is consistent with the fact that there is a genetic effect on phenotype variance but none on phenotype mean and with the highly significant $p$-value for the vQTL test but non-significant $p$-value for the mQTL test.
+
+For phenotype4, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the third chromosome.
+The estimates fo the genotype effects differ in both the horizontal and vertical axes.
+This pattern is consistent with the fact that there is a genetic effect on both phenotype mean and variance and with the highly significant $p$-value for the mvQTL test and the marginally significant $p$-values for the other two tests.
+
+\section*{Conclusion}
+
+We have demonstrated typical usage of the new \texttt{vqtl} \texttt{R} package for QTL mapping in experimental crosses.
+This package is most appropriate for crosses and phenotypes where nuisance covariates or genetic factors are known or suspected to influence phenotype variance.
+In the case of genetic factors, they can be mapped.
+In the case of nuisance covariates, they can be accommodated.
+
+The central function of this package, \texttt{scanonevar}, carries out the initial genome scan.
+The permutation function, \texttt{scanonevar.perm}, executes a carefully-constructed set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
+A suite of plotting functions, e.g. the \texttt{mean\_var\_plot}, allows geneticists to evaluate the allelic and covariate effects that underlie detected signals.
+
 \bibliography{10_Aim1-2016_G3_PackageVQTL}
 
+\newpage
+\section*{Appendix}
+
+\begin{figure}[ht!]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4x.pdf}
+    \end{subfigure}
+
+    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+The traditional test is in green and globally similar to the mean test.
+\label{fig:apdx_lod_score_scans}}
+\end{figure}
+
+\begin{figure}[ht!]
+    \begin{subfigure}{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.5\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4x.pdf}
+    \end{subfigure}
+
+    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
+\label{fig:apdx_empir_p_scans}}
+\end{figure}
+
+\begin{figure}[htbp]
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1x.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2x.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3x.pdf}
+    \end{subfigure}  
+    \hfill
+    \begin{subfigure}[t]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4x.pdf}
+    \end{subfigure}
+
+    \caption{mean\_var plots show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
+More information on the interpretation of these plots is provided in the `Investigate Significant Findings' section.
+\label{fig:apdx_mean_var_plots}}
+\end{figure}
+
 \end{document}

### 2016-09-15 FWER_correct_scans.R

@@ -1,131 +1,0 @@
-library(tidyr); library(dplyr); library(vqtl)
-
-set.seed(27599)
-
-my.cross <- qtl::sim.cross(map = qtl::sim.map(len = rep(100, 3), n.mar = 20, eq.spacing = TRUE, include.x = FALSE),
-													 n.ind = 400,
-													 type = 'f2')
-my.cross$pheno$sex <- rep(x = c(0, 1), each = 200)
-my.cross <- qtl::calc.genoprob(my.cross)
-
-my.cross$pheno$phenotype1 <- rnorm(n = qtl::nind(my.cross))
-my.cross$pheno$phenotype2 <- rnorm(n = qtl::nind(my.cross), mean = 0.6*my.cross$geno$`1`$data[,10])
-my.cross$pheno$phenotype3 <- rnorm(n = qtl::nind(my.cross), sd = exp(0.35*(my.cross$geno$`2`$data[,10] - 2)))
-my.cross$pheno$phenotype4 <- rnorm(n = qtl::nind(my.cross), mean = 0.3*my.cross$geno$`3`$data[,10], sd = exp(0.25*(my.cross$geno$`3`$data[,10] - 2)))
-
-
-# make fig 1 -- LOD score scans
-ymax <- 30
-
-a0 <- qtl::scanone(cross = my.cross,
-									 pheno.col = 'phenotype1')
-a1 <- scanonevar(cross = my.cross,
-								 mean.formula = phenotype1 ~ sex + mean.QTL.add + mean.QTL.dom,
-								 var.formula = ~sex + var.QTL.add + var.QTL.dom)
-p <- plot(x = a1, y = a0, ylim = c(0, ymax))
-p
-ggplot2::ggsave(filename = 'images/LOD_scan_phen1.pdf', height = 3, width = 9)
-
-
-b0 <- qtl::scanone(cross = my.cross,
-									 pheno.col = 'phenotype2')
-b1 <- scanonevar(cross = my.cross,
-								 mean.formula = phenotype2 ~ sex + mean.QTL.add + mean.QTL.dom,
-								 var.formula = ~sex + var.QTL.add + var.QTL.dom)
-p <- plot(x = b1, y = b0, ylim = c(0, ymax))
-p
-ggplot2::ggsave(filename = 'images/LOD_scan_phen2.pdf', height = 3, width = 9)
-
-
-c0 <- qtl::scanone(cross = my.cross,
-									 pheno.col = 'phenotype3')
[…]

### 2016-09-15 make_predictive_plots.R

@@ -1,15 +1,0 @@
-library(tidyr); library(dplyr); library(vqtl)
-
-set.seed(27599)
-
-my.cross <- sim.cross(map = sim.map(len = rep(100, 4), n.mar = 30, eq.spacing = TRUE, include.x = FALSE),
-											n.ind = 200,
-											type = 'f2')
-my.cross$pheno$sex <- rep(x = c(0, 1), each = 100)
-my.cross <- calc.genoprob(my.cross)
-
-my.cross$pheno$phenotype1 <- rnorm(n = nind(my.cross))
-my.cross$pheno$phenotype2 <- rnorm(n = nind(my.cross), mean = 0.8*my.cross$geno$`1`$data[,15])
-my.cross$pheno$phenotype3 <- rnorm(n = nind(my.cross), sd = my.cross$geno$`2`$data[,15])
-my.cross$pheno$phenotype4 <- rnorm(n = nind(my.cross), mean = my.cross$geno$`3`$data[,15], sd = my.cross$geno$`3`$data[,15])
-

### 2016-09-15 scripts/8_make_mean_var_plots_appendix.R

@@ -1,0 +1,32 @@
+xlims <- c(-0.3, 0.5)
+ylims <- c(0.6, 1.6)
+
+plot(p1 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype1x',
+																		 focal.covariate.names = 'D1M1',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p2 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype2x',
+																		 focal.covariate.names = 'D1M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p3 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype3x',
+																		 focal.covariate.names = 'D2M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p4 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype4',
+																		 focal.covariate.names = 'D3M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+
+ggplot2::ggsave(plot = p1, filename = 'images/mean_var_plot_phen1x.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p2, filename = 'images/mean_var_plot_phen2x.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p3, filename = 'images/mean_var_plot_phen3x.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p4, filename = 'images/mean_var_plot_phen4x.pdf', height = 3, width = 4)

### 2016-09-15 scripts/copy_over_bib.sh

@@ -1,1 +1,0 @@
-cp ~/Dropbox\ \(ValdarLab\)/bibtex/10_Aim1-2016_G3_PackageVQTL.bib .

### 2016-09-15 scripts/genome_scans_for_manuscript.R

@@ -1,0 +1,79 @@
+args <- commandArgs(trailingOnly = TRUE)
+focal.phen.name <- args[1]
+
+#### SIMULATE CROSS AND PHENOTYPES ####
+library(tidyr); library(dplyr); library(qtl); library(vqtl);
+set.seed(27599)
+
+# simulate the cross
+my.cross <- qtl::sim.cross(map = qtl::sim.map(len = rep(100, 3), n.mar = 11, eq.spacing = TRUE, include.x = FALSE, anchor.tel = TRUE),
+													 n.ind = 400,
+													 type = 'f2')
+my.cross$pheno$sex <- rep(x = c(0, 1), each = 200)
+my.cross$pheno$batch <- sample(x = 10, size = 400, replace = TRUE)
+my.cross <- qtl::calc.genoprob(cross = my.cross, step = 2)
+batch.effects <- runif(n = 10, min = -0.5, max = 0.5)
+
+# simulate phenotype1 through phenotype4, for the body of the paper
+my.cross$pheno$phenotype1 <- rnorm(n = qtl::nind(my.cross),
+																	 mean = 0,
+																	 sd = 1)
+my.cross$pheno$phenotype2 <- rnorm(n = qtl::nind(my.cross),
+																	 mean = 0.28*(my.cross$geno$`1`$data[,6] - 2),
+																	 sd = 1)
+my.cross$pheno$phenotype3 <- rnorm(n = qtl::nind(my.cross),
+																	 mean = 0,
+																	 sd = exp(0.23*(my.cross$geno$`2`$data[,6] - 2)))
+my.cross$pheno$phenotype4 <- rnorm(n = qtl::nind(my.cross),
+																	 mean = 0.24*(my.cross$geno$`3`$data[,6] - 2),
+																	 sd = exp(0.16*(my.cross$geno$`3`$data[,6] - 2)))
+
+# simulate phenotype1x through phenotype4x, for the appendix
+my.cross$pheno$phenotype1x <- rnorm(n = qtl::nind(my.cross),
+																		mean = 0,
+																		sd = exp(my.cross$pheno$sex - 0.5))
+my.cross$pheno$phenotype2x <- rnorm(n = qtl::nind(my.cross),
+																		mean = 0.48*(my.cross$geno$`1`$data[,6] - 2),
+																		sd = exp(0.5*batch.effects[my.cross$pheno$batch]))
+my.cross$pheno$phenotype3x <- rnorm(n = qtl::nind(my.cross),
[…]

### 2016-09-15 scripts/make_LOD_score_scans_figure.R

@@ -1,52 +1,0 @@
-library(vqtl)
-
-set.seed(27599)
-
-my.cross <- sim.cross(map = sim.map(len = rep(100, 4), n.mar = 30, eq.spacing = TRUE, include.x = FALSE),
-											n.ind = 200,
-											type = 'f2')
-my.cross$pheno$sex <- rep(x = c(0, 1), each = 100)
-my.cross <- calc.genoprob(my.cross)
-
-my.cross$pheno$phenotype1 <- rnorm(n = nind(my.cross))
-my.cross$pheno$phenotype2 <- rnorm(n = nind(my.cross), mean = 0.8*my.cross$geno$`1`$data[,15])
-my.cross$pheno$phenotype3 <- rnorm(n = nind(my.cross), sd = my.cross$geno$`2`$data[,15])
-my.cross$pheno$phenotype4 <- rnorm(n = nind(my.cross), mean = my.cross$geno$`3`$data[,15], sd = my.cross$geno$`3`$data[,15])
-
-
-a1 <- scanonevar(cross = my.cross,
-								 mean.formula = 'phenotype1 ~ sex + mean.QTL.add + mean.QTL.dom',
-								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
-								 return.models = TRUE)$varscan
-a2 <- scanone(cross = my.cross,
-							pheno.col = 'phenotype1')
-
-b1 <- scanonevar(cross = my.cross,
-								 mean.formula = 'phenotype2 ~ sex + mean.QTL.add + mean.QTL.dom',
-								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
-								 return.models = TRUE)$varscan
-b2 <- scanone(cross = my.cross,
-							pheno.col = 'phenotype2')
-
-c1 <- scanonevar(cross = my.cross,
-								 mean.formula = 'phenotype3 ~ sex + mean.QTL.add + mean.QTL.dom',
-								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
-								 return.models = TRUE)$varscan
-c2 <- scanone(cross = my.cross,
-							pheno.col = 'phenotype3')
-
-d1 <- scanonevar(cross = my.cross,
-								 mean.formula = 'phenotype4 ~ sex + mean.QTL.add + mean.QTL.dom',
-								 var.formula = '~sex + var.QTL.add + var.QTL.dom',
-								 return.models = TRUE)$varscan
[…]

### 2016-09-15 scripts/make_mean_var_plots.R

@@ -1,0 +1,68 @@
+#### MEAN_VAR PLOTS FOR BODY OF PAPER ####
+xlims <- c(-0.5, 0.3)
+ylims <- c(0.6, 1.5)
+
+plot(p1 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype1',
+																		 focal.covariate.names = 'D1M1',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p2 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype2',
+																		 focal.covariate.names = 'D1M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p3 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype3',
+																		 focal.covariate.names = 'D2M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p4 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype4',
+																		 focal.covariate.names = 'D3M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+
+ggplot2::ggsave(plot = p1, filename = 'images/mean_var_plot_phen1.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p2, filename = 'images/mean_var_plot_phen2.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p3, filename = 'images/mean_var_plot_phen3.pdf', height = 3, width = 4)
+ggplot2::ggsave(plot = p4, filename = 'images/mean_var_plot_phen4.pdf', height = 3, width = 4)
+
+
+#### MEAN VAR PLOTS FOR APPENDIX ####
+xlims <- c(-0.4, 0.5)
+ylims <- c(0.6, 1.6)
+
+plot(p1 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype1x',
+																		 focal.covariate.names = 'D1M1',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
+plot(p2 <- mean_var_plot_model_based(cross = my.cross,
+																		 phenotype.name = 'phenotype2x',
+																		 focal.covariate.names = 'D1M6',
+																		 xlim = xlims,
+																		 ylim = ylims))
+
[…]

### 2016-09-15 2016_G3_PackageVQTL_Corty.tex
added some clarifying prose
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 
-One assumption that has been constant throughout all these advances is that the extent of residual variation is constant across all organisms in a study population.
-Said another way, it has always been assumed that no environmental factors and no genetic factors influence the extent of residual phenotype variation.
+One common assumption throughout is that the extent of residual variation is constant across all organisms in a study population.
+Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the extent of residual phenotype variation.
 In the companion piece of this article, we introduced a statistical modeling approach that accommodates heterogeneity in residual variation within a study population.
@@
 \begin{enumerate}
-	\item It allows for the detection of genetic loci that influence residual phenotype variation, which are likely to play central roles in the network of molecular interactions that gives rise to a complex trait.
+	\item It allows for the detection of genetic loci that influence the extent of residual variation, which are likely to play central roles in the network of molecular interactions that gives rise to a complex trait.
 	\item It allows for accurate consideration of the quantity of information provided by each individual.
-Individuals with less residual phenotype variance provide more information about the mean of the groups to which they belong, as is evident in the standard equation for the standard error of the mean: $\text{SE}_\mu = \sigma/\sqrt{n}$.
+Individuals with less residual variation provide more information about the mean of the groups to which they belong, as is evident in the formula for the standard error of the mean of a normally-distributed random variable: $\text{SE}_\mu = \sigma/\sqrt{n}$.
 \end{enumerate}
 
-The companion piece of this article is based on our reanalysis of an F2 intercross of two mouse strains carried out and published in 2008, but some of the results are completely novel.
-In support of other researchers who may be interested in reanalysing previously-conducted mapping studies using the simultaneous mean-variance mapping approach, we have composed an R package that provides functions for conducting mean-variance genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
-R package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
+Using the ``simultaneous mean-variance'' approach, we reanalyzed an F2 intercross of two inbred mouse strains that was carried out and published in 2008.
+This reanalysis uncovered some novel results.
+In support of other researchers who may be interested in reanalysing previously-conducted mapping studies using the simultaneous mean-variance approach, we have composed \texttt{R} package \texttt{vqtl}, which provides functions for conducting mean-variance genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
+\texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
 
-Here, we demonstrate typical usage of the \texttt{vqtl} package.
-The code used to generate all statistics and figures in this paper is available at \texttt{github.com/rcorty}.
+Here, we demonstrate typical use of the \texttt{vqtl} package.
+The code used to simulate the phenotypes, calculate the statistics, estimate their significance, and visualize significant results is available at \texttt{github.com/rcorty}.
 
@@
 
-\section*{Conduct a Genome Scans}
+\section*{Conduct Genome Scans}
 
@@
 
-\begin{figure}[ht!]
+\begin{figure}
     \begin{subfigure}{0.5\textwidth}
@@
 The LOD score is the traditional test statistic in QTL mapping, but its interpretation can be difficult for two reasons:
-1) LOD scores must be interpreted differently in autosomes and sex chromosomes.  The sex chromosomes are typically fit with fewer parameters, so the LOD scores are typically lower.
-2) LOD scores from different tests must be interpreted differently.
-For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests, because they are nested tests.
+(1) LOD scores must be interpreted differently in autosomes and sex chromosomes.
+The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
+(2) LOD scores from different tests must be interpreted differently.
+For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
+This relationship is due to the nested nature of the mvQTL test and the other two tests.
+All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
 
-These difficulties in interpreting the LOD score are borne out in figure~\ref{fig:lod_score_scans}.
-Given the scale of the vertical axis, it seems that there are no important signicals in the genome scan of \texttt{phenotype1}.
-It is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
-But it is not evident how to compare the results of the mvQTL test to the results of the other tests.
-Nor would it be clear how to compare the results of tests from sex chromosomes, if they were present, to those from autosomes.
-Nor is is possible to make an appropriate multple testing correction to account for the many tests that are conducted in a genome scan.
+These difficulties in interpreting the LOD score evident in figure~\ref{fig:lod_score_scans}.
+It seems that there are no important signials in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
+But important questions remain:
+(1) If we didn't have the genome scans from \texttt{phenotype2} - \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
+(2) How can we compare the results of the mvQTL test to the results of the other tests?
+(3) How could we compare the results of tests on autosomes to tests on sex chromosomes (if they were present)?
+(4) How often do we expect to observe results of this magnitude or greater when there is no true association, due simply to sampling variation and the multiplicity of tests conducted?
 
 To put all genetic loci and all three tests on a level playing field, we consider two types of $p$-values:
-1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom.
-Though these $p$-values overcome the two problems of working with LOD scores described above, they leave unresolved the problem of interpreting $p$-values when many tests were conducted.
-2) Empirical, family-wide error rate (FWER) -corrected $p$-values resolve the multiple-testing issue, thus they are the recommended means of assessing the significance of QTL mapping results.
+(1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom for each locus.
+Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave problem 4 unresolved.
+(2) Empirical, family-wide error rate (FWER) -corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
 Methods for calculating them are described in the next section.
@@
 Calling \texttt{summary} on this object produces a summary of how the scan was conducted and what the results were.
+With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
 
@@
 
-The effective number of statistical tests conducted in a family of tests (a genome scan) typically must be estimated to control family-wide error rate (FWER).
+The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much or more evidence against the null at this locus, given that there is no true effect?''
+This question is not entirely relevant for genetic mapping, however, because we typically test for association at many loci, with no special interest in any individual locus \textit{a priori}.
+Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more evidence against the null at some locus in the genome, given that there are no true effects?''
+This is precisely the reasing behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
+
+To calculate a FWER-controlling $p$-value, one typically must estimate the effective number of statistical tests conducted.
 The effective number of tests in a genome scan, however, is difficult to estimate.
@@
 Due to the randomization in meiosis no two non-syntenic loci are correlated in an experimental cross and therefore tests on different chromosomes are always independent.
-But there are many tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
+But there are many non-identical tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
 One upper bound on the effective number of tests is the total number of loci.
@@
 Our empirical approach sidesteps estimation of the effective number of tests.
-We conduct many genomes scans, each executing a carefully-constructed model comparison to maintain all mean and variance effects of covariates and any non-focal genetic effects on mean and variance.
+We conduct many permuted genomes scans, each executing a carefully-constructed model comparison to maintain all mean and variance effects of covariates and any non-focal genetic effects on mean and variance.
 The details of these model comparisons are provided in the companion article.
@@
 The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value distribution to estimate the FWER-controlling $p$-values.
-This approach is implemented in \texttt{scanonevar.perm}.
+This approach is implemented in the function, \texttt{scanonevar.perm}.
 \texttt{scanonevar.perm} takes two required inputs:
@@
 The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with one important additional piece of information, an empirical $p$-value for each test at each locus.
-These $p$-values are FWER-corrected, so a value of $0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe \textit{some} locus this significant or more significant.
-A list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
+These $p$-values are FWER-corrected, so a value of $0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe \textit{some} locus with this much or more evidence of association.
+Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
 
-\begin{figure}[ht!]
+\begin{figure}
     \begin{subfigure}{0.5\textwidth}
@@
 When many phenotypes are studied, or the 5-50 minute runtime is problematic, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
-This function combines the permutations from all the inputted scans, re-evaluates the observed LOD scores in light of all available permutations, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values in about one second.
+This function combines the permutations from all the inputted scans, re-evaluates the observed LOD scores in the context of all available permutations, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}
     \begin{subfigure}[t]{0.24\textwidth}
@@
 This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
-There are both ``model-free'' and ``model-based'' versions of this plotting utility, but here we show only the ``model-based'' version for conciseness.
+There are both ``model-free'' and ``model-based'' versions of this plotting utility, but here we show only the ``model-based'' version here.
 
 Figure~\ref{fig:mean_var_plots} shows a model-based \texttt{mean\_var\_plot} for each of the four phenotypes.
-In each plot, the location of the dot shows the estimated mean and variance effect of each genotype, with the mean effect indicated by the horizontal position and the variance effect indicated by the vertical position.
-The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the variance estimate.
-Additional grouping factors, such as sex, treatment group, or batch can also be specified and all combinations of effects for which an observation was made are plotted, but we here we show only the allele effects for clarity.
+In each plot, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
+The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
+There are two types of grouping factors considered by the function \texttt{mean\_var\_plot\_model\_based}:
+(1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
+For example, a genetic marker is the \texttt{focal.group} in each plot in figure~\ref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
+(2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
+When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
 
-For phenotype1, the \texttt{mean\_var\_plot} is shown at the first marker of the first chromosome.
+For \texttt{phenotype1}, the \texttt{mean\_var\_plot} is shown at the first marker of the first chromosome.
 The estimates of the genotype effects on phenotype mean and variance are within one standard error of each other.
@@
 
-For phenotype2, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the first chromosome.
+For \texttt{phenotype2}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the first chromosome.
 The estimates of the genotype effects differ in the horizontal axis, but not the vertical axis.
@@
 
-For phenotype3, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the second chromosome.
+For \texttt{phenotype3}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the second chromosome.
 The estimates of the genotype effects differ in the vertical axis, but on on the horizontal axis.
@@
 
-For phenotype4, the \texttt{mean\_var\_plot} is shown at the most significant marker, the tenth marker on the third chromosome.
-The estimates fo the genotype effects differ in both the horizontal and vertical axes.
+For \texttt{phenotype4}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the third chromosome.
+The estimates fo the genotype effects differ somewhat in both the horizontal and vertical axes, but not as much on the horizontal axis as \texttt{phenotype2} and not as much on the vertical axis as \texttt{phenotype3}.
 This pattern is consistent with the fact that there is a genetic effect on both phenotype mean and variance and with the highly significant $p$-value for the mvQTL test and the marginally significant $p$-values for the other two tests.
+
+Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online vignette for the \texttt{vqtl} package, available at \texttt{github.com/rcorty}.
 
@@
 \bibliography{10_Aim1-2016_G3_PackageVQTL}
+\vfill
 

### 2016-09-15 scripts/make_mean_var_plots.R

@@ -4,25 +4,25 @@ ylims <- c(0.6, 1.5)
 
 plot(p1 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype1',
-																		 focal.covariate.names = 'D1M1',
+																		 focal.groups = 'D1M1',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p2 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype2',
-																		 focal.covariate.names = 'D1M6',
+																		 focal.groups = 'D1M6',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p3 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype3',
-																		 focal.covariate.names = 'D2M6',
+																		 focal.groups = 'D2M6',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p4 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype4',
-																		 focal.covariate.names = 'D3M6',
+																		 focal.groups = 'D3M6',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
@@ -39,25 +39,25 @@ ylims <- c(0.6, 1.6)
 
 plot(p1 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype1x',
-																		 focal.covariate.names = 'D1M1',
+																		 focal.groups = 'D1M1',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p2 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype2x',
-																		 focal.covariate.names = 'D1M6',
+																		 focal.groups = 'D1M6',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p3 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype3x',
-																		 focal.covariate.names = 'D2M6',
+																		 focal.groups = 'D2M6',
 																		 xlim = xlims,
 																		 ylim = ylims))
 
 plot(p4 <- mean_var_plot_model_based(cross = my.cross,
 																		 phenotype.name = 'phenotype4',
-																		 focal.covariate.names = 'D3M6',
[…]

### 2017-11-08 Package_vqt_Manuscript.tex
minor improvements
Signed-off-by: Robert Corty <rcorty@gmail.com>

--- 2016-09-15 2016_G3_PackageVQTL_Corty.tex
@@
-\documentclass[9pt,twocolumn,twoside]{gsag3jnl}
+\documentclass{article}
 
-\articletype{inv} % article type
+% \articletype{inv} % article type
 
+\usepackage[margin = 1in]{geometry}
+
+\usepackage[style=authoryear,natbib=true]{biblatex}
+\bibliography{../../bibtex/10_Aim1.bib}
+
+\usepackage{amsmath}
 \usepackage{bm}
@@
 
-\title{\texttt{vqtl}: An \texttt{R} package for QTL Mapping on Phenotypes with Heterogeneous Variance}
+\title{\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping}
+\author{Robert W. Corty and William Valdar}
+% \author[$\ast$]{Robert W. Corty}
+% \author[$\ast, 1$]{William Valdar}
+% \affil[$\ast$]{Department of Genetics, University of North Carolina at Chapel Hill}
 
-\author[$\ast$]{Robert W. Corty}
-\author[$\ast, 1$]{William Valdar}
-\affil[$\ast$]{Department of Genetics, University of North Carolina at Chapel Hill}
-
-\keywords{QTL mapping, variance heterogeneity}
-\runningtitle{R package vqtl} % For use in the footer
-\correspondingauthor{Robert Corty}
-\setboolean{displaycopyright}{true}
-
-\begin{abstract}
-Existing methods for QTL mapping in experimental crosses assume that the amount of residual variation is constant across all individuals.
-Many common situations violate this assumption.
-For example, female mice often have more variable phenotypes than males, some experimenters make more precise measurements than others, and specific genetic loci may influence the extent of variation between individuals.
-In all such cases, the heterogeneous variance modeling approach demonstrated here provides higher power, better protection against false positives, and allows for detection of QTL that influence phenotype variance, termed vQTL.
-The \texttt{R} package \texttt{vqtl} makes it easy for geneticists to apply the heterogeneous variance model, control family-wide error rate (FWER), and visualize and interpret their results.
-Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze the results of their experimental crosses with \texttt{vqtl}, possibly discovering new QTL.
-Here, we demonstrate typical usage.
-\end{abstract}
+% \keywords{QTL mapping, variance heterogeneity}
+% \runningtitle{R package mvqtl} % For use in the footer
+% \correspondingauthor{Robert Corty}
+% \setboolean{displaycopyright}{true}
 
@@
 \maketitle
-\thispagestyle{firststyle}
-\logomark
-\articletypemark
-\marginmark
-\firstpagefootnote
-\correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
-\vspace{-11pt}%
+% \thispagestyle{firststyle}
+% \logomark
+% \articletypemark
+% \marginmark
+% \firstpagefootnote
+% \correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
+% \vspace{-11pt}%
 
-\noindent Experimental crosses of inbred organisms have been a cornerstone of forward genetics for over a century \citep{mendel1866}.
-They have provided important insights on nearly every trait of interest in human disease, agriculture, and livestock production.
-Studies across all these diverse fields were enabled by advances in methods for efficient breeding, phenotyping \citep{Yang2014a}, genotyping \citep{Williams1990}, statistical methods \citep{Lander1989a}, and software tools \citep{Broman2003}.
+\begin{abstract}
+Most existing methods for QTL mapping in experimental crosses assume that the residual variance is constant across all individuals.
+But common situations violate this assumption.
+For many phenotypes, one sex is more variable than the other, some experimenters make more precise measurements than others, and specific genetic factors influence environmental sensitivity.
+In these cases, mean-variance QTL mapping provides higher power and better protection against false positives.
+It also allows for detection of QTL that influence phenotype variance, termed vQTL.
 
-One common assumption throughout is that the extent of residual variation is constant across all organisms in a study population.
-Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the extent of residual phenotype variation.
-In the companion piece of this article, we introduced a statistical modeling approach that accommodates heterogeneity in residual variation within a study population.
-We further demonstrated two critical benefits of using this ``simultaneous mean-variance'' modeling approach.
+We present \texttt{R} package \texttt{vqtl}.
+This package makes it easy for geneticists to apply the mean-variance QTL mapping approach, control family-wide error rate (FWER), and visualize and interpret their results.
+Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
+\end{abstract}
 
-\begin{enumerate}
-	\item It allows for the detection of genetic loci that influence the extent of residual variation, which are likely to play central roles in the network of molecular interactions that gives rise to a complex trait.
-	\item It allows for accurate consideration of the quantity of information provided by each individual.
-Individuals with less residual variation provide more information about the mean of the groups to which they belong, as is evident in the formula for the standard error of the mean of a normally-distributed random variable: $\text{SE}_\mu = \sigma/\sqrt{n}$.
-\end{enumerate}
+\noindent QTL mapping studies have provided important insights on nearly every trait of interest in human health and disease.
+Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003} have supported these discoveries.
 
-Using the ``simultaneous mean-variance'' approach, we reanalyzed an F2 intercross of two inbred mouse strains that was carried out and published in 2008.
-This reanalysis uncovered some novel results.
-In support of other researchers who may be interested in reanalysing previously-conducted mapping studies using the simultaneous mean-variance approach, we have composed \texttt{R} package \texttt{vqtl}, which provides functions for conducting mean-variance genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
+One common assumption in the design and analysis of these studies is that, across all organisms in a study population, the residual variance is constant.
+Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.
+
+% In fact, gene-gene interactions, gene-environment interactions, genetic factors that influence microenvironmental sensitivity, and environmental factors can influence residual variance.
+
+``vQTL'' analysis challenges that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Ronnegard2011a,Ronnegard2012}.
+In the companion piece [Corty2017a], we describe an elaboration of ``vQTL'' analysis, the mean-variance QTL mapping approach.
+This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
+We compare its behavior to other QTL mapping approaches, and illustrate the discovery of a new QTL from an existing study.
+
+In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
 \texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
@@
 
-\section*{Simulate an Experimental Cross}
+\section*{Simulation of an Illustrative Dataset}
 
-We used \texttt{R/qtl} to simulate the experimental cross to be analyzed.
-We simulated a population of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden markov model.
+We used the popular \texttt{R/qtl} package to simulate an example experimental cross.
+This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model.
+
 We simulated four phenotypes:
@@
 	\item \texttt{phenotype4} is influenced by the 6th (middle) marker on chromosome three.
-The marker influences both the mean and the variance of the phenotype, so it will serve as an example of a joint ``mvQTL''.
+The marker influences both the mean and the variance of the phenotype, so it will serve as an example of a mixed ``mvQTL''.
 \end{enumerate}
@@
 
-\section*{Conduct Genome Scans}
+\section*{Genome Scans}
 
-The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone}.
-Analogously, the central function for genetic mapping in package \texttt{vqtl} is \texttt{scanonevar}.
-\texttt{scanonevar} takes three required inputs:
+\begin{figure}[t]
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
+    \end{subfigure}
+    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+The traditional test is in green and globally similar to the mean test.
+\label{fig:lod_score_scans}}
+\end{figure}
+
+The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
+Analogously, the central function for genetic mapping in package \texttt{mvqtl} is \texttt{scanonevar}.
+It takes three required inputs:
 
@@
 The results of \texttt{scanonevar} on each of the four described phenotypes is shown in figure~\ref{fig:lod_score_scans}.
-The details of the null and alternative models used in each of the three tests can be found in the companion article.
+The details of the null and alternative models used in each of the three tests can be found in the companion article [Corty2017a, Corty2017b].
 
-\begin{figure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
-    \end{subfigure}
-
-    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
-The traditional test is in green and globally similar to the mean test.
-\label{fig:lod_score_scans}}
-\end{figure}
-
-The LOD score is the traditional test statistic in QTL mapping, but its interpretation can be difficult for two reasons:
+\subsection*{The LOD Score -- Problems and an Alternative}
+The traditional test statistic in QTL mapping is the LOD score, but its interpretation can be difficult for two reasons:
 (1) LOD scores must be interpreted differently in autosomes and sex chromosomes.
@@
 
-These difficulties in interpreting the LOD score evident in figure~\ref{fig:lod_score_scans}.
+These difficulties in interpreting the LOD score are evident in figure~\ref{fig:lod_score_scans}.
 It seems that there are no important signials in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
@@
 Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave problem 4 unresolved.
-(2) Empirical, family-wide error rate (FWER) -corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
+(2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
 Methods for calculating them are described in the next section.
@@
 
-The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much or more evidence against the null at this locus, given that there is no true effect?''
+\begin{figure}[t]
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.9\textwidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
+    \end{subfigure}
+
+    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
+\label{fig:empir_p_scans}}
+\end{figure}
+
+The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much of a deviation from the null at this locus, given that there is no true effect?''
 This question is not entirely relevant for genetic mapping, however, because we typically test for association at many loci, with no special interest in any individual locus \textit{a priori}.
-Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more evidence against the null at some locus in the genome, given that there are no true effects?''
-This is precisely the reasing behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
+Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more deviation from the null at \textit{some} locus in the genome, given that there are no true effects?''
+This is precisely the reasoning behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
 
@@
 One upper bound on the effective number of tests is the total number of loci.
-But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate.
+But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate \citep{Lander1989a}.
 
-Our empirical approach sidesteps estimation of the effective number of tests.
+Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, our permutation approach sidesteps estimation of the effective number of tests.
 We conduct many permuted genomes scans, each executing a carefully-constructed model comparison to maintain all mean and variance effects of covariates and any non-focal genetic effects on mean and variance.
-The details of these model comparisons are provided in the companion article.
-For each test, we extract the highest observed value of the test statistic from each permutation scan and use those to model a generalized extreme value distribution.
-The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value distribution to estimate the FWER-controlling $p$-values.
-This approach is implemented in the function, \texttt{scanonevar.perm}.
-\texttt{scanonevar.perm} takes two required inputs:
+The details of these model comparisons are provided in the companion article [Corty2017a].
+For each test, we extract the highest observed value of the test statistic from each permutation scan and use those to model a generalized extreme value (GEV) density \citep{Stephenson2002}.
+The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
+This approach is implemented in the function, \texttt{scanonevar.perm}, which requires two inputs:
 
@@
 
-\begin{figure}
-    \begin{subfigure}{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.5\textwidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
-    \end{subfigure}
-
-    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
-Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
-\label{fig:empir_p_scans}}
-\end{figure}
-
 Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
-We recommend at least 100, and rarely more than 1000.
+We recommend at least 100, and rarely more than 1000 \citep{Churchill1994,Carlborg2002}.
 These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument.
-On an Intel Core i5, running 100 permutations takes about five minutes.
-When many phenotypes are studied, or the 5-50 minute runtime is problematic, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
+On an Intel Core i5, running 100 permutations on this dataset packtakes about five minutes.
+When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
 This function combines the permutations from all the inputted scans, re-evaluates the observed LOD scores in the context of all available permutations, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
@@
 
-\begin{figure}
-    \begin{subfigure}[t]{0.24\textwidth}
+\begin{figure}[ht]
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
@@
 
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
@@
 
-    \caption{mean\_var plots show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+    \caption{\texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
 Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
@@
 This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
-There are both ``model-free'' and ``model-based'' versions of this plotting utility, but here we show only the ``model-based'' version here.
+There are both ``model-free'' and ``model-based'' versions of this plotting utility.
+Here we show only the ``model-based'' version.
 
@@
 
+% \section*{Mathematical Details}
+
+% [todo]
+
+% mean-var plot what is on x what is on y and how are SE's calculated
+
+% DGLM with other links and distribs
+
 \section*{Conclusion}
@@
 
-\bibliography{10_Aim1-2016_G3_PackageVQTL}
-\vfill
-
+\clearpage
 \newpage
@@
 \begin{figure}[ht!]
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4x.pdf}
@@
 \begin{figure}[ht!]
-    \begin{subfigure}{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3x.pdf}
@@
 
-    \begin{subfigure}[b]{0.5\textwidth}
+    \begin{subfigure}[b]{0.9\textwidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4x.pdf}
@@
 \begin{figure}[htbp]
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen1x.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen2x.pdf}
@@
 
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen3x.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.24\textwidth}
+    \begin{subfigure}[t]{0.5\textwidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen4x.pdf}

### 2017-11-08 scripts/genome_scans_for_manuscript.R

@@ -1,15 +1,12 @@
-args <- commandArgs(trailingOnly = TRUE)
-focal.phen.name <- args[1]
-
 #### SIMULATE CROSS AND PHENOTYPES ####
 library(tidyr); library(dplyr); library(qtl); library(vqtl);
+
 set.seed(27599)
 
 # simulate the cross
 my.cross <- qtl::sim.cross(map = qtl::sim.map(len = rep(100, 3), n.mar = 11, eq.spacing = TRUE, include.x = FALSE, anchor.tel = TRUE),
 													 n.ind = 400,
 													 type = 'f2')
-my.cross$pheno$sex <- rep(x = c(0, 1), each = 200)
 my.cross$pheno$batch <- sample(x = 10, size = 400, replace = TRUE)
 my.cross <- qtl::calc.genoprob(cross = my.cross, step = 2)
 batch.effects <- runif(n = 10, min = -0.5, max = 0.5)
@@ -19,61 +16,103 @@ my.cross$pheno$phenotype1 <- rnorm(n = qtl::nind(my.cross),
 																	 mean = 0,
 																	 sd = 1)
 my.cross$pheno$phenotype2 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0.28*(my.cross$geno$`1`$data[,6] - 2),
+																	 mean = 0.31*(my.cross$geno$`1`$data[,6] - 2),
 																	 sd = 1)
 my.cross$pheno$phenotype3 <- rnorm(n = qtl::nind(my.cross),
 																	 mean = 0,
-																	 sd = exp(0.23*(my.cross$geno$`2`$data[,6] - 2)))
+																	 sd = exp(0.2*(my.cross$geno$`2`$data[,6] - 2)))
 my.cross$pheno$phenotype4 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0.24*(my.cross$geno$`3`$data[,6] - 2),
-																	 sd = exp(0.16*(my.cross$geno$`3`$data[,6] - 2)))
+																	 mean = 0.23*(my.cross$geno$`3`$data[,6] - 2),
+																	 sd = exp(0.17*(my.cross$geno$`3`$data[,6] - 2)))
 
 # simulate phenotype1x through phenotype4x, for the appendix
 my.cross$pheno$phenotype1x <- rnorm(n = qtl::nind(my.cross),
 																		mean = 0,
-																		sd = exp(my.cross$pheno$sex - 0.5))
[…]

### 2017-11-10 Package_vqt_Manuscript.tex
titled intro section
Signed-off-by: Robert Corty <rcorty@gmail.com>
added tiny section on bootstrap
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 In these cases, mean-variance QTL mapping provides higher power and better protection against false positives.
-It also allows for detection of QTL that influence phenotype variance, termed vQTL.
+It also allows for detection of QTL that influence phenotype variance, termed vQTL and QTL that influence some mixture of phenotype mean and variance, termed mvQTL.
 
@@
 
-\noindent QTL mapping studies have provided important insights on nearly every trait of interest in human health and disease.
+\section*{Introduction}
+
+QTL mapping studies have provided important insights on nearly every trait of interest in human health and disease.
 Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003} have supported these discoveries.
@@
 
-``vQTL'' analysis challenges that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Ronnegard2011a,Ronnegard2012}.
-In the companion piece [Corty2017a], we describe an elaboration of ``vQTL'' analysis, the mean-variance QTL mapping approach.
-This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
-We compare its behavior to other QTL mapping approaches, and illustrate the discovery of a new QTL from an existing study.
+``vQTL'' analysis challenges that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Ronnegard2011a,Ronnegard2012,Cao2014}.
+In the companion pieces, we describe an elaboration of ``vQTL'' analysis, the mean-variance QTL mapping approach and use it to discover new QTL from existing data resources [companion B, C].
+% This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
+% We compare its behavior to other QTL mapping approaches, and illustrate the discovery of a new QTL from an existing study.
 
@@
 Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
+This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
 
 Unlike \texttt{scanone}, which only tests for association between each locus and the phenotype mean, \texttt{scanonevar} computes three tests for each locus -- association with phenotype mean, association with phenotype variance, and joint association with phenotype mean and variance.
-The statistic for each of these associations is a LOD score, the log of the ratio of the likelihood of the alternative model to the null model.
+The statistic for each of these associations is a LOD score, a the log of the ratio of the likelihood of the alternative model to the null model.
 The results of \texttt{scanonevar} on each of the four described phenotypes is shown in figure~\ref{fig:lod_score_scans}.
@@
 We recommend at least 100, and rarely more than 1000 \citep{Churchill1994,Carlborg2002}.
-These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
 On an Intel Core i5, running 100 permutations on this dataset packtakes about five minutes.
@@
 
-Having identified some QTL, we want to visualize the allele effects at those loci.
+Having identified some QTL, we want to visualize the estimated genetic and covariate effects at the QTL.
 Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
@@
 
+Finally, it is important to assess the genetic precision of a discovered QTL.
+This step allows for bioinformatic follow-up.
+The function \texttt{scanonevar.boot} implements the non-parametric and Bayesian bootstrap.
+This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
+
+In our experience, 1000 resamples is an appropriate number to establish 80\% and 90\% confidence intervals.
+With the datasets simulated here, it takes 15 minutes to run 500 bootstrap resamples on an Intel core i5.
+
 % \section*{Mathematical Details}
@@
 
-We have demonstrated typical usage of the new \texttt{vqtl} \texttt{R} package for QTL mapping in experimental crosses.
+We have demonstrated typical usage of the \texttt{vqtl} \texttt{R} package for QTL mapping in experimental crosses.
 This package is most appropriate for crosses and phenotypes where nuisance covariates or genetic factors are known or suspected to influence phenotype variance.
@@
 The central function of this package, \texttt{scanonevar}, carries out the initial genome scan.
-The permutation function, \texttt{scanonevar.perm}, executes a carefully-constructed set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
+The permutation function, \texttt{scanonevar.perm}, executes a set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
 A suite of plotting functions, e.g. the \texttt{mean\_var\_plot}, allows geneticists to evaluate the allelic and covariate effects that underlie detected signals.
+The bootstrap function, \texttt{scanonevar.boot}, executes bootstrap resampling to assess the genetic precision of a QTL and extablish a confidence interval.
+
+\printbibliography
 

### 2018-02-13 Package_vqt_Manuscript.tex
improve conciseness

@@
 \begin{abstract}
-Most existing methods for QTL mapping in experimental crosses assume that the residual variance is constant across all individuals.
-But common situations violate this assumption.
-For many phenotypes, one sex is more variable than the other, some experimenters make more precise measurements than others, and specific genetic factors influence environmental sensitivity.
-In these cases, mean-variance QTL mapping provides higher power and better protection against false positives.
-It also allows for detection of QTL that influence phenotype variance, termed vQTL and QTL that influence some mixture of phenotype mean and variance, termed mvQTL.
-
-We present \texttt{R} package \texttt{vqtl}.
-This package makes it easy for geneticists to apply the mean-variance QTL mapping approach, control family-wide error rate (FWER), and visualize and interpret their results.
+QTL mapping in experimental crosses is an important component of forward genetics.
+In a companion article, we describe the mean-variance approach to QTL mapping, which addresses both mean and variance heterogeneity.
+Here, we present \texttt{vqtl}, an \texttt{R} package that implements mean-variance QTL mapping.
+The goal of \texttt{vqtl} is to provide a straightforward tool for geneticists to perform mean-variance genome scans, assess the statistical significance of their findings, and communicate their results with informative plots.
 Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
+% % Mean-variance QTL mapping, a relatively novel approach to QTL mapping, elevates phenotype variance to the same level of consideration as phenotype mean.
+% Most existing methods for QTL mapping in experimental crosses assume that the residual variance is constant across all individuals.
+% But common situations violate this assumption.
+% For many phenotypes, one sex is more variable than the other, some experimenters make more precise measurements than others, and specific genetic factors influence environmental sensitivity.
+% In these cases, mean-variance QTL mapping provides higher power and better protection against false positives.
+% It also allows for detection of QTL that influence phenotype variance, termed vQTL and QTL that influence some mixture of phenotype mean and variance, termed mvQTL.
+% We present \texttt{R} package \texttt{vqtl}.
+% This package makes it easy for geneticists to apply the mean-variance QTL mapping approach, control family-wide error rate (FWER), and visualize and interpret their results.
 \end{abstract}
@@
 
-Here, we demonstrate typical use of the \texttt{vqtl} package.
-The code used to simulate the phenotypes, calculate the statistics, estimate their significance, and visualize significant results is available at \texttt{github.com/rcorty}.
+Here, we demonstrate typical use of the \texttt{vqtl} package to discover three QTL.
+% a QTL that influence phenotype mean (an mQTL), a QTL that influences phenotype variance (vQTL) and a QTL that influences a combination of mean and variance (an mvQTL).
+% The code used to simulate the phenotypes, calculate the statistics, estimate their significance, and visualize significant results is available at \texttt{github.com/rcorty}.
 
@@
 The results of \texttt{scanonevar} on each of the four described phenotypes is shown in figure~\ref{fig:lod_score_scans}.
-The details of the null and alternative models used in each of the three tests can be found in the companion article [Corty2017a, Corty2017b].
+The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
 
@@
 (2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
-Methods for calculating them are described in the next section.
+Statistical methods for empirical FWER control in mean-variance QTL mapping are described in [CompanionC].
 
@@
 Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
-We recommend at least 100, and rarely more than 1000 \citep{Churchill1994,Carlborg2002}.
+We recommend 1000 \citep{Churchill1994,Carlborg2002}.
 These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
-On an Intel Core i5, running 100 permutations on this dataset packtakes about five minutes.
+On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
 When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
@@
 There are both ``model-free'' and ``model-based'' versions of this plotting utility.
-Here we show only the ``model-based'' version.
+For brevity, we show only the more useful ``model-based'' version.
 
-Figure~\ref{fig:mean_var_plots} shows a model-based \texttt{mean\_var\_plot} for each of the four phenotypes.
-In each plot, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
+In each \texttt{mean\_var\_plot} in Figure~\ref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
 The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
@@
 
-Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online vignette for the \texttt{vqtl} package, available at \texttt{github.com/rcorty}.
+Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online vignette for the \texttt{vqtl} package, available at \texttt{https://CRAN.R-project.org/package=vqtl}.
 
-Finally, it is important to assess the genetic precision of a discovered QTL.
-This step allows for bioinformatic follow-up.
-The function \texttt{scanonevar.boot} implements the non-parametric and Bayesian bootstrap.
+Finally, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
+The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
 This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
+As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores.
 
-In our experience, 1000 resamples is an appropriate number to establish 80\% and 90\% confidence intervals.
-With the datasets simulated here, it takes 15 minutes to run 500 bootstrap resamples on an Intel core i5.
+We recommend 1000 resamples to establish 80\% and 90\% confidence intervals.
+With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
 
@@
 
-We have demonstrated typical usage of the \texttt{vqtl} \texttt{R} package for QTL mapping in experimental crosses.
-This package is most appropriate for crosses and phenotypes where nuisance covariates or genetic factors are known or suspected to influence phenotype variance.
+We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in experimental crosses.
+This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
 In the case of genetic factors, they can be mapped.
-In the case of nuisance covariates, they can be accommodated.
+In the case of covariates, they can be accommodated, the benefits of which are described in the companion article [BVH\_paper].
 

### 2018-02-15 plot_genome_scans.R
delete R stuff
put that in separate repo

@@ -1,0 +1,35 @@
+#### SIMULATE CROSS AND PHENOTYPES ####
+library(tidyr); library(dplyr); library(qtl); library(vqtl);
+
+r <- readRDS('scans_2017-05-07 12:09:54.RDS')
+
+for (phen.name in names(r[['sos']])) {
+
+	p <- plot(x = r[['sovs']][[phen.name]], y = r[['sos']][[phen.name]], plotting.units = 'LOD') +
+		ggplot2::coord_cartesian(ylim = c(0, 5))
+
+	ggplot2::ggsave(plot = p, height = 2.5, width = 9,
+									filename = paste0('images/LOD_scan_', phen.name,'.pdf'))
+
+	p <- plot(x = r[['sovs']][[phen.name]], y = r[['sos']][[phen.name]]) +
+		ggplot2::coord_cartesian(ylim = c(0, 2.5))
+
+	ggplot2::ggsave(plot = p, height = 2.5, width = 9,
+									filename = paste0('images/empir_p_scan_', phen.name,'.pdf'))
+}
+
+
+for (phen.name in names(r[['soxs']])) {
+
+	p <- plot(x = r[['sovxs']][[phen.name]], y = r[['soxs']][[phen.name]], plotting.units = 'LOD') +
+		ggplot2::coord_cartesian(ylim = c(0, 6))
+
+	ggplot2::ggsave(plot = p, height = 2.5, width = 9,
+									filename = paste0('images/LOD_scan_', phen.name,'.pdf'))
+
+	p <- plot(x = r[['sovxs']][[phen.name]], y = r[['soxs']][[phen.name]]) +
+		ggplot2::coord_cartesian(ylim = c(0, 2.5))
+
+	ggplot2::ggsave(plot = p, height = 2.5, width = 9,
+									filename = paste0('images/empir_p_scan_', phen.name,'.pdf'))
+}

### 2018-02-15 scripts/8_make_mean_var_plots_appendix.R

@@ -1,32 +1,0 @@
-xlims <- c(-0.3, 0.5)
-ylims <- c(0.6, 1.6)
-
-plot(p1 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype1x',
-																		 focal.covariate.names = 'D1M1',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p2 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype2x',
-																		 focal.covariate.names = 'D1M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p3 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype3x',
-																		 focal.covariate.names = 'D2M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p4 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype4',
-																		 focal.covariate.names = 'D3M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-
-ggplot2::ggsave(plot = p1, filename = 'images/mean_var_plot_phen1x.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p2, filename = 'images/mean_var_plot_phen2x.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p3, filename = 'images/mean_var_plot_phen3x.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p4, filename = 'images/mean_var_plot_phen4x.pdf', height = 3, width = 4)

### 2018-02-15 scripts/genome_scans_for_manuscript.R

@@ -1,118 +1,0 @@
-#### SIMULATE CROSS AND PHENOTYPES ####
-library(tidyr); library(dplyr); library(qtl); library(vqtl);
-
-set.seed(27599)
-
-# simulate the cross
-my.cross <- qtl::sim.cross(map = qtl::sim.map(len = rep(100, 3), n.mar = 11, eq.spacing = TRUE, include.x = FALSE, anchor.tel = TRUE),
-													 n.ind = 400,
-													 type = 'f2')
-my.cross$pheno$batch <- sample(x = 10, size = 400, replace = TRUE)
-my.cross <- qtl::calc.genoprob(cross = my.cross, step = 2)
-batch.effects <- runif(n = 10, min = -0.5, max = 0.5)
-
-# simulate phenotype1 through phenotype4, for the body of the paper
-my.cross$pheno$phenotype1 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0,
-																	 sd = 1)
-my.cross$pheno$phenotype2 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0.31*(my.cross$geno$`1`$data[,6] - 2),
-																	 sd = 1)
-my.cross$pheno$phenotype3 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0,
-																	 sd = exp(0.2*(my.cross$geno$`2`$data[,6] - 2)))
-my.cross$pheno$phenotype4 <- rnorm(n = qtl::nind(my.cross),
-																	 mean = 0.23*(my.cross$geno$`3`$data[,6] - 2),
-																	 sd = exp(0.17*(my.cross$geno$`3`$data[,6] - 2)))
-
-# simulate phenotype1x through phenotype4x, for the appendix
-my.cross$pheno$phenotype1x <- rnorm(n = qtl::nind(my.cross),
-																		mean = 0,
-																		sd = exp(0.5*batch.effects[my.cross$pheno$batch]))
-my.cross$pheno$phenotype2x <- rnorm(n = qtl::nind(my.cross),
-																		mean = 0.45*(my.cross$geno$`1`$data[,6] - 2),
-																		sd = exp(batch.effects[my.cross$pheno$batch]))
-my.cross$pheno$phenotype3x <- rnorm(n = qtl::nind(my.cross),
-																		mean = 0,
[…]

### 2018-02-15 scripts/make_mean_var_plots.R

@@ -1,68 +1,0 @@
-#### MEAN_VAR PLOTS FOR BODY OF PAPER ####
-xlims <- c(-0.5, 0.3)
-ylims <- c(0.6, 1.5)
-
-plot(p1 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype1',
-																		 focal.groups = 'D1M1',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p2 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype2',
-																		 focal.groups = 'D1M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p3 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype3',
-																		 focal.groups = 'D2M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p4 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype4',
-																		 focal.groups = 'D3M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-
-ggplot2::ggsave(plot = p1, filename = 'images/mean_var_plot_phen1.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p2, filename = 'images/mean_var_plot_phen2.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p3, filename = 'images/mean_var_plot_phen3.pdf', height = 3, width = 4)
-ggplot2::ggsave(plot = p4, filename = 'images/mean_var_plot_phen4.pdf', height = 3, width = 4)
-
-
-#### MEAN VAR PLOTS FOR APPENDIX ####
-xlims <- c(-0.4, 0.5)
-ylims <- c(0.6, 1.6)
-
-plot(p1 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype1x',
-																		 focal.groups = 'D1M1',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p2 <- mean_var_plot_model_based(cross = my.cross,
-																		 phenotype.name = 'phenotype2x',
-																		 focal.groups = 'D1M6',
-																		 xlim = xlims,
-																		 ylim = ylims))
-
-plot(p3 <- mean_var_plot_model_based(cross = my.cross,
[…]

### 2018-02-15 Package_vqt_Manuscript.tex
switch to G3 format
and cut some more text to get under 2500
add render new images and add them

@@
-\documentclass{article}
+\documentclass[9pt,twocolumn,twoside]{gsag3jnl}
 
-% \articletype{inv} % article type
-
-\usepackage[margin = 1in]{geometry}
-
-\usepackage[style=authoryear,natbib=true]{biblatex}
-\bibliography{../../bibtex/10_Aim1.bib}
-
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{tabu}
-\usepackage{graphicx}
-\usepackage{caption}
 \usepackage{subcaption}
 
-\input{styles/Corty2015commands}
+\articletype{inv} % article type
 
 \title{\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping}
-\author{Robert W. Corty and William Valdar}
-% \author[$\ast$]{Robert W. Corty}
-% \author[$\ast, 1$]{William Valdar}
-% \affil[$\ast$]{Department of Genetics, University of North Carolina at Chapel Hill}
 
-% \keywords{QTL mapping, variance heterogeneity}
-% \runningtitle{R package mvqtl} % For use in the footer
-% \correspondingauthor{Robert Corty}
-% \setboolean{displaycopyright}{true}
+\author[$\ast$,1]{Robert W. Corty}
+\author[$\ast, 1$]{William Valdar}
 
-\begin{document}
+\affil[$\ast$]{University of North Carolina at Chapel Hill, Department of Genetics}
 
-\maketitle
-% \thispagestyle{firststyle}
-% \logomark
-% \articletypemark
-% \marginmark
-% \firstpagefootnote
-% \correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
-% \vspace{-11pt}%
+\keywords{QTL mapping\\variance heterogeneity\\vQTL, mvQTL\\R, DGLM}
+
+\runningtitle{Package vqtl} % For use in the footer
+
+\runningauthor{Corty and Valdar}
+
+\graphicspath{{./images/}}
 
@@
 
+\setboolean{displaycopyright}{true}
+
+% \usepackage[margin = 1in]{geometry}
+
+% \usepackage[style=authoryear,natbib=true]{biblatex}
+% \bibliography{../../bibtex/10_Aim1.bib}
+
+% \usepackage{amsmath}
+% \usepackage{bm}
+% \usepackage{tabu}
+% \usepackage{graphicx}
+% \usepackage{caption}
+% \usepackage{subcaption}
+
+% \input{styles/Corty2015commands}
+
+% \linespread{2}
+
+\begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\logomark
+\articletypemark
+\marginmark
+\firstpagefootnote
+\correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
+\vspace{-24pt}%
+
 \section*{Introduction}
 
-QTL mapping studies have provided important insights on nearly every trait of interest in human health and disease.
+% significance of the software or resource presented
+QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health and disease.
 Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003} have supported these discoveries.
 
-One common assumption in the design and analysis of these studies is that, across all organisms in a study population, the residual variance is constant.
+% discuss novelty
+One common assumption in the design and analysis of many of these studies is that, across all organisms in a study population, the residual variance is constant.
 Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.
@@
 
-``vQTL'' analysis challenges that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Ronnegard2011a,Ronnegard2012,Cao2014}.
-In the companion pieces, we describe an elaboration of ``vQTL'' analysis, the mean-variance QTL mapping approach and use it to discover new QTL from existing data resources [companion B, C].
+Recent variance QTL (vQTL) analyses have challenged that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
+In the companion pieces, we describe an elaboration of vQTL analysis, which we term ``mean-variance QTL mapping'', and use it to discover new QTL from existing data resources [companion B, C].
 % This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
@@
 
+% overview of the software or resource
 In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
+
+% describe how researchers can access it.
 \texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
-
-Here, we demonstrate typical use of the \texttt{vqtl} package to discover three QTL.
+Here, we demonstrate typical use of the \texttt{vqtl} package to analyze four simulated traits and discover three QTL.
 % a QTL that influence phenotype mean (an mQTL), a QTL that influences phenotype variance (vQTL) and a QTL that influences a combination of mean and variance (an mvQTL).
@@
 \begin{figure}[t]
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
     \end{subfigure}
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
     \end{subfigure}
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
     \end{subfigure}
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
@@
 \begin{figure}[t]
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
@@
 
-The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much of a deviation from the null at this locus, given that there is no true effect?''
-This question is not entirely relevant for genetic mapping, however, because we typically test for association at many loci, with no special interest in any individual locus \textit{a priori}.
-Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more deviation from the null at \textit{some} locus in the genome, given that there are no true effects?''
-This is precisely the reasoning behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
+% The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much of a deviation from the null at this locus, given that there is no true effect?''
+% This question is not entirely relevant for genetic mapping, however, because we typically test for association at many loci, with no special interest in any individual locus \textit{a priori}.
+% Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more deviation from the null at \textit{some} locus in the genome, given that there are no true effects?''
+% This is precisely the reasoning behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
 
-To calculate a FWER-controlling $p$-value, one typically must estimate the effective number of statistical tests conducted.
-The effective number of tests in a genome scan, however, is difficult to estimate.
-One lower bound is the number of chromosomes.
-Due to the randomization in meiosis no two non-syntenic loci are correlated in an experimental cross and therefore tests on different chromosomes are always independent.
-But there are many non-identical tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
-One upper bound on the effective number of tests is the total number of loci.
-But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate \citep{Lander1989a}.
+% To calculate a FWER-controlling $p$-value, one typically must estimate the effective number of statistical tests conducted.
+% The effective number of tests in a genome scan, however, is difficult to estimate.
+% One lower bound is the number of chromosomes.
+% Due to the randomization in meiosis no two non-syntenic loci are correlated in an experimental cross and therefore tests on different chromosomes are always independent.
+% But there are many non-identical tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
+% One upper bound on the effective number of tests is the total number of loci.
+% But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate \citep{Lander1989a}.
 
+Given the many tests involved in a genome scan, some multiple hypothesis testing correction must be made.
+We advocate using a permutation-based approach, detailed in [CompanionC] to control the family wise error rate (FWER).
 Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, our permutation approach sidesteps estimation of the effective number of tests.
-We conduct many permuted genomes scans, each executing a carefully-constructed model comparison to maintain all mean and variance effects of covariates and any non-focal genetic effects on mean and variance.
-The details of these model comparisons are provided in the companion article [Corty2017a].
+
+This approach involves conducting many permuted genomes scans, each executing a model comparison that maintains all mean and variance effects of covariates and any non-focal genetic factors.
 For each test, we extract the highest observed value of the test statistic from each permutation scan and use those to model a generalized extreme value (GEV) density \citep{Stephenson2002}.
@@
 
-\section*{Investigate Significant Findings}
+\section*{Communicate Significant Findings}
 
 \begin{figure}[ht]
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
@@
 
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
@@
 
-    \caption{\texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+    \caption{
+        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
         Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
-More information on the interpretation of these plots is provided in the `Investigate Significant Findings' section.
-\label{fig:mean_var_plots}}
+        For \texttt{phenotype1}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+        For \texttt{phenotype2}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL.
+        For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
+        For \texttt{phenotype4}, the pattern of two dimensional separation without either total horizontal or vertical separation illustrates an mvQTL with neither mean nor variance effect strong enough to define an mQTL or vQTL.
+        \label{fig:mean_var_plots}
+    }
 \end{figure}
@@
 Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
-To investigate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
+To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
 This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
@@
 
-For \texttt{phenotype1}, the \texttt{mean\_var\_plot} is shown at the first marker of the first chromosome.
-The estimates of the genotype effects on phenotype mean and variance are within one standard error of each other.
-This pattern is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+The identification of an mQTL is consistent with horizontal distances between genotype groups that exceed the horizontal standard error bars.
+The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
+And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
 
-For \texttt{phenotype2}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the first chromosome.
-The estimates of the genotype effects differ in the horizontal axis, but not the vertical axis.
-This pattern is consistent with the fact that there is a genetic effect on phenotype mean but none on phenotype variance and with the highly significant $p$-value for the mQTL test, but non-significant $p$-value for the vQTL test.
-
-For \texttt{phenotype3}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the second chromosome.
-The estimates of the genotype effects differ in the vertical axis, but on on the horizontal axis.
-This pattern is consistent with the fact that there is a genetic effect on phenotype variance but none on phenotype mean and with the highly significant $p$-value for the vQTL test but non-significant $p$-value for the mQTL test.
-
-For \texttt{phenotype4}, the \texttt{mean\_var\_plot} is shown at the most significant marker, the sixth marker on the third chromosome.
-The estimates fo the genotype effects differ somewhat in both the horizontal and vertical axes, but not as much on the horizontal axis as \texttt{phenotype2} and not as much on the vertical axis as \texttt{phenotype3}.
-This pattern is consistent with the fact that there is a genetic effect on both phenotype mean and variance and with the highly significant $p$-value for the mvQTL test and the marginally significant $p$-values for the other two tests.
-
-Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online vignette for the \texttt{vqtl} package, available at \texttt{https://CRAN.R-project.org/package=vqtl}.
+Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
 
@@
 In the case of genetic factors, they can be mapped.
-In the case of covariates, they can be accommodated, the benefits of which are described in the companion article [BVH\_paper].
+In the case of covariates, they can be accommodated, which can increase power and control false positive rate, as described in a companion article [CompanionC].
 
-The central function of this package, \texttt{scanonevar}, carries out the initial genome scan.
-The permutation function, \texttt{scanonevar.perm}, executes a set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
-A suite of plotting functions, e.g. the \texttt{mean\_var\_plot}, allows geneticists to evaluate the allelic and covariate effects that underlie detected signals.
-The bootstrap function, \texttt{scanonevar.boot}, executes bootstrap resampling to assess the genetic precision of a QTL and extablish a confidence interval.
+\section*{Resources}
+The scripts used to create the data and plots published here are available at \url{ZENODO}.
+The software and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.
 
-\printbibliography
+% The central function of this package, \texttt{scanonevar}, carries out the initial genome scan.
+% The permutation function, \texttt{scanonevar.perm}, executes a set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
+% A suite of plotting functions, e.g. the \texttt{mean\_var\_plot}, allows geneticists to evaluate the allelic and covariate effects that underlie detected signals.
+% The bootstrap function, \texttt{scanonevar.boot}, executes bootstrap resampling to assess the genetic precision of a QTL and extablish a confidence interval.
+
+\section{Acknowledgements}
+This work was primarily funded by a National Institutes of General Medical Sciences (NIGMS) grant to WV (R01-GM104125) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
+RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.
+
+\bibliography{10_Aim1}
 
@@
 
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
 \begin{figure}[ht!]
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4x.pdf}
@@
 \begin{figure}[ht!]
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3x.pdf}
@@
 
-    \begin{subfigure}[b]{0.9\textwidth}
+    \begin{subfigure}[b]{\linewidth}
         \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4x.pdf}
@@
 \begin{figure}[htbp]
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen1x.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen2x.pdf}
@@
 
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen3x.pdf}
@@
     \hfill
-    \begin{subfigure}[t]{0.5\textwidth}
+    \begin{subfigure}[t]{0.5\linewidth}
         \includegraphics[width=\textwidth]{images/mean_var_plot_phen4x.pdf}

### 2018-02-15 Package_vqt_Manuscript.tex
improved keywords and some captions

@@
 
-\keywords{QTL mapping\\variance heterogeneity\\vQTL, mvQTL\\R, DGLM}
+\keywords{QTL mapping, vQTL, mvQTL, variance heterogeneity, DGLM, R}
 
@@
 
-    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+    \caption{For each of the four simulated phenotypes with background variance heterogeneity, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
 The traditional test is in green and globally similar to the mean test.
@@
 
-    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+    \caption{For each of the four simulated phenotypes with background variance heterogeneity, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
 Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
@@
 
-    \caption{\texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+    \caption{
+         \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
         Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
-More information on the interpretation of these plots is provided in the `Investigate Significant Findings' section.
-\label{fig:apdx_mean_var_plots}}
+        For \texttt{phenotype1x}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+        For \texttt{phenotype2x}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL on a background of variance heterogeneity.
+        For \texttt{phenotype3x}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL on a background of variance heterogeneity.
+        For \texttt{phenotype4x}, the pattern of two dimensional separation without either total horizontal or vertical separation illustrates an mvQTL with neither mean nor variance effect strong enough to define an mQTL or vQTL on a background of variance heterogeneity.
+        \label{fig:mean_var_plots}
+    }
 \end{figure}

### 2018-02-16 Package_vqt_Manuscript.tex
add zenodo DOI

@@
 \section*{Resources}
-The scripts used to create the data and plots published here are available at \url{ZENODO}.
-The software and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.
+The scripts used to create the data and plots published here are available as a public, static Zenodo repository at \url{DOI: 10.5281/zenodo.1173799}.
+The package \texttt{vqtl} and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.
 

### 2018-02-16 Package_vqt_Manuscript.tex
explain whats at Zenodo better

@@
 \section*{Resources}
-The scripts used to create the data and plots published here are available as a public, static Zenodo repository at \url{DOI: 10.5281/zenodo.1173799}.
+The scripts used to simulate genotypes and phenotypes, conduct the genome scans, and plot the results are available as a public, static Zenodo repository at \url{DOI: 10.5281/zenodo.1173799}.
 The package \texttt{vqtl} and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.

### 2018-02-22 Package_vqt_Manuscript.tex
working away

@@
 \usepackage{subcaption}
+
+\usepackage{color,xcolor,xspace}
+\input{macros}
 
@@
 \begin{abstract}
-QTL mapping in experimental crosses is an important component of forward genetics.
-In a companion article, we describe the mean-variance approach to QTL mapping, which addresses both mean and variance heterogeneity.
-Here, we present \texttt{vqtl}, an \texttt{R} package that implements mean-variance QTL mapping.
-The goal of \texttt{vqtl} is to provide a straightforward tool for geneticists to perform mean-variance genome scans, assess the statistical significance of their findings, and communicate their results with informative plots.
+We present \texttt{vqtl}, an \texttt{R} package that implements mean-variance QTL mapping.
+This approach for QTL mapping tests for genetic loci that influence the mean of the phenotype, termed mean QTL, the variance of the phenotype, termed variance QTL, or some combination of the two, termed mean-variance QTL.
+It is unique in its ability to correct for variance heterogeneity arising from nuisance factors, such as sex, batch, or housing.
+This package provides a functions for geneticists to conduct genome scans using the mean-variance approach to QTL mapping, conduct permutations to assess the statistical significance of their findings, and make informative plots to communicate their results.
 Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
@@
 
-% significance of the software or resource presented
 QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health and disease.
-Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003} have supported these discoveries.
+Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003,Mulligan2017} have supported these discoveries.
 
-% discuss novelty
 One common assumption in the design and analysis of many of these studies is that, across all organisms in a study population, the residual variance is constant.
-Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.
+Recent variance QTL (vQTL) analyses have challenged that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
+In the companion pieces, we describe an elaboration of vQTL analysis, which we term ``mean-variance QTL mapping'', and use it to discover new QTL from existing data resources [\CortyReanalysisPaper, \CortyMethodsPaper].
+
+Here, we demonstrate typical usage of \texttt{R} package \texttt{vqtl}, which implements the mean-variance approach to QTL mapping.
+First, we simulate an F2 intercross using the popular \texttt{R/qtl} package.
+For the organisms in this cross, we simulate four example phenotypes --- a phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
+We conduct a genome scan on each of the four phenotypes, using the standard interval mapping approach, and testing for mQTL, vQTL, and mvQTL using the mean-variance approach to QTL mapping.
+We plot all four tests in the units of LOD scores, and discuss some drawbacks to this traditional plotting unit.
+We conduct permutation scans and use the results of the permutation scans to put the original four tests on a level, and more easily interpretable, playing field.
+And finally, we plot the results to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
+
+% Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.
 
@@
 
-Recent variance QTL (vQTL) analyses have challenged that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
-In the companion pieces, we describe an elaboration of vQTL analysis, which we term ``mean-variance QTL mapping'', and use it to discover new QTL from existing data resources [companion B, C].
 % This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
@@
 % overview of the software or resource
-In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
+% In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
 
 % describe how researchers can access it.
-\texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
-Here, we demonstrate typical use of the \texttt{vqtl} package to analyze four simulated traits and discover three QTL.
+% \texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
+% Here, we demonstrate typical use of the \texttt{vqtl} package to analyze four simulated traits and discover three QTL.
 % a QTL that influence phenotype mean (an mQTL), a QTL that influences phenotype variance (vQTL) and a QTL that influences a combination of mean and variance (an mvQTL).

### 2018-02-23 Package_vqt_Manuscript.tex
edits
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 We present \texttt{vqtl}, an \texttt{R} package that implements mean-variance QTL mapping.
-This approach for QTL mapping tests for genetic loci that influence the mean of the phenotype, termed mean QTL, the variance of the phenotype, termed variance QTL, or some combination of the two, termed mean-variance QTL.
+This QTL mapping approach tests for genetic loci that influence the mean of the phenotype, termed mean QTL, the variance of the phenotype, termed variance QTL, or some combination of the two, termed mean-variance QTL.
 It is unique in its ability to correct for variance heterogeneity arising from nuisance factors, such as sex, batch, or housing.
-This package provides a functions for geneticists to conduct genome scans using the mean-variance approach to QTL mapping, conduct permutations to assess the statistical significance of their findings, and make informative plots to communicate their results.
+This package provides a functions for geneticists to conduct genome scans, run permutations to assess the statistical significance of their findings, and make informative plots to communicate their results.
 Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
@@
 \correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
-\vspace{-24pt}%
+\vspace{-30pt}%
 

### 2018-02-27 Package_vqt_Manuscript.tex
edits
last page still hasn’t been rewritten

Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 \correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
-\vspace{-30pt}%
+\vspace{-24pt}%
 
@@
 QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health and disease.
-Advances in breeding, phenotyping \citep{Yang2014a}, and genotyping \citep{Williams1990} model organisms as well as statistical methods \citep{Lander1989a} and software tools \citep{Broman2003,Mulligan2017} have supported these discoveries.
+Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools \citep{Broman2003,Mulligan2017} have supported these discoveries.
 
-One common assumption in the design and analysis of many of these studies is that, across all organisms in a study population, the residual variance is constant.
-Recent variance QTL (vQTL) analyses have challenged that assumption by seeking to identify genetic factors that influence the extent of residual variation \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
-In the companion pieces, we describe an elaboration of vQTL analysis, which we term ``mean-variance QTL mapping'', and use it to discover new QTL from existing data resources [\CortyReanalysisPaper, \CortyMethodsPaper].
+Traditional analyses assumed that the mapping population had homogenous variance and it was analyzed in search of ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean.
+But, recent work has challenged that homogenous variance assumption by seeking to identify an additional type of QTL; those that
+%, by driving heterogeneity of epistatic or environmental variance, 
+influence the extent of residual variation, termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
 
-Here, we demonstrate typical usage of \texttt{R} package \texttt{vqtl}, which implements the mean-variance approach to QTL mapping.
-First, we simulate an F2 intercross using the popular \texttt{R/qtl} package.
-For the organisms in this cross, we simulate four example phenotypes --- a phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
-We conduct a genome scan on each of the four phenotypes, using the standard interval mapping approach, and testing for mQTL, vQTL, and mvQTL using the mean-variance approach to QTL mapping.
-We plot all four tests in the units of LOD scores, and discuss some drawbacks to this traditional plotting unit.
-We conduct permutation scans and use the results of the permutation scans to put the original four tests on a level, and more easily interpretable, playing field.
+QTL analysis seeking to identify vQTL has recently entered the mainstream of genetic mapping efforts \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, but there remains great heterogeneity among the statistical methods used to conduct these analyses.
+We developed a standardized method for QTL mapping that can detect mQTL, vQTL, and a generalization of the two that we term ``mvQTL''.
+This approach, which we term ``mean-variance QTL mapping'', can be applied to intercross and backcross designs.
+In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity [\CortyMethodsPaper] and use it to discover new QTL from two existing data resources [\CortyReanalysisPaper]
+
+Here, we provide a practical guide to using the \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping.
+First, we simulate an F2 intercross using the popular \texttt{R/qtl} package and four phenotypes --- a phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
+We conduct a genome scan on each of the four phenotypes, using the standard interval mapping approach \citep{Lander1989a}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
+We plot the association statistics of all four tests in LOD score units and discuss some drawbacks to this traditional plotting unit.
+We conduct permutation scans and use the results of the permutation scans to put the four tests on a level, and more easily interpretable, playing field.
 And finally, we plot the results to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
@@
 
-\section*{Simulation of an Illustrative Dataset}
+\section*{Simulate an F2 Intercross and Four Traits}
 
@@
 	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
-	\item \texttt{phenotype2} is influenced by the 6th (middle) marker on chromosome one.
-The marker influences the mean of the phenotype, but not the variance, so it will serve as an example of a pure ``mQTL''.
-	\item \texttt{phenotype3} is influenced by the 6th (middle) marker on chromosome two.
-The marker influences the variance of the phenotype, but not the mean, so it will serve as an example of a pure ``vQTL''.
-	\item \texttt{phenotype4} is influenced by the 6th (middle) marker on chromosome three.
-The marker influences both the mean and the variance of the phenotype, so it will serve as an example of a mixed ``mvQTL''.
+	\item \texttt{phenotype2} has an mQTL at the center of chromosome one.
+	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
+	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
 \end{enumerate}
@@
 The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
-Analogously, the central function for genetic mapping in package \texttt{mvqtl} is \texttt{scanonevar}.
+Analogously, the central function for in package \texttt{vqtl} is \texttt{scanonevar}.
 It takes three required inputs:
@@
 \begin{enumerate}
-    \item \texttt{cross} contains the genetic and phenotypic information from an exerimental cross.  This object can be the same \texttt{cross} object used in package \texttt{qtl}.
-    \item \texttt{mean.formula} specifies the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted (\texttt{additive} and \texttt{dominance} components by default).
-The \texttt{mean.formula} uses the standard \texttt{R} formula notation.
-    \item \texttt{var.formula} specifies the covariates to be corrected for as well as the QTL terms to be fitted (\texttt{additive} and \texttt{dominance} components by default) in modeling the residual variance.
-The \texttt{var.formula} also uses the standard \texttt{R} formula notation.
+    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an exerimental cross, as defined in package \texttt{qtl}.
+    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
+    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
 \end{enumerate}
 
-Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
-Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
-This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
+% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
+% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
+% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
 
-Unlike \texttt{scanone}, which only tests for association between each locus and the phenotype mean, \texttt{scanonevar} computes three tests for each locus -- association with phenotype mean, association with phenotype variance, and joint association with phenotype mean and variance.
-The statistic for each of these associations is a LOD score, a the log of the ratio of the likelihood of the alternative model to the null model.
-The results of \texttt{scanonevar} on each of the four described phenotypes is shown in figure~\ref{fig:lod_score_scans}.
-The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
+% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
+For example, to scan a phenotype named \texttt{p1}, we run:
+\begin{verbatim}
+scanonevar(cross = test_cross,
+    mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
+    var.formula = ~ sex + var.QTL.add + var.QTL.dom)
+\end{verbatim}
+
+Unlike \texttt{scanone}, which only tests for mQTL, \texttt{scanonevar} computes a test statistic for mQTL, vQTL, and mvQTL at each locus.
+For each type of QTL, the test statistic is the LOD score, the base 10 logarithm of the ratio of the likelihood of an alternative model to that of a null model.
+For all three types of QTL, the alternative model is the same:
+\begin{align*}
+    \text{mean} &\propto \text{covariate effects} + \text{locus effects}\\
+    \text{variance} &\propto \text{covariate effects} + \text{locus effects}    
+\end{align*}
+
+The mQTL test compares that alternative against a null model that omits the locus effects on phenotype mean, whereas the vQTL test compares it with a null model that omits the locus effects on phenotype variance, and the mvQTL test omits both locus effects.
+
+The results of \texttt{scanonevar} on each of the four phenotypes can be plotted directly, with the LOD score as the measure of association (\autoref{fig:lod_score_scans}).
+% The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
+
+The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
+Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
+Calling \texttt{summary} on this object produces a summary of how the scan was conducted and what the results were.
+With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
 
 \subsection*{The LOD Score -- Problems and an Alternative}
-The traditional test statistic in QTL mapping is the LOD score, but its interpretation can be difficult for two reasons:
-(1) LOD scores must be interpreted differently in autosomes and sex chromosomes.
-The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
-(2) LOD scores from different tests must be interpreted differently.
-For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
-This relationship is due to the nested nature of the mvQTL test and the other two tests.
-All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
+The LOD score is the traditional association statistic in QTL mapping, but its interpretation can be difficult becuase it cannot be compared across tests with different degrees of freedom.
+For example, associations from a sex chromosome are not be comparable with those from autosomes when, as is common, the number of parameters is not constant.
+Another example, directly relevant here, is that the mvQTL test has more parameters than the mQTL test and the vQTL test and thus the LOD score is not comparable across those test classes.
+% (2) Another problem with the LOD score is that it relates only to a single locus.
+% Geneticists must take into account the many hypotheses that are tested.
+% When a significance threshold is established, \eg by permutation, to control family-wise error rate, it divides
 
-These difficulties in interpreting the LOD score are evident in figure~\ref{fig:lod_score_scans}.
-It seems that there are no important signials in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
+% (1) It must be interpreted differently in autosomes and sex chromosomes.
+% The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
+% (2) It's interpretation depends on which test was conducted -- mQTLLOD scores from different tests must be interpreted differently.
+% For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
+% This relationship is due to the nested nature of the mvQTL test and the other two tests, where they all share an alternative model and the mvQTL test uses a null model that combines the restrictions of the mQTL null and vQTL null.
+% (3) It relates only to a single locus.
+
+% All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
+
+These difficulties in interpreting the LOD score are evident in \autoref{fig:lod_score_scans}.
+It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
 But important questions remain:
@@
 (1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom for each locus.
-Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave problem 4 unresolved.
+Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave question 4 unresolved.
 (2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
-Statistical methods for empirical FWER control in mean-variance QTL mapping are described in [CompanionC].
-
-The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
-Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
-Calling \texttt{summary} on this object produces a summary of how the scan was conducted and what the results were.
-With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
+Statistical methods for empirical FWER control in mean-variance QTL mapping are described in [\CortyMethodsPaper].
 
@@
     \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
-Thus, a value of 3 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per thousand genome scans when there is no true effect.
+Thus, a value of 2 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per hundred genome scans when there is no true effect.
 \label{fig:empir_p_scans}}
@@
 
-Given the many tests involved in a genome scan, some multiple hypothesis testing correction must be made.
-We advocate using a permutation-based approach, detailed in [CompanionC] to control the family wise error rate (FWER).
-Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, our permutation approach sidesteps estimation of the effective number of tests.
+To calculate the empirical, FWER controlling, $p$-values of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
+% Given the many tests involved in a genome scan, some multiple hypothesis testing correction must be made.
+% We advocate using a permutation-based approach, detailed in [CompanionC] to control the family wise error rate (FWER).
+Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps estimation of the effective number of tests.
 
-This approach involves conducting many permuted genomes scans, each executing a model comparison that maintains all mean and variance effects of covariates and any non-focal genetic factors.
-For each test, we extract the highest observed value of the test statistic from each permutation scan and use those to model a generalized extreme value (GEV) density \citep{Stephenson2002}.
+In brief, this approach involves conducting many genomes scans on pseudo-null data generated through permutation to maintain as much of the character of the data as possible, while breaking the tested phenotype-genotype association.
+Specifically, the design matrix of the QTL is permuted in the mean portion of the mQTL alternative model, the variance portion of the vQTL alternative model, and in both portions of the mvQTL alternative model.
+
+For each test (mQTL, vQTL, and mvQTL), the highest observed test statistic is extracted from each permutation scan and the collection of statistics that results is used to fit a generalized extreme value (GEV) density \citep{Stephenson2002}.
 The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
@@
 
-The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with one important additional piece of information, an empirical $p$-value for each test at each locus.
-These $p$-values are FWER-corrected, so a value of $0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe \textit{some} locus with this much or more evidence of association.
-Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
+These FWER-corrected $p$-values are straightforwardly interpretable -- $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
+% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
 
@@
 When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
-This function combines the permutations from all the inputted scans, re-evaluates the observed LOD scores in the context of all available permutations, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
+This function combines the permutations from all the inputted scans, re-estimates the extreme value density, re-evaluates the observed LOD scores in the context of new extreme value density, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
 
@@
 
-In each \texttt{mean\_var\_plot} in Figure~\ref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
+In each \texttt{mean\_var\_plot} in \autoref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
 The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
@@
 (1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
-For example, a genetic marker is the \texttt{focal.group} in each plot in figure~\ref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
+For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
 (2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.

### 2018-02-27 Package_vqt_Manuscript.tex
edits

@@
 
-We additionally consider \texttt{phenotype1x} through \texttt{phenotype4x}, which have the same type of genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, but additionally have covariate effects on phenotype variance.
+We additionally consider \texttt{phenotype1x} through \texttt{phenotype4x}, which have the same type of genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, and, as an additional wrinkle, females have greater residual variance than males.
 All the same analyses and plots that are shown for \texttt{phenotype1} through \texttt{phenotype4} are shown for \texttt{phenotype1x} through \texttt{phenotype4x} in the appendix.
 
-\section*{Genome Scans}
+\section*{Scan the Genome}
 
@@
     \text{mean} &\propto \text{covariate effects} + \text{locus effects}\\
-    \text{variance} &\propto \text{covariate effects} + \text{locus effects}    
+    \text{log(variance)} &\propto \text{covariate effects} + \text{locus effects}    
 \end{align*}
@@
 The results of \texttt{scanonevar} on each of the four phenotypes can be plotted directly, with the LOD score as the measure of association (\autoref{fig:lod_score_scans}).
+Calling \texttt{summary} on the output of \texttt{scanonevar} produces a summary of how the scan was conducted and what the results were.
+With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
 % The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
 
-The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
-Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
-Calling \texttt{summary} on this object produces a summary of how the scan was conducted and what the results were.
-With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
+% The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
+% Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
 
@@
 The LOD score is the traditional association statistic in QTL mapping, but its interpretation can be difficult becuase it cannot be compared across tests with different degrees of freedom.
-For example, associations from a sex chromosome are not be comparable with those from autosomes when, as is common, the number of parameters is not constant.
+For example, associations from a sex chromosome are not be comparable with those from an autosome when, as is common, fewer parameters are used to model a locus on a sex chromosome.
 Another example, directly relevant here, is that the mvQTL test has more parameters than the mQTL test and the vQTL test and thus the LOD score is not comparable across those test classes.
@@
 These difficulties in interpreting the LOD score are evident in \autoref{fig:lod_score_scans}.
-It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one two and three, respectively.
+It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one, two, and three, respectively.
 But important questions remain:
-(1) If we didn't have the genome scans from \texttt{phenotype2} - \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
+(1) If we didn't have the genome scans from \texttt{phenotype2} -- \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
 (2) How can we compare the results of the mvQTL test to the results of the other tests?
@@
 Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave question 4 unresolved.
-(2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above issues with interpreting LOD scores, thus they are the recommended means of assessing the significance of QTL mapping results.
-Statistical methods for empirical FWER control in mean-variance QTL mapping are described in [\CortyMethodsPaper].
+(2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above questions with interpreting LOD scores so they are our preferred association statistics in most cases.
+% thus they are the recommended means of assessing the significance of QTL mapping results.
+% Statistical methods for empirical FWER control in mean-variance QTL mapping are described in \CortyMethodsPaper.
 
@@
 
-To calculate the empirical, FWER controlling, $p$-values of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
+To calculate the empirical, FWER-controlling, $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
 % Given the many tests involved in a genome scan, some multiple hypothesis testing correction must be made.
@@
 The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
-These FWER-corrected $p$-values are straightforwardly interpretable -- $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
+These FWER-corrected $p$-values are straightforwardly interpretable --- $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
 % Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
@@
 We recommend 1000 \citep{Churchill1994,Carlborg2002}.
-These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 1.
 On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
@@
 
-\begin{figure}[ht]
+Having identified some QTL, we want to visualize the estimated genetic and covariate effects at the QTL.
+Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
+To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
+This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
+There are both ``model-free'' and ``model-based'' versions of this plotting utility.
+For brevity, we show only the more useful ``model-based'' version.
+
+In each \texttt{mean\_var\_plot} in \autoref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
+The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
+There are two types of grouping factors considered by the function \texttt{mean\_var\_plot\_model\_based}:
+(1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
+For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
+(2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
+When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
+
+% The identification of an mQTL is consistent, roughly speaking, with horizontal distances between genotype groups that exceed the horizontal standard error bars.
+% The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
+% And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
+
+Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
+
+\section{Establish a Confidence Interval for the QTL}
+
+Finally, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
+The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
+This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
+As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.
+
+We recommend 1000 resamples to establish 80\% and 90\% confidence intervals.
+With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
+
+\begin{figure}[t]
     \begin{subfigure}[t]{0.5\linewidth}
@@
 
-Having identified some QTL, we want to visualize the estimated genetic and covariate effects at the QTL.
-Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
-To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
-This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
-There are both ``model-free'' and ``model-based'' versions of this plotting utility.
-For brevity, we show only the more useful ``model-based'' version.
-
-In each \texttt{mean\_var\_plot} in \autoref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
-The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
-There are two types of grouping factors considered by the function \texttt{mean\_var\_plot\_model\_based}:
-(1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
-For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
-(2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
-When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
-
-The identification of an mQTL is consistent with horizontal distances between genotype groups that exceed the horizontal standard error bars.
-The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
-And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
-
-Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
-
-Finally, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
-The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
-This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
-As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores.
-
-We recommend 1000 resamples to establish 80\% and 90\% confidence intervals.
-With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
-
 % \section*{Mathematical Details}
@@
 
-We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in experimental crosses.
+We have demonstrated typical usage of the \texttt{R} package \texttt{vqtl} for mean-variance QTL mapping in an F2 intercross.
 This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
-In the case of genetic factors, they can be mapped.
-In the case of covariates, they can be accommodated, which can increase power and control false positive rate, as described in a companion article [CompanionC].
+In the case of genetic factors, they can be mapped, as illustrated in \CortyReanalysisPaper.
+In the case of covariates, they can be accommodated, which can increase power and control false positive rate, as described in \CortyMethodsPaper.
 
@@
         For \texttt{phenotype4x}, the pattern of two dimensional separation without either total horizontal or vertical separation illustrates an mvQTL with neither mean nor variance effect strong enough to define an mQTL or vQTL on a background of variance heterogeneity.
-        \label{fig:mean_var_plots}
+        \label{fig:mean_var_plots_appendix}
     }

### 2018-02-28 Package_vqt_Manuscript.tex
edits

--- 2018-02-27 Package_vqt_Manuscript.tex
@@
 This QTL mapping approach tests for genetic loci that influence the mean of the phenotype, termed mean QTL, the variance of the phenotype, termed variance QTL, or some combination of the two, termed mean-variance QTL.
-It is unique in its ability to correct for variance heterogeneity arising from nuisance factors, such as sex, batch, or housing.
+It is unique in its ability to correct for variance heterogeneity arising not only from the QTL itself but also from nuisance factors, such as sex, batch, or housing.
 This package provides a functions for geneticists to conduct genome scans, run permutations to assess the statistical significance of their findings, and make informative plots to communicate their results.
-Because this package is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be easy for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
+Because it is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
 % % Mean-variance QTL mapping, a relatively novel approach to QTL mapping, elevates phenotype variance to the same level of consideration as phenotype mean.
@@
 QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health and disease.
-Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools \citep{Broman2003,Mulligan2017} have supported these discoveries.
+Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.
 
-Traditional analyses assumed that the mapping population had homogenous variance and it was analyzed in search of ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean.
-But, recent work has challenged that homogenous variance assumption by seeking to identify an additional type of QTL; those that
+Traditional analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
+Recent work, however, has challenged the homogenous variance assumption and sought to identify an additional type of QTL --- those that
 %, by driving heterogeneity of epistatic or environmental variance, 
-influence the extent of residual variation, termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
+influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
 
-QTL analysis seeking to identify vQTL has recently entered the mainstream of genetic mapping efforts \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, but there remains great heterogeneity among the statistical methods used to conduct these analyses.
+Although the search for vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, the statistical methods used for this purpose remain heterogeneous.
+% and sometimes idiosyncratic, with connections to methods for mapping mean QTL often lacking or unclear.
+% a clear or intuitive connection to methods used for detecting mean QTL.
+%there remains great heterogeneity among the statistical methods used to conduct these analyses.
 We developed a standardized method for QTL mapping that can detect mQTL, vQTL, and a generalization of the two that we term ``mvQTL''.
 This approach, which we term ``mean-variance QTL mapping'', can be applied to intercross and backcross designs.
-In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity [\CortyMethodsPaper] and use it to discover new QTL from two existing data resources [\CortyReanalysisPaper]
+In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper)
 
 Here, we provide a practical guide to using the \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping.
-First, we simulate an F2 intercross using the popular \texttt{R/qtl} package and four phenotypes --- a phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
-We conduct a genome scan on each of the four phenotypes, using the standard interval mapping approach \citep{Lander1989a}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
-We plot the association statistics of all four tests in LOD score units and discuss some drawbacks to this traditional plotting unit.
-We conduct permutation scans and use the results of the permutation scans to put the four tests on a level, and more easily interpretable, playing field.
-And finally, we plot the results to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
+First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
+On each phenotype we then conduct a genome scan using \WV{standard approximations to interval mapping \citep{Lander1989a}}{Surely HK and MC?
+We don't actually use the Lander mixture model.}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
+The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
+Permutation scans are used to determine empirically adjusted p-values, and using these in place of LOD scores is shown to to make the results of the four tests more comparable.
+Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
 
@@
     \end{subfigure}
-    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
+    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test --- mean, variance, and joint --- in blue, red, and black, respectively.
 The traditional test is in green and globally similar to the mean test.
@@
 The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
-Analogously, the central function for in package \texttt{vqtl} is \texttt{scanonevar}.
+Analogously, the central function for in package \texttt{vqtl} is \WV{\texttt{scanonevar}}{Did Karl originally implement this? If so, we should say something about it.}.
 It takes three required inputs:
-
 \begin{enumerate}
@@
 \end{enumerate}
-
 % Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
@@
 % This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
-
 % \RC{}{Doesn't fit.  It's already in the online code. Delete?}
@@
 \begin{align*}
-    \text{mean} &\propto \text{covariate effects} + \text{locus effects}\\
-    \text{variance} &\propto \text{covariate effects} + \text{locus effects}    
+    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
+    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
 \end{align*}
-
-The mQTL test compares that alternative against a null model that omits the locus effects on phenotype mean, whereas the vQTL test compares it with a null model that omits the locus effects on phenotype variance, and the mvQTL test omits both locus effects.
+The mQTL test compares this alternative against a null model that omits the locus effects on phenotype mean; the vQTL test compares it with a null model that omits the locus effects on phenotype variance; and the mvQTL test compares it with a null that omits both locus effects.
 

### 2018-02-28 Package_vqt_Manuscript.tex
respond to edits
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
-On each phenotype we then conduct a genome scan using \WV{standard approximations to interval mapping \citep{Lander1989a}}{Surely HK and MC?
-We don't actually use the Lander mixture model.}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
+On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
 The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
@@
 The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
-Analogously, the central function for in package \texttt{vqtl} is \WV{\texttt{scanonevar}}{Did Karl originally implement this? If so, we should say something about it.}.
+Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
 It takes three required inputs:

### 2018-02-28 Package_vqt_Manuscript.tex
edits

--- 2018-02-28 Package_vqt_Manuscript.tex
@@
 
-\section*{Simulate an F2 Intercross and Four Traits}
+\section*{Example data: Simulated F2 Intercross and Four Traits}
 
@@
 
-\subsection*{The LOD Score -- Problems and an Alternative}
-The LOD score is the traditional association statistic in QTL mapping, but its interpretation can be difficult becuase it cannot be compared across tests with different degrees of freedom.
-For example, associations from a sex chromosome are not be comparable with those from an autosome when, as is common, fewer parameters are used to model a locus on a sex chromosome.
-Another example, directly relevant here, is that the mvQTL test has more parameters than the mQTL test and the vQTL test and thus the LOD score is not comparable across those test classes.
+\subsection*{The LOD Score --- Problems and an Alternative}
+\WV{}{Edits to here so far}
+QTL mapping in model organisms is traditionally reported via the LOD score.
+However, for reporting the results of mean-variance QTL mapping, the LOD score is cumbersome because it is not comparable across tests with different degrees of freedom; in this case, the mvQTL test has twice as many degrees of freedom as the mQTL or vQTL tests, meaning that each type of test needs to be presented with its own threshold.
+
 % (2) Another problem with the LOD score is that it relates only to a single locus.

### 2018-02-28 Package_vqt_Manuscript.tex
edits

@@
 It is unique in its ability to correct for variance heterogeneity arising not only from the QTL itself but also from nuisance factors, such as sex, batch, or housing.
-This package provides a functions for geneticists to conduct genome scans, run permutations to assess the statistical significance of their findings, and make informative plots to communicate their results.
+This package provides functions to conduct genome scans, run permutations to assess the statistical significance, and make informative plots to communicate results.
 Because it is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.

### 2018-03-01 Package_vqt_Manuscript.tex
edits

@@
 Traditional analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
-Recent work, however, has challenged the homogenous variance assumption and sought to identify an additional type of QTL --- those that
+Recent work, however, has challenged the assumption of homogenous variance and sought to identify an additional type of QTL --- those that
 %, by driving heterogeneity of epistatic or environmental variance, 
@@
 
-Although the search for vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, the statistical methods used for this purpose remain heterogeneous.
+Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, the statistical methods used for this purpose remain heterogeneous.
 % and sometimes idiosyncratic, with connections to methods for mapping mean QTL often lacking or unclear.
@@
 
-We used the popular \texttt{R/qtl} package to simulate an example experimental cross.
+To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using the popular \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
 This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model.
-
-We simulated four phenotypes:
-
+We then generated four phenotypes:
 \begin{enumerate}
@@
 
-We additionally consider \texttt{phenotype1x} through \texttt{phenotype4x}, which have the same type of genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, and, as an additional wrinkle, females have greater residual variance than males.
+We additionally consider \texttt{phenotype1x} through \texttt{phenotype4x}, which have the same type of genetic effects as \texttt{phenotype1} through \texttt{phenotype4}, but have the additional feature that females have greater residual variance than males.
 All the same analyses and plots that are shown for \texttt{phenotype1} through \texttt{phenotype4} are shown for \texttt{phenotype1x} through \texttt{phenotype4x} in the appendix.
@@
 \section*{Scan the Genome}
+
+The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
+Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
+It takes three required inputs:
+\begin{enumerate}
+    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an exerimental cross, as defined in package \texttt{qtl}.
+    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
+    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
+\end{enumerate}
+% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
+% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
+% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
+% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
+For example, to scan a phenotype named \texttt{p1}, we run:
+\begin{verbatim}
+scanonevar(cross = test_cross,
+    mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
+    var.formula = ~ sex + var.QTL.add + var.QTL.dom)
+\end{verbatim}
+At each locus in turn, this function tests for the presence of an mQTL, a vQTL, and an mvQTL. The basis of these tests is a comparison between the fit of an alternative model of the form
+\begin{align*}
+    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
+    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
+\end{align*}
+with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
+\WV{(Note that the mQTL test is different from the traditional test for mean-affecting QTL: in the traditional test does not have variance predictors of any kind in either null or alternative models.)}{Robert: the scans show the traditional test --- is this traditional test performed as standard?}
+
+\subsection{LOD scores and nominal p-values}
+
+Each type of test (mQTL, vQTL, and mvQTL) yeilds two association statistics: the LOD score, and the (nominal) p-value.
+The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
+Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
+The p-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\text{LR}$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
+With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
+
+The p-values described above, however, are nominal: they do not take into account multiple testing across the genome.
+They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
+More robust p-values that are corrected for genomewide significance via control of the familywise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
+
+\subsection{Robust, genomewide-adjusted p-values}
+
+To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
+Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.
+
+In brief, this approach involves conducting many genomes scans on pseudo-null data generated through permutation to maintain as much of the character of the data as possible, while breaking the tested phenotype-genotype association.
+Specifically, the design matrix of the QTL is permuted in the mean portion of the mQTL alternative model, the variance portion of the vQTL alternative model, and in both portions of the mvQTL alternative model.
+
+For each test (mQTL, vQTL, and mvQTL), the highest observed test statistic is extracted from each permutation scan and the collection of statistics that results is used to fit a generalized extreme value (GEV) density \citep{Stephenson2002,Dudbridge2004,Valdar06cc}.
+The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
+This approach is implemented in the function, \texttt{scanonevar.perm}, which requires two inputs:
+\begin{enumerate}
+	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed through permutation.
+	\item \texttt{n.perms} is the number of permutations to conduct.
+\end{enumerate}
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
+These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
+% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
+
+Accurate estimation of the FWER-controlled $p$-values requires many permutation scans: traditionally recommended is 1,000 (\eg, \citealt{Churchill1994,Carlborg2002}), although the efficiency gain of using the GEV rather than raw quantiles means that fewer may be adequate in practice.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 1.
+On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
+When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
+This function combines the permutations from all the inputted scans, re-estimates the extreme value density, re-evaluates the observed LOD scores in the context of new extreme value density, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
+
+\subsection*{Reporting and plotting genome scans}
 
@@
 \end{figure}
-
-The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
-Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
-It takes three required inputs:
-\begin{enumerate}
-    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an exerimental cross, as defined in package \texttt{qtl}.
-    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
-    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
-\end{enumerate}
-% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
-% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
-% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
-% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
-For example, to scan a phenotype named \texttt{p1}, we run:
-\begin{verbatim}
-scanonevar(cross = test_cross,
-    mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
-    var.formula = ~ sex + var.QTL.add + var.QTL.dom)
-\end{verbatim}
-
-Unlike \texttt{scanone}, which only tests for mQTL, \texttt{scanonevar} computes a test statistic for mQTL, vQTL, and mvQTL at each locus.
-For each type of QTL, the test statistic is the LOD score, the base 10 logarithm of the ratio of the likelihood of an alternative model to that of a null model.
-For all three types of QTL, the alternative model is the same:
-\begin{align*}
-    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
-    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
-\end{align*}
-The mQTL test compares this alternative against a null model that omits the locus effects on phenotype mean; the vQTL test compares it with a null model that omits the locus effects on phenotype variance; and the mvQTL test compares it with a null that omits both locus effects.
-
-The results of \texttt{scanonevar} on each of the four phenotypes can be plotted directly, with the LOD score as the measure of association (\autoref{fig:lod_score_scans}).
-Calling \texttt{summary} on the output of \texttt{scanonevar} produces a summary of how the scan was conducted and what the results were.
-With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
-% The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
-
-% The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
-% Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
-
-\subsection*{The LOD Score --- Problems and an Alternative}
-\WV{}{Edits to here so far}
-QTL mapping in model organisms is traditionally reported via the LOD score.
-However, for reporting the results of mean-variance QTL mapping, the LOD score is cumbersome because it is not comparable across tests with different degrees of freedom; in this case, the mvQTL test has twice as many degrees of freedom as the mQTL or vQTL tests, meaning that each type of test needs to be presented with its own threshold.
-
-% (2) Another problem with the LOD score is that it relates only to a single locus.
-% Geneticists must take into account the many hypotheses that are tested.
-% When a significance threshold is established, \eg by permutation, to control family-wise error rate, it divides
-
-% (1) It must be interpreted differently in autosomes and sex chromosomes.
-% The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
-% (2) It's interpretation depends on which test was conducted -- mQTLLOD scores from different tests must be interpreted differently.
-% For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
-% This relationship is due to the nested nature of the mvQTL test and the other two tests, where they all share an alternative model and the mvQTL test uses a null model that combines the restrictions of the mQTL null and vQTL null.
-% (3) It relates only to a single locus.
-
-% All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
-
-These difficulties in interpreting the LOD score are evident in \autoref{fig:lod_score_scans}.
-It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one, two, and three, respectively.
-But important questions remain:
-(1) If we didn't have the genome scans from \texttt{phenotype2} -- \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
-(2) How can we compare the results of the mvQTL test to the results of the other tests?
-(3) How could we compare the results of tests on autosomes to tests on sex chromosomes (if they were present)?
-(4) How often do we expect to observe results of this magnitude or greater when there is no true association, due simply to sampling variation and the multiplicity of tests conducted?
-
-To put all genetic loci and all three tests on a level playing field, we consider two types of $p$-values:
-(1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom for each locus.
-Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave question 4 unresolved.
-(2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above questions with interpreting LOD scores so they are our preferred association statistics in most cases.
-% thus they are the recommended means of assessing the significance of QTL mapping results.
-% Statistical methods for empirical FWER control in mean-variance QTL mapping are described in \CortyMethodsPaper.
-
-\section*{Assess the Significance of Results}
 
@@
 
+The results of \texttt{scanonevar} on each of the four phenotypes can be plotted by calling \texttt{plot} on the \texttt{scanonevar} output object.
+This produces a publication-quality figure that shows the association of the phenotype for each location in the genome as different colors for type of test, with y-axis scale being specified by the user, via option \WV{XXX}{} as the LOD (\autoref{fig:lod_score_scans}), nominal p-value, or, provided permutations have been run, FWER-controlled p-value (\autoref{fig:empir_p_scans}).
+Of the available y-axis scales, we recommend using the FWER-controlled p-values since this scale puts all tests on a level-footing (unlike the LOD), and allows direct identification of genomewide significance and thereby relevance (unlike the nominal p-value).
+
+Calling \texttt{summary} on the output of \texttt{scanonevar} produces a summary of how the scan was conducted and what the results were.
+
+% The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
+
+% The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
+% Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
+
+% \subsection*{The LOD Score --- Problems and an Alternative}
+% \WV{}{Edits to here so far}
+% QTL mapping in model organisms is traditionally reported via the LOD score.
+However, for reporting the results of mean-variance QTL mapping, the LOD score is cumbersome because it is not comparable across tests with different degrees of freedom; in this case, the mvQTL test has twice as many degrees of freedom as the mQTL or vQTL tests, meaning that each type of test needs to be presented with its own threshold.
+
+% (2) Another problem with the LOD score is that it relates only to a single locus.
+% Geneticists must take into account the many hypotheses that are tested.
+% When a significance threshold is established, \eg by permutation, to control family-wise error rate, it divides
+
+% (1) It must be interpreted differently in autosomes and sex chromosomes.
+% The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
+% (2) It's interpretation depends on which test was conducted -- mQTLLOD scores from different tests must be interpreted differently.
+% For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
+% This relationship is due to the nested nature of the mvQTL test and the other two tests, where they all share an alternative model and the mvQTL test uses a null model that combines the restrictions of the mQTL null and vQTL null.
+% (3) It relates only to a single locus.
+
+% All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
+
+% These difficulties in interpreting the LOD score are evident in \autoref{fig:lod_score_scans}.
+% It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one, two, and three, respectively.
+% But important questions remain:
+% (1) If we didn't have the genome scans from \texttt{phenotype2} -- \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
+% (2) How can we compare the results of the mvQTL test to the results of the other tests?
+% (3) How could we compare the results of tests on autosomes to tests on sex chromosomes (if they were present)?
+% (4) How often do we expect to observe results of this magnitude or greater when there is no true association, due simply to sampling variation and the multiplicity of tests conducted?
+
+% To put all genetic loci and all three tests on a level playing field, we consider two types of $p$-values:
+% (1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom for each locus.
+% Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave question 4 unresolved.
+% (2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above questions with interpreting LOD scores so they are our preferred association statistics in most cases.
+% % thus they are the recommended means of assessing the significance of QTL mapping results.
+% % Statistical methods for empirical FWER control in mean-variance QTL mapping are described in \CortyMethodsPaper.
+
+%\section*{Assess the Significance of Results}
+
 % The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much of a deviation from the null at this locus, given that there is no true effect?''
@@
 
-To calculate the empirical, FWER-controlling, $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
-% Given the many tests involved in a genome scan, some multiple hypothesis testing correction must be made.
-% We advocate using a permutation-based approach, detailed in [CompanionC] to control the family wise error rate (FWER).
-Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps estimation of the effective number of tests.
-
-In brief, this approach involves conducting many genomes scans on pseudo-null data generated through permutation to maintain as much of the character of the data as possible, while breaking the tested phenotype-genotype association.
-Specifically, the design matrix of the QTL is permuted in the mean portion of the mQTL alternative model, the variance portion of the vQTL alternative model, and in both portions of the mvQTL alternative model.
-
-For each test (mQTL, vQTL, and mvQTL), the highest observed test statistic is extracted from each permutation scan and the collection of statistics that results is used to fit a generalized extreme value (GEV) density \citep{Stephenson2002}.
-The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
-This approach is implemented in the function, \texttt{scanonevar.perm}, which requires two inputs:
-
-\begin{enumerate}
-	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed through permutation.
-	\item \texttt{n.perms} is the number of permutations to conduct.
-\end{enumerate}
-
-The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
-These FWER-corrected $p$-values are straightforwardly interpretable --- $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
-% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
-
-Accurate estimation of the FWER-controlled $p$-values requires many permutation scans.
-We recommend 1000 \citep{Churchill1994,Carlborg2002}.
-These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 1.
-On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
-When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
-This function combines the permutations from all the inputted scans, re-estimates the extreme value density, re-evaluates the observed LOD scores in the context of new extreme value density, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
-
 \section*{Communicate Significant Findings}
-
-Having identified some QTL, we want to visualize the estimated genetic and covariate effects at the QTL.
-Because the \texttt{vqtl} package models both mean and variance effects, existing plotting utilities aren't able to display the entirety of the modeling results.
-To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
-This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
-There are both ``model-free'' and ``model-based'' versions of this plotting utility.
-For brevity, we show only the more useful ``model-based'' version.
-
-In each \texttt{mean\_var\_plot} in \autoref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
-The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
-There are two types of grouping factors considered by the function \texttt{mean\_var\_plot\_model\_based}:
-(1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
-For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
-(2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
-When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
-
-% The identification of an mQTL is consistent, roughly speaking, with horizontal distances between genotype groups that exceed the horizontal standard error bars.
-% The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
-% And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
-
-Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
-
-\section{Establish a Confidence Interval for the QTL}
-
-Finally, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
-The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
-This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
-As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.
-
-We recommend 1000 resamples to establish 80\% and 90\% confidence intervals.
-With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
 
@@
 
+Having identified interesting QTL, we want to visualize the their estimated genetic and covariate effects.
+Because the \texttt{vqtl} package models effects for both mean and variance, existing plotting utilities are not able to display the entirety of the modeling results.
+To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
+This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
+There are both ``model-free'' and ``model-based'' versions of this plotting utility.
+For brevity, we show only the more useful ``model-based'' version.\WV{}{Hint at what model-free means, because I can't guess}
+
+In each \texttt{mean\_var\_plot} in \autoref{fig:mean_var_plots}, the location of the dot shows the estimated mean and standard deviation of each genotype group, with the mean indicated by the horizontal position and the standard deviation indicated by the vertical position.
+The horizontal lines extending to the left and right from each dot show the standard error of the mean estimate, and the vertical lines extending up and down from each dot show the standard error of the standard deviation estimate.
+There are two types of grouping factors considered by the function \texttt{mean\_var\_plot\_model\_based}:
+(1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
+For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
+(2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
+When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
+
+% The identification of an mQTL is consistent, roughly speaking, with horizontal distances between genotype groups that exceed the horizontal standard error bars.
+% The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
+% And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
+
+Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
+
+\section{Establish a Confidence Interval for the QTL}
+
+Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
+The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
+This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
+As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.
+
+We recommend 1000 resamples to establish 80\% and 90\% confidence intervals.
+With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
+
 % \section*{Mathematical Details}
@@
 
-\bibliography{10_Aim1}
+\bibliography{10_Aim1,wv}
 

### 2018-03-01 Package_vqt_Manuscript.tex
respond to edits

@@
 This package provides functions to conduct genome scans, run permutations to assess the statistical significance, and make informative plots to communicate results.
-Because it is interoperable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
+Because it is inter-operable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
 % % Mean-variance QTL mapping, a relatively novel approach to QTL mapping, elevates phenotype variance to the same level of consideration as phenotype mean.
@@
 
-Traditional analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
-Recent work, however, has challenged the assumption of homogenous variance and sought to identify an additional type of QTL --- those that
+Traditional analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
+Recent work, however, has challenged the assumption of homogeneous variance and sought to identify an additional type of QTL --- those that
 %, by driving heterogeneity of epistatic or environmental variance, 
@@
 The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
-Permutation scans are used to determine empirically adjusted p-values, and using these in place of LOD scores is shown to to make the results of the four tests more comparable.
+Permutation scans are used to determine empirically adjusted $p$-values, and using these in place of LOD scores is shown to to make the results of the four tests more comparable.
 Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
@@
 \begin{enumerate}
-    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an exerimental cross, as defined in package \texttt{qtl}.
+    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an experimental cross, as defined in package \texttt{qtl}.
     \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
@@
 with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
-\WV{(Note that the mQTL test is different from the traditional test for mean-affecting QTL: in the traditional test does not have variance predictors of any kind in either null or alternative models.)}{Robert: the scans show the traditional test --- is this traditional test performed as standard?}
+\RC{\WV{(Note that the mQTL test is different from the traditional test QTL: the traditional test does not have variance predictors of any kind in either null or alternative models.)}{Robert: the scans show the traditional test --- is this traditional test performed as standard?}}{Nope.
+It works like: \texttt{plot(x = sov, y = so)}}
 
@@
 
-Each type of test (mQTL, vQTL, and mvQTL) yeilds two association statistics: the LOD score, and the (nominal) p-value.
+Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.
 The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
 Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
-The p-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\text{LR}$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
+The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
 With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
 
-The p-values described above, however, are nominal: they do not take into account multiple testing across the genome.
+The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
 They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
-More robust p-values that are corrected for genomewide significance via control of the familywise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
+More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
 
@@
 The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
-These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe a locus with this much or more evidence of association in this test.
+These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe some locus with this much or more evidence of association in this test.
 % Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
 
-Accurate estimation of the FWER-controlled $p$-values requires many permutation scans: traditionally recommended is 1,000 (\eg, \citealt{Churchill1994,Carlborg2002}), although the efficiency gain of using the GEV rather than raw quantiles means that fewer may be adequate in practice.
-These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 1.
+Accurate estimation of the FWER-controlled $p$-values requires many permutation scans: traditionally recommended is 1,000 (\eg, \citealt{Churchill1994,Carlborg2002}), although the efficiency gain of using the GEV rather than raw quantiles means that fewer may be adequate in practice \citep{Valdar06cc}.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
 On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
@@
 
-The results of \texttt{scanonevar} on each of the four phenotypes can be plotted by calling \texttt{plot} on the \texttt{scanonevar} output object.
-This produces a publication-quality figure that shows the association of the phenotype for each location in the genome as different colors for type of test, with y-axis scale being specified by the user, via option \WV{XXX}{} as the LOD (\autoref{fig:lod_score_scans}), nominal p-value, or, provided permutations have been run, FWER-controlled p-value (\autoref{fig:empir_p_scans}).
-Of the available y-axis scales, we recommend using the FWER-controlled p-values since this scale puts all tests on a level-footing (unlike the LOD), and allows direct identification of genomewide significance and thereby relevance (unlike the nominal p-value).
+The results of \texttt{scanonevar} can be plotted by calling \texttt{plot} on the \texttt{scanonevar} output object.
+This produces a publication-quality figure that shows the association of the phenotype for each location in the genome as different colors for type of test, with y-axis scale being specified by the user, via option \texttt{units} as the LOD (\autoref{fig:lod_score_scans}), nominal $p$-value, or, provided permutations have been run, FWER-controlled $p$-value (\autoref{fig:empir_p_scans}).
+Of the available y-axis scales, we recommend using the FWER-controlled $p$-values since this scale puts all tests on a level-footing (unlike the LOD), and allows direct identification of genomewide significance and thereby relevance (unlike the nominal $p$-value).
 
@@
 To understand and communicate the results of a \texttt{vqtl} scan at one particular locus, we developed the \texttt{mean\_var\_plot}.
-This plot shows information about the phenotype mean on the horizontal axis and information about the phenotype variance on the vertical axis.
-There are both ``model-free'' and ``model-based'' versions of this plotting utility.
-For brevity, we show only the more useful ``model-based'' version.\WV{}{Hint at what model-free means, because I can't guess}
+This plot illustrates how the mean sub-model and variance sub-model of the DGLM fit the data at a given locus.
 
@@
 (2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.
-When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that determine the mean and variance of the phenotype.
+When there are many grouping factors thought to play a role in determining the mean and variance of an individual's phenotype, such as sex, treatment, and batch, we recommend putting just one or two in \texttt{focal.groups} and the others in \texttt{nuisance.groups} for clarity, cycling through which are displayed to gain a thorough understanding of the factors that influence the phenotype.
 
@@
 Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
-The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \cite{Visscher1996}.
-This function takes, as arguments, a \texttt{scanonevar} object, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
+The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \citep{Visscher1996}.
+This function takes, as arguments, a \texttt{scanonevar} object, the type of QTL detected, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
 As with \texttt{scanonevar.perm}, the \texttt{n.cores} argument can be used to spread the bootstraps over many computational cores and defaults to the number of cores available minus two, and bootstraps can be run on separate computers and combined with \texttt{c} to increase the precision of the estimate of the confidence interval.
@@
 
-\section{Acknowledgements}
+\section{Acknowledgments}
 This work was primarily funded by a National Institutes of General Medical Sciences (NIGMS) grant to WV (R01-GM104125) and a National Institutes of Mental Health grant to RWC (F30-MH108265).

### 2018-03-01 Package_vqt_Manuscript.tex
edits

@@
 \begin{abstract}
-We present \texttt{vqtl}, an \texttt{R} package that implements mean-variance QTL mapping.
+We present \texttt{vqtl}, an \texttt{R} package for mean-variance QTL mapping.
 This QTL mapping approach tests for genetic loci that influence the mean of the phenotype, termed mean QTL, the variance of the phenotype, termed variance QTL, or some combination of the two, termed mean-variance QTL.
@@
 This package provides functions to conduct genome scans, run permutations to assess the statistical significance, and make informative plots to communicate results.
-Because it is inter-operable with the popular \texttt{R/qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{R/vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
+Because it is inter-operable with the popular \texttt{qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
 % % Mean-variance QTL mapping, a relatively novel approach to QTL mapping, elevates phenotype variance to the same level of consideration as phenotype mean.
@@
 First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
-On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and the mean-variance approach to QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
+On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
 The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
-Permutation scans are used to determine empirically adjusted $p$-values, and using these in place of LOD scores is shown to to make the results of the four tests more comparable.
+Permutation scans are used to determine empirically adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more comparable.
 Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
@@
 \begin{verbatim}
-scanonevar(cross = test_cross,
+scanonevar(
+ cross = test_cross,
  mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
-    var.formula = ~ sex + var.QTL.add + var.QTL.dom)
+ var.formula = ~ sex + var.QTL.add + var.QTL.dom
+)
 \end{verbatim}
@@
 with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
-\RC{\WV{(Note that the mQTL test is different from the traditional test QTL: the traditional test does not have variance predictors of any kind in either null or alternative models.)}{Robert: the scans show the traditional test --- is this traditional test performed as standard?}}{Nope.
-It works like: \texttt{plot(x = sov, y = so)}}
+(Note that the mQTL test in mean-variance QTL mapping is different from the traditional test: the traditional test does not have variance predictors of any kind in either null or alternative models.)
 
@@
 \end{enumerate}
-The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional piece of information, an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional pieces of information: an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
 These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe some locus with this much or more evidence of association in this test.
@@
     \end{subfigure}
-    \caption{For each of the four simulated phenotypes, the genome scan shows the LOD score of each test --- mean, variance, and joint --- in blue, red, and black, respectively.
-The traditional test is in green and globally similar to the mean test.
-\label{fig:lod_score_scans}}
+    \caption{
+        For each of the four simulated phenotypes, the genome scan shows the LOD score of each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
+        The traditional test is in green and globally similar to the mQTL test.
+    }
+    \label{fig:lod_score_scans}
 \end{figure}
@@
 
-    \caption{For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value of each test -- mean, variance, and joint -- in blue, red, and black, respectively.
-Thus, a value of 2 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per hundred genome scans when there is no true effect.
-\label{fig:empir_p_scans}}
+    \caption{
+        For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value for each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
+        The traditional test is in green and globally similar to the mQTL test.
+        A value of 2 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per hundred phenotypes no QTL.
+    }
+    \label{fig:empir_p_scans}
 \end{figure}
@@
 The results of \texttt{scanonevar} can be plotted by calling \texttt{plot} on the \texttt{scanonevar} output object.
-This produces a publication-quality figure that shows the association of the phenotype for each location in the genome as different colors for type of test, with y-axis scale being specified by the user, via option \texttt{units} as the LOD (\autoref{fig:lod_score_scans}), nominal $p$-value, or, provided permutations have been run, FWER-controlled $p$-value (\autoref{fig:empir_p_scans}).
+This produces a publication-quality figure that shows the association of the phenotype for each location in the genome as different colors for type of test, with y-axis scale being specified by the user, via option \texttt{plotting.units} as the LOD (\autoref{fig:lod_score_scans}), nominal $p$-value, or, provided permutations have been run, empirical, FWER-controlling $p$-value (\autoref{fig:empir_p_scans}).
 Of the available y-axis scales, we recommend using the FWER-controlled $p$-values since this scale puts all tests on a level-footing (unlike the LOD), and allows direct identification of genomewide significance and thereby relevance (unlike the nominal $p$-value).
@@
     \caption{
-        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus, with mean effects on the horizontal axis and variance effects on the vertical axis.
+        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus with mean effects on the horizontal axis and variance effects on the vertical axis.
         Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
@@
 (1) \texttt{focal.groups} are groups that are modeled and the prediction for each group is plotted.
-For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, etc.
+For example, a genetic marker is the \texttt{focal.group} in each plot in \autoref{fig:mean_var_plots}; \texttt{D1M1} in the top left, \texttt{D1M6} in the top right, \texttt{D2M6} in the bottom left, and \texttt{D3M6} in the bottom right.
 (2) \texttt{nuisance.groups} are groups that are modeled, but then averaged over before plotting.

### 2018-04-29 Package_vqt_Manuscript.tex
minor

@@
 \setboolean{displaycopyright}{true}
-
-% \usepackage[margin = 1in]{geometry}
-
-% \usepackage[style=authoryear,natbib=true]{biblatex}
-% \bibliography{../../bibtex/10_Aim1.bib}
-
-% \usepackage{amsmath}
-% \usepackage{bm}
-% \usepackage{tabu}
-% \usepackage{graphicx}
-% \usepackage{caption}
-% \usepackage{subcaption}
-
-% \input{styles/Corty2015commands}
 
@@
         For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
-        For \texttt{phenotype4}, the pattern of two dimensional separation without either total horizontal or vertical separation illustrates an mvQTL with neither mean nor variance effect strong enough to define an mQTL or vQTL.
+        For \texttt{phenotype4}, the pattern of two-dimensional separation illustrates an mvQTL.
         \label{fig:mean_var_plots}

### 2018-04-29 revision/response.tex
start on revision

\documentclass[11pt]{article} 
\usepackage[letterpaper, headheight=100pt]{geometry}
\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=2.5cm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{xspace}
\newcommand{\eg}{\emph{e.g.}\xspace}
\newcommand{\ie}{\emph{i.e.}\xspace}
%\newcommand{\WV}[2]{\textcolor{red}{#1}\footnote{\textcolor{red}{#2}}}
\newcommand{\WV}[2]{\textcolor{red}{#1\footnote{\textcolor{red}{#2}}}}
\newcommand{\RC}[2]{\textcolor{purple}{#1\footnote{\textcolor{purple}{#2}}}}
\newcommand{\reply}{\ensuremath{\hookrightarrow}\xspace}

% Bibliography
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
\bibliographystyle{genetics}
%\addto{\captionsenglish}{\renewcommand{\refname}{Literature Cited}}
\setlength{\bibsep}{0.0pt}

\usepackage{color}
\usepackage[usenames,dvipsnames]{xcolor}

\newcommand{\alert}[1]{\textcolor{red}{#1}}
\usepackage[left,running,mathlines]{lineno}
\renewcommand\linenumberfont{\color{gray}\normalfont\tiny\sffamily}
\usepackage{listings}

\newcommand{\auassign}[2]{\mbox{}\marginpar{\footnotesize\textbf{#1}\newline \textcolor{red}{#2}}}

\lhead{Manuscript ID\\G3/2018/200195 }
\chead{\bfseries Response to Reviewer Comments}
\rhead{\today\\William Valdar}
\title{BVH: Response To Editors/Reviewers}

\newcounter{ReviewerNum}
\setcounter{ReviewerNum}{1}

\newcounter{QuestionNum}[ReviewerNum]
\setcounter{QuestionNum}{0}

\newcounter{EditorNum}
\setcounter{EditorNum}{1}

\newcounter{EdQuestionNum}[EditorNum]
\setcounter{EdQuestionNum}{0}

\newcommand{\ReviewerQuestion}[1]{
  \vspace{5pt}\goodbreak
%  \refstepcounter{QuestionNum}
  \noindent\fbox{Reviewer:} #1
  \normalfont\par
}

\newcommand{\EditorQuestion}[1]{
  \vspace{5pt}\goodbreak
%  \refstepcounter{QuestionNum}
  \noindent\fbox{Editor:} #1
  \normalfont\par
}

\newcommand{\Response}[1]{\textcolor{blue}{#1}}

\begin{document}
\reversemarginpar

\section*{Editor Comments}

\subsection*{Multi-manuscript Issues (identical across responses to reviewers)}

\EditorQuestion{
  Now that you have all three manuscripts potentially going forward for publication in G3, they can be more tightly linked.
}
\Response{
  We have updated the manuscripts such that each one refers to the other two in the introduction and discussion.
}

\EditorQuestion{
   For instance, one the papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation [which experiment?].
  It is described in the ``benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
  (They are also in the Zenodo repository of the reanalysis paper.)
}

\subsection*{Issues Specific to this Manuscript}

\EditorQuestion{
  The main point to be addressed is the inclusion of a benchmark study demonstrating how the software scale (computational time and memory usage) with the number of SNP and sample size. But there are also many other comments that are useful and need to be addressed.
}

\EditorQuestion{
  As suggested by Reviewer 1 it will be very useful to have a benchmark study with times/memory requirements as a function of number of markers and sample size.
  This study will give those interested on using the package an understanding of how this package would scale.
  It will be important to include in the discussion also elements related to scalability: is the package suitable for very large-p/large-n? 
}
\Response{
  temp
}

\EditorQuestion{
  There is a very rich literature in the study of genetic control in environmental variance in both plant and animal breeding as well as model organisms.
  This literature should be mentioned. This comment may also apply to some of the companion papers.
}

% Mulder HA, Bijma P, Hill WG. Selection for Uniformity in Livestock by Exploiting Genetic Heterogeneity of Residual Variance. Genet Sel Evol. 2008;40:37-59.

% Hill WG, Zhang XS. Effects on Phenotypic Variability of Directional Selection Arising Through Genetic Differences in Residual Variability. Genet Res (Camb). 2004;83:121-132.

% Yang Y, Schön CC, Sorensen D. The genetics of environmental variation for dry matter grain yield in maize. Genet Res (Camb). 2012;94:113-119.

% Ibáñez N, Moreno A, Nieto B, Piqueras P, Salgado C, Gutierrez JP. Genetic Parameters Related to Environmental Variability of Weight of Mice; Signs of Correlated Canalised Response. Genet Sel Evol. 2008;40:279-293.

% Ibáñez N, Varona L, Sorensen D, Noguera JL. A study of heterogeneity of environmental variance for slaughter weight in pigs. Animal. 2007;2:19-26.

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Response to Reviewer 1 comment}

\ReviewerQuestion{
  The authors mention "in almost all traits of interest in human health and disease".
  The detection of QTL has also been widely used in plant and animal populations.
}

\Response{True.}

\ReviewerQuestion{
  The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
  I would like to see an example with a larger genotype data set.
}

\Response{Fair.}

\ReviewerQuestion{
  In the simulation, the authors should indicate the magnitude of the genetic variance explained by the QTL.
}
\Response{It's in there.}

\ReviewerQuestion{
  The CPU time is indicated for the data set of the example.
  I would like to have the CPU time information in a more realistic data set.
  Perhaps authors can use a free data set from the web.
}

\Response{addressed in Editor comment.}

\ReviewerQuestion{
  I think the authors should also indicate if the software is capable in terms of computing power to analyze the data from SNP genotyping devices or even sequencing.
}

\Response{Permutation probably not.  Could do Efron trick.}

\end{document}

### 2018-04-30 Package_vqt_Manuscript.tex
respond to simple comment

@@
 
-QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health and disease.
+QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, and livestock and crop improvement.
 Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.

### 2018-04-30 revision/response.tex
respond to simple comment
address effect size comment

@@
 \ReviewerQuestion{
-  The authors mention "in almost all traits of interest in human health and disease".
+  The authors mention ``in almost all traits of interest in human health and disease''.
   The detection of QTL has also been widely used in plant and animal populations.
@@
 
-\Response{True.}
+\Response{
+  We have edited the introduction to reflect this fact.
+}
 
@@
 }
-\Response{It's in there.}
+\Response{
+  Thank you for this recommendation.
+  We have added this information to the manuscript and it now appears on page 2.
+}
 

### 2018-04-30 Package_vqt_Manuscript.tex
address effect size comment

@@
 	\item \texttt{phenotype1} consists only of random noise and will serve as an example of negative results for all tests.
-	\item \texttt{phenotype2} has an mQTL at the center of chromosome one.
+	\item \texttt{phenotype2} has an mQTL that explains 4\% of phenotype variance at the center of chromosome one.
 	\item \texttt{phenotype3} has a vQTL at the center of chromosome two.
+        This vQTL acts additively on the log standard deviation scale, and results in residual standard deviation of [0.8, 1, 1.25] for the three genotype groups.
 	\item \texttt{phenotype4} has an mvQTL at the center of chromosome three.
+        This mvQTL has a mean effect that explains 2.7\% of phenotype variance and a variance effect that acts additively on the standard deviation scale, resulting in residual standard deviation of [0.85, 1, 1.17] for the three genotype groups.
 \end{enumerate}

### 2018-05-05 revision/response.tex
responding to reviews

@@
 
+\newcommand{\CortyRPaper}{\textcolor{purple}{Corty and Valdar 2018+ [r/vqtl]}\xspace}
+\newcommand{\CortyReanalysisPaper}{\textcolor{purple}{Corty \emph{et al} 2018+}\xspace}
+\newcommand{\CortyMethodsPaper}{\textcolor{purple}{Corty and Valdar 2018+ [BVH]}\xspace}
+
 % Bibliography
@@
 
-\lhead{Manuscript ID\\G3/2018/200195 }
+\lhead{Manuscript ID\\G3/2018/200159}
 \chead{\bfseries Response to Reviewer Comments}
@@
   \vspace{5pt}\goodbreak
-%  \refstepcounter{QuestionNum}
   \noindent\fbox{Reviewer:} #1
@@
 }
-
 \newcommand{\EditorQuestion}[1]{
   \vspace{5pt}\goodbreak
-%  \refstepcounter{QuestionNum}
   \noindent\fbox{Editor:} #1
@@
 }
-
-\newcommand{\Response}[1]{\textcolor{blue}{#1}}
+\newcommand{\Response}[1]{
+  \goodbreak
+  \textcolor{blue}{#1}
+  \normalfont\par
+}
 
@@
 \EditorQuestion{
-   For instance, one the papers could provide the benchmark data that is required for [the software] manuscript.
+   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
    So the software paper can include those files and refer to the other manuscript for details.
@@
 \Response{
-  We have incorporated this recommendation [which experiment?].
-  It is described in the ``benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
-  (They are also in the Zenodo repository of the reanalysis paper.)
+  We have incorporated this recommendation, using the dataset from Kumar et al.
+2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
+  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
+  (These files are now in two places, since they are also in the Zenodo repository of the reanalysis paper.)
 }
@@
 \subsection*{Issues Specific to this Manuscript}
-
-\EditorQuestion{
-  The main point to be addressed is the inclusion of a benchmark study demonstrating how the software scale (computational time and memory usage) with the number of SNP and sample size.
-But there are also many other comments that are useful and need to be addressed.
-}
 
@@
 \Response{
-  temp
+  We have added this information in the ``Performance Benchmarks'' section.
+  See detailed response to Reviewer 1 comment 3.
 }
@@
   This literature should be mentioned. This comment may also apply to some of the companion papers.
+  [Citation recommendations provided by the editor.]
+}
+\Response{
+  We have added a section in the introduction that describes how projects in plant and animal breeding seeking to modulate environmental variance are related to the work presented here.
 }
@@
 
-\section*{Response to Reviewer 1 comment}
+\section*{Reviewer 1 Comments}
 
@@
 }
-
 \Response{
-  We have edited the introduction to reflect this fact.
+  The reviewer is quite correct and we have edited the introduction to reflect the fact that QTL mapping has been applied productively in plant and animal research.
 }
@@
 }
+\Response{
+  In order to most clearly illustrate the standard workflow associated with package \texttt{vqtl}, the authors have demonstrated its usage on a an unrealistically-small genome.
+  The authors are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
+  This manuscript has been submitted as part of a three-paper companionship, where (call this one paper one), paper two contains two real-data reanalyses and paper three contains one real-data reanalysis.
+  Thus, the authors contend it is appropriate to maintain this paper having only limited mention of real data, those mentions in the ``Performance Benchmarks'' section.
+}
 
-\Response{Fair.}
+\ReviewerQuestion{
+  The CPU time is indicated for the data set of the example.
+  I would like to have the CPU time information in a more realistic data set.
+  Perhaps authors can use a free data set from the web.
+}
+\Response{
+  Thank you for this recommendation.
+  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package -- we have added two sets of benchmarks.
+}
+\Response{
+  First, we demonstrate the performance of the package according to number of computer cores used and number of permutations conducted.
+  For this set of benchmarks, we used the intercross from Kumar et.
+al 2013 which we reanalyzed in \CortyReanalysisPaper, which involved 244 mice, 93 markers, and after using a hidden Markov model to estimate genotype probabilities at locations between markers, 582 loci.
+  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in a companion manuscript.
+  This addition can be found on page 4.
+}
+\Response{
+  Second, we demonstrate the performance of the package according to the size of the mapping population and the number of genetic markers, holding constant the number of permutations at 1000 and the number of cores at 32.
+  For this set of benchmarks, we used crosses simulated with \texttt{R/qtl} so we could control the population size and number of genetic markers.
+  This addition can be found on page 4.
+}
 
@@
 \ReviewerQuestion{
-  The CPU time is indicated for the data set of the example.
-  I would like to have the CPU time information in a more realistic data set.
-  Perhaps authors can use a free data set from the web.
-}
-
-\Response{addressed in Editor comment.}
-
-\ReviewerQuestion{
   I think the authors should also indicate if the software is capable in terms of computing power to analyze the data from SNP genotyping devices or even sequencing.
@@
 
-\Response{Permutation probably not.  Could do Efron trick.}
+\Response{
+  Thank you for this recommendation.
+  We have addressed this question in the discussion section, on page 4.
+}
 

### 2018-05-05 Package_vqt_Manuscript.tex
responding to reviews

@@
 \usepackage{subcaption}
-
+\usepackage{multirow}
 \usepackage{color,xcolor,xspace}
@@
 
-QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, and livestock and crop improvement.
+QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
 Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.
@@
 
-\section*{Example data: Simulated F2 Intercross and Four Traits}
+\section*{Example data: Simulated F2 Intercross}
 
@@
 \section*{Scan the Genome}
-
-The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
-Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
-It takes three required inputs:
-\begin{enumerate}
-    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an experimental cross, as defined in package \texttt{qtl}.
-    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
-    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
-\end{enumerate}
-% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
-% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
-% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
-% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
-For example, to scan a phenotype named \texttt{p1}, we run:
-\begin{verbatim}
-scanonevar(
- cross = test_cross,
- mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
- var.formula = ~ sex + var.QTL.add + var.QTL.dom
-)
-\end{verbatim}
-At each locus in turn, this function tests for the presence of an mQTL, a vQTL, and an mvQTL. The basis of these tests is a comparison between the fit of an alternative model of the form
-\begin{align*}
-    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
-    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
-\end{align*}
-with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
-(Note that the mQTL test in mean-variance QTL mapping is different from the traditional test: the traditional test does not have variance predictors of any kind in either null or alternative models.)
-
-\subsection{LOD scores and nominal p-values}
-
-Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.
-The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
-Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
-The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
-With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
-
-The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
-They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
-More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
-
-\subsection{Robust, genomewide-adjusted p-values}
-
-To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
-Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.
-
-In brief, this approach involves conducting many genomes scans on pseudo-null data generated through permutation to maintain as much of the character of the data as possible, while breaking the tested phenotype-genotype association.
-Specifically, the design matrix of the QTL is permuted in the mean portion of the mQTL alternative model, the variance portion of the vQTL alternative model, and in both portions of the mvQTL alternative model.
-
-For each test (mQTL, vQTL, and mvQTL), the highest observed test statistic is extracted from each permutation scan and the collection of statistics that results is used to fit a generalized extreme value (GEV) density \citep{Stephenson2002,Dudbridge2004,Valdar06cc}.
-The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
-This approach is implemented in the function, \texttt{scanonevar.perm}, which requires two inputs:
-\begin{enumerate}
-	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed through permutation.
-	\item \texttt{n.perms} is the number of permutations to conduct.
-\end{enumerate}
-The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional pieces of information: an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
-These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe some locus with this much or more evidence of association in this test.
-% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
-
-Accurate estimation of the FWER-controlled $p$-values requires many permutation scans: traditionally recommended is 1,000 (\eg, \citealt{Churchill1994,Carlborg2002}), although the efficiency gain of using the GEV rather than raw quantiles means that fewer may be adequate in practice \citep{Valdar06cc}.
-These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
-On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
-When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
-This function combines the permutations from all the inputted scans, re-estimates the extreme value density, re-evaluates the observed LOD scores in the context of new extreme value density, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
-
-\subsection*{Reporting and plotting genome scans}
 
@@
 \end{figure}
+
+The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
+Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
+It takes three required inputs:
+\begin{enumerate}
+    \item \texttt{cross} is an object that contains the genetic and phenotypic information from an experimental cross, as defined in package \texttt{qtl}.
+    \item \texttt{mean.formula} is a two-sided formula, specifying the phenotype to be mapped, the covariates to be corrected for, and the QTL terms to be fitted, with keywords \texttt{mean.QTL.add} and \texttt{mean.QTL.dom}
+    \item \texttt{var.formula} is a one-sided formula, specifying the variance covariates to be corrected for as well as the QTL terms to be fitted, using keywords \texttt{var.QTL.add} and \texttt{var.QTL.dom}.
+\end{enumerate}
+% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
+% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
+% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
+% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
+For example, to scan a phenotype named \texttt{p1}, we run:
+\begin{verbatim}
+scanonevar(
+ cross = test_cross,
+ mean.formula = p1 ~ sex + mean.QTL.add + mean.QTL.dom,
+ var.formula = ~ sex + var.QTL.add + var.QTL.dom
+)
+\end{verbatim}
+At each locus in turn, this function tests for the presence of an mQTL, a vQTL, and an mvQTL. The basis of these tests is a comparison between the fit of an alternative model of the form
+\begin{align*}
+    \text{mean} &= \text{covariate effects} + \text{locus effects}\\
+    \log(\text{variance}) &= \text{covariate effects} + \text{locus effects}    
+\end{align*}
+with a null model that omits specific terms: for the mQTL test, the null model omits locus effects on phenotype mean; for the vQTL test, the null omits the locus effects on phenotype variance; and for the mvQTL test, the null omits locus effects on both mean and variance.
+(Note that the mQTL test in mean-variance QTL mapping is different from the traditional test: the traditional test does not have variance predictors of any kind in either null or alternative models.)
+
+\subsection{LOD scores and nominal p-values}
+
+\begin{figure}[t]
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
+    \end{subfigure}
+
+    \caption{
+        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus with mean effects on the horizontal axis and variance effects on the vertical axis.
+        Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
+        For \texttt{phenotype1}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+        For \texttt{phenotype2}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL.
+        For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
+        For \texttt{phenotype4}, the pattern of two-dimensional separation illustrates an mvQTL.
+        \label{fig:mean_var_plots}
+    }
+\end{figure}
+
+Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
+Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
+The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
+With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
+
+The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
+They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
+More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
+
+\subsection{Robust, genomewide-adjusted p-values}
+
+To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
+Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.
+
+In brief, this approach involves conducting many genomes scans on pseudo-null data generated through permutation to maintain as much of the character of the data as possible, while breaking the tested phenotype-genotype association.
+Specifically, the design matrix of the QTL is permuted in the mean portion of the mQTL alternative model, the variance portion of the vQTL alternative model, and in both portions of the mvQTL alternative model.
+
+For each test (mQTL, vQTL, and mvQTL), the highest observed test statistic is extracted from each permutation scan and the collection of statistics that results is used to fit a generalized extreme value (GEV) density \citep{Stephenson2002,Dudbridge2004,Valdar06cc}.
+The observed LOD scores from the genome scan are then transformed by the cumulative distribution function of the extreme value density to estimate the FWER-controlling $p$-values.
+This approach is implemented in the function, \texttt{scanonevar.perm}, which requires two inputs:
+\begin{enumerate}
+	\item \texttt{sov} is the \texttt{scanonevar} object, the statistical significance of which will be assessed through permutation.
+	\item \texttt{n.perms} is the number of permutations to conduct.
+\end{enumerate}
+The object returned by \texttt{scanonevar.perm} is a \texttt{scanonevar} object with two additional pieces of information: an empirical $p$-value for each test at each locus and the per-permutation maxima that were used to calculate those $p$-values.
+These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe some locus with this much or more evidence of association in this test.
+% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
+
+Accurate estimation of the FWER-controlled $p$-values requires many permutation scans: traditionally recommended is 1,000 (\eg, \citealt{Churchill1994,Carlborg2002}), although the efficiency gain of using the GEV rather than raw quantiles means that fewer may be adequate in practice \citep{Valdar06cc}.
+These permutation scans can be run on multiple processors by specifying the optional \texttt{n.cores} argument, which defaults to the total number of cores on the computer minus 2.
+On an Intel Core i5, running 100 permutations on this dataset takes about five minutes.
+When many phenotypes are studied, or if faster runtimes are needed, these permutation scans can be broken into groups with different values for \texttt{random.seed}, run on separate computers, and combined with the \texttt{c} function.
+This function combines the permutations from all the inputted scans, re-estimates the extreme value density, re-evaluates the observed LOD scores in the context of new extreme value density, and returns a new \texttt{scanonevar} object with more precisely estimated empirical $p$-values.
+
+\subsection*{Reporting and plotting genome scans}
 
@@
 \begin{figure}[t]
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
-    \end{subfigure}
-
+    \includegraphics[width = \linewidth]{images/benchmark_kumar.pdf}
     \caption{
-        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus with mean effects on the horizontal axis and variance effects on the vertical axis.
-        Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
-        For \texttt{phenotype1}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
-        For \texttt{phenotype2}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL.
-        For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
-        For \texttt{phenotype4}, the pattern of two-dimensional separation illustrates an mvQTL.
-        \label{fig:mean_var_plots}
+        Time taken to run \texttt{scanonevar.perm} on the data from \citet{Kumar2013} which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
+        For a given number of cores, there is a linear relationship between number of permutations conducted and time required.
+        The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
     }
+    \label{fig:benchmark_kumar}
 \end{figure}
@@
 
+\begin{figure}[t]
+    \includegraphics[width = \linewidth]{images/benchmark_sim_cross.pdf}
+    \caption{
+        Time taken to run 1000 permutation scans on 32 cores on simulated data using \texttt{scanonevar.perm}, varying the number of individuals in the mapping population and the number of markers in the genome.
+        For a given population size, there is a slightly supra-linear relationship between number of markers and time required.
+        The average slope of the line indicates the average time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
+    }
+    \label{fig:benchmark_sim_cross}
+\end{figure}
+
 Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
@@
 
+\section*{Performance Benchmarks}
+
+By far, the most computationally-intensive step in the mean-variance QTL mapping process is the assessment of genome-wide statistical significance by permutation.
+The original genome scan is much faster, because it involves only a single scan, and the bootstrap is much faster because it involves only a single chromosome.
+
+For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
+For a given number of cores, there is a linear relationship between number of permutations conducted and time required \autoref{fig:benchmark_kumar}.
+The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
+
+For the second benchmark, we ran \texttt{scanonevar.perm} on simulated data, always conducting 1000 permutations and using 32 cores, but varying the number of individuals in the mapping population and the number of markers in the genome.
+For a given population size, there is a slightly supra-linear relationship between number of markers and time required \autoref{fig:benchmark_sim_cross}, which reflects a linear increase in the time taken to conduct the permuted genome scans, plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
+The slope of the line indicates the time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
+
+% Performance benchmarks of package \texttt{vqtl} are presented in \autoref{fig:benchmark_kumar} and \autoref{fig:benchmark_sim_cross}.
+
+% \begin{table}
+%     \centering
+%     \begin{tabular}{r|rrrr}
+%         \multirow{2}{1in}{\raggedleft \textbf{number of permutations}}& \multicolumn{4}{c}{\textbf{number of cores}}\\
+%         \cline{2-5}
+%         & 4 & 8 & 16 & 32\\
+%         \hline
+%         100  & 11 & 6 & 4 & 4\\
+%         1000 & -  & 53 & 34 & 20\\
+%         \hline
+%     \end{tabular}
+%     \caption{
+%         Time in minutes required to conduct permuations for all three tests on the dataset from \citet{Kumar2013}, which has 244 individuals and 582 loci, according to number of permutations desired and number of computer cores available.
+%         We recommend 1000 permutations for publication-quality precision.
+%         We include the time taken to conduct 100 and 500 permutations to illustrate the linear relation between number of permutations and time required.
+%         This table also illustrates the linear relationship between number of computer cores used and time taken to complete permutations.
+%     }
+% \end{table}
+%
+% \begin{table}
+%     \centering
+%     \begin{tabular}{r|rrrr}
+%         \multirow{2}{1in}{\raggedleft \textbf{number of markers}}& \multicolumn{4}{c}{\textbf{number of individuals}}\\
+%         \cline{2-5}
+%         & 100 & 200 & 400 & 800\\
+%         \hline
+%         100 & 2 & 3 & 4 & 4\\
+%         200 & 5 & 5 & 7 & 9\\
+%         400 & 9 & 11 & 13 & 19\\
+%         800 & 19 & 23 & 31 & 46\\
+%         \hline
+%     \end{tabular}
+%     \caption{Time in minutes required to conduct 1000 permuations on 32 cores for all three tests on a simulated F2 intercross, according to number of individuals in the mapping population and number of markers in the genome.}
+% \end{table}
+
 \section*{Conclusion}
@@
 In the case of genetic factors, they can be mapped, as illustrated in \CortyReanalysisPaper.
-In the case of covariates, they can be accommodated, which can increase power and control false positive rate, as described in \CortyMethodsPaper.
+In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as described in \CortyMethodsPaper.
 
@@
 
-\begin{figure}[ht!]
+\begin{figure}[ht]
     \begin{subfigure}[b]{\linewidth}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[ht]
     \begin{subfigure}[t]{0.5\linewidth}

### 2018-05-05 revision/response.tex
edits

@@
 2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
-  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
-  (These files are now in two places, since they are also in the Zenodo repository of the reanalysis paper.)
+  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
 }
@@
 \Response{
-  We have added a section in the introduction that describes how projects in plant and animal breeding seeking to modulate environmental variance are related to the work presented here.
+  The editor suggested references that relate to the heritabiliyty of environmental variance.
+  This literature was foundational toward the development of the work presented in this manuscript, so we agree with the editor that it should be cited in the introduction.
+  We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping.
 }
@@
 \Response{
-  In order to most clearly illustrate the standard workflow associated with package \texttt{vqtl}, the authors have demonstrated its usage on a an unrealistically-small genome.
+  To most clearly illustrate the standard workflow associated with package \texttt{vqtl}, the authors have demonstrated its usage on a simulated cross with an unrealistically-small genome.
   The authors are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
-  This manuscript has been submitted as part of a three-paper companionship, where (call this one paper one), paper two contains two real-data reanalyses and paper three contains one real-data reanalysis.
-  Thus, the authors contend it is appropriate to maintain this paper having only limited mention of real data, those mentions in the ``Performance Benchmarks'' section.
+  This manuscript has been submitted as part of a three-paper companionship, where each of the other papers contains an analysis of real data using package \texttt{vqtl}.
+  Thus, the authors contend it is appropriate to maintain this paper with only limited reference to real data --- a real dataset is used in in the first performance benchmark, as described below.
 }
@@
   Thank you for this recommendation.
-  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package -- we have added two sets of benchmarks.
+  The authors agree that including this information will make the paper more useful to geneticists who are considering using the package --- we have added two sets of benchmarks.
 }
@@
   First, we demonstrate the performance of the package according to number of computer cores used and number of permutations conducted.
-  For this set of benchmarks, we used the intercross from Kumar et.
-al 2013 which we reanalyzed in \CortyReanalysisPaper, which involved 244 mice, 93 markers, and after using a hidden Markov model to estimate genotype probabilities at locations between markers, 582 loci.
-  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in a companion manuscript.
+  For this set of benchmarks, we used the intercross from Kumar et. al 2013 which we reanalyzed in \CortyReanalysisPaper.
+  This intercross population has 244 mice and was genotyped at 93 markers.
+  After applying the hidden Markov model in \texttt{R/qtl} with \texttt{step.size} of 2, the number of loci to be tested was 582.
+  As recommended by the editor, this example helps tie together the companions, because the results of this analysis are presented in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
   This addition can be found on page 4.
@@
   Thank you for this recommendation.
-  We have addressed this question in the discussion section, on page 4.
+  In short, the workflow presented here, which relies on permutation to assess statistical significance, would not be practical with a dense genotyping array with $\textgreater$ 10,000 markers, such as the MegaMUGA ($\approx$ 78,000 markers) GigaMuga ($\approx$ 150,000 markers), or the Mouse Diversity Genotyping array ($\approx$ 623,000 markers).
+  We have addressed this question in the ``Performance Benchmarks'' section, starting on page 4 and continuing onto page 5.
 }

### 2018-05-05 Package_vqt_Manuscript.tex
edits

@@
 
-Traditional analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
+It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Lynch1998}.
+This phenomenon has been described theoretically \citep{Hill2010,Hill2004-uo} and demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+
+However, traditional QTL mappinng analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
 Recent work, however, has challenged the assumption of homogeneous variance and sought to identify an additional type of QTL --- those that
@@
 
-\begin{figure}[t]
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
-    \end{subfigure}
-    \caption{
-        For each of the four simulated phenotypes, the genome scan shows the LOD score of each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
-        The traditional test is in green and globally similar to the mQTL test.
-    }
-    \label{fig:lod_score_scans}
-\end{figure}
-
-\begin{figure}[t]
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{\linewidth}
-        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
-    \end{subfigure}
-
-    \caption{
-        For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value for each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
-        The traditional test is in green and globally similar to the mQTL test.
-        A value of 2 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per hundred phenotypes no QTL.
-    }
-    \label{fig:empir_p_scans}
-\end{figure}
-
 The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
@@
 \begin{figure}[t]
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype1.pdf}
     \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype2.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype3.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/LOD_scan_phenotype4.pdf}
+    \end{subfigure}
+    \caption{
+        For each of the four simulated phenotypes, the genome scan shows the LOD score of each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
+        The traditional test is in green and globally similar to the mQTL test.
+    }
+    \label{fig:lod_score_scans}
+\end{figure}
+
+Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.
+The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
+Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
+The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
+% With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
+
+The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
+\RC{
+    They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
+}{
+    Could change this to read ``They also rely on the alternative model being exactly true; severe model misspecification is severe this can lead to inflated statistics.''
+    Would be more accurate, but I don't know the rules with regards to things the reviewer didn't mention.
+}
+
+More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
+
+\subsection{Robust, genomewide-adjusted p-values}
+
+\begin{figure}[t]
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype1.pdf}
     \end{subfigure}
 
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype2.pdf}
     \end{subfigure}
-    \hfill
-    \begin{subfigure}[t]{0.5\linewidth}
-        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
+
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype3.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{\linewidth}
+        \includegraphics[width=\textwidth]{images/empir_p_scan_phenotype4.pdf}
     \end{subfigure}
@@
     \caption{
-        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus with mean effects on the horizontal axis and variance effects on the vertical axis.
-        Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
-        For \texttt{phenotype1}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
-        For \texttt{phenotype2}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL.
-        For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
-        For \texttt{phenotype4}, the pattern of two-dimensional separation illustrates an mvQTL.
-        \label{fig:mean_var_plots}
+        For each of the four simulated phenotypes, the genome scan shows the -log10 of the FWER-corrected $p$-value for each test --- mQTL, vQTL, and mvQTL --- in blue, red, and black, respectively.
+        The traditional test is in green and globally similar to the mQTL test.
+        A value of 2 implies that the quantity of evidence against the null is such that we expect to see this much or more evidence once per hundred phenotypes no QTL.
     }
+    \label{fig:empir_p_scans}
 \end{figure}
-
-Each type of test (mQTL, vQTL, and mvQTL) yields two association statistics: the LOD score, and the (nominal) $p$-value.The LOD is a raw measure of association equal to the base 10 logarithm of the likelihood ratio (LR) between the fitted alternative and null models.
-Higher values indicate greater association when considered across loci for the same type of test; but LOD scores between different types of tests, namely between mvQTL test vs either mQTL or vQTL tests, are not readily comparable.
-The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
-With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
-
-The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
-They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
-More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
-
-\subsection{Robust, genomewide-adjusted p-values}
 
@@
 \begin{figure}[t]
-    \includegraphics[width = \linewidth]{images/benchmark_kumar.pdf}
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen1.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen2.pdf}
+    \end{subfigure}
+
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen3.pdf}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[t]{0.5\linewidth}
+        \includegraphics[width=\textwidth]{images/mean_var_plot_phen4.pdf}
+    \end{subfigure}
+
     \caption{
-        Time taken to run \texttt{scanonevar.perm} on the data from \citet{Kumar2013} which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
-        For a given number of cores, there is a linear relationship between number of permutations conducted and time required.
-        The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
+        \texttt{mean\_var\_plot}s show the estimated genotype effects at a locus with mean effects on the horizontal axis and variance effects on the vertical axis.
+        Horizontal lines indicate standard errors for mean effects and vertical lines indicate standard errors for variance effects.
+        For \texttt{phenotype1}, the pattern of overlapping estimates and standard errors is consistent with the fact that there are no genetic effects, and the $p$-value was not statistically significant at any locus.
+        For \texttt{phenotype2}, the pattern of horizontal, but not vertical, separation visually illustrates the identified mQTL.
+        For \texttt{phenotype3}, the pattern of vertical, but not horizontal, separation visually illustrates the identified vQTL.
+        For \texttt{phenotype4}, the pattern of two-dimensional separation illustrates an mvQTL.
+        \label{fig:mean_var_plots}
     }
-    \label{fig:benchmark_kumar}
 \end{figure}
@@
 
-\begin{figure}[t]
-    \includegraphics[width = \linewidth]{images/benchmark_sim_cross.pdf}
-    \caption{
-        Time taken to run 1000 permutation scans on 32 cores on simulated data using \texttt{scanonevar.perm}, varying the number of individuals in the mapping population and the number of markers in the genome.
-        For a given population size, there is a slightly supra-linear relationship between number of markers and time required.
-        The average slope of the line indicates the average time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
-    }
-    \label{fig:benchmark_sim_cross}
-\end{figure}
-
 Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
@@
 
+\begin{figure}[t]
+    \includegraphics[width = \linewidth]{images/benchmark_kumar.pdf}
+    \caption{
+        Time taken to run \texttt{scanonevar.perm} on the data from \citet{Kumar2013} which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
+        For a given number of cores, there is a linear relationship between number of permutations conducted and time required.
+        The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
+    }
+    \label{fig:benchmark_kumar}
+\end{figure}
+
+\begin{figure}[t]
+    \includegraphics[width = \linewidth]{images/benchmark_sim_cross.pdf}
+    \caption{
+        Time taken to run 1000 permutation scans on 32 cores on simulated data using \texttt{scanonevar.perm}, varying the number of individuals in the mapping population and the number of markers in the genome.
+        For a given population size, there is a slightly supra-linear relationship between number of markers and time required.
+        The average slope of the line indicates the average time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
+    }
+    \label{fig:benchmark_sim_cross}
+\end{figure}
+
 By far, the most computationally-intensive step in the mean-variance QTL mapping process is the assessment of genome-wide statistical significance by permutation.
@@
 The slope of the line indicates the time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
+
+Based on these benchmarks, it is clear that the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
+Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci are relevant, could not be practically analyzed with package \texttt{vqtl} in this way.
+However, both statistical and computational steps could be taken to make such a study feasible.
+Statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004}.
+Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
 

### 2018-05-06 Package_vqt_Manuscript.tex
edits

@@
 
-It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Lynch1998}.
-This phenomenon has been described theoretically \citep{Hill2010,Hill2004-uo} and demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
+This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
 
@@
 Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
+
+In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
+In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
+The dataset from one of these intercrosses is used to benchmark the performance of package \texttt{vqtl}, as is described in the ``Performance Benchmarks'' section.
+In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).
 
@@
 This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
-In the case of genetic factors, they can be mapped, as illustrated in \CortyReanalysisPaper.
-In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as described in \CortyMethodsPaper.
+In the case of genetic factors, they can be mapped, as illustrated in one companion article (\CortyReanalysisPaper).
+In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as illustrated in another companion article (\CortyMethodsPaper).
 

### 2018-06-07 revision/response.tex
minor
edited first couple of R1 comments

@@
 \usepackage[letterpaper, headheight=100pt]{geometry}
-\geometry{verbose,tmargin=3cm,bmargin=3cm,lmargin=3cm,rmargin=2.5cm}
+\geometry{verbose,tmargin=3cm,bmargin=2cm,lmargin=2.5cm,rmargin=2.5cm}
 \usepackage{amsmath}
@@
 \Response{
-  The editor suggested references that relate to the heritabiliyty of environmental variance.
-  This literature was foundational toward the development of the work presented in this manuscript, so we agree with the editor that it should be cited in the introduction.
+  Thank you for these suggestions. 
+  This literature was foundational toward the development of the work presented in this manuscript, and we agree with the editor that it should be cited in the introduction.
   We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping.
@@
 \Response{
-  To most clearly illustrate the standard workflow associated with package \texttt{vqtl}, the authors have demonstrated its usage on a simulated cross with an unrealistically-small genome.
-  The authors are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
-  This manuscript has been submitted as part of a three-paper companionship, where each of the other papers contains an analysis of real data using package \texttt{vqtl}.
-  Thus, the authors contend it is appropriate to maintain this paper with only limited reference to real data --- a real dataset is used in in the first performance benchmark, as described below.
+    Our use of an unrealistically-small genome is to most clearly illustrate the standard workflow.
+%    But we are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
+    Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of three full sets of real data using the R package, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
+Nonetheless, we note that we have added use of a full, real dataset is used in the newly added performance benchmark, as described below.
 }

### 2018-06-07 Package_vqt_Manuscript.tex
edited first couple of R1 comments

@@
 Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci are relevant, could not be practically analyzed with package \texttt{vqtl} in this way.
-However, both statistical and computational steps could be taken to make such a study feasible.
+\WV{However, both statistical and computational steps could be taken to make such a study feasible.}{Are there ways to make efficient the analysis of multiple traits on the same population?
+For example, in the standard analysis of gene expression data, rint-ing a probe then performing permutations on it provides thresholds that will be valid for all other rint-ed probes (assuming no missingness; eg, ask Greg).
+Not sure if this applies in our situation...
+needs more thought.}
 Statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004}.
-Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
+Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.\WV{}{Does this section extend the paper beyond the required limits.}
 

### 2018-06-07 Package_vqt_Manuscript.tex
edits

--- 2018-05-06 Package_vqt_Manuscript.tex
@@
 Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci are relevant, could not be practically analyzed with package \texttt{vqtl} in this way.
+\RC
+{
+    \WV
+    {
         However, both statistical and computational steps could be taken to make such a study feasible.
+    }{
+        Are there ways to make efficient the analysis of multiple traits on the same population?
+For example, in the standard analysis of gene expression data, rint-ing a probe then performing permutations on it provides thresholds that will be valid for all other rint-ed probes (assuming no missingness; eg, ask Greg).
+Not sure if this applies in our situation...
+needs more thought.
+    }
+}
+{
+    I think simply suggesting that larger-scale analysis might be possible is appropriate here, since we haven't done it and haven't even worked out how to do it.
+    I don't think it makes sense for us to try to figure out how we would have done a different study and report exactly how we would have done that.
+}
 Statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004}.
 Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
+\RC
+{
+    \WV{}
+    {Does this section extend the paper beyond the required limits.}
+}
+{
+    Yes.  Do you think they will notice/care?
+}
 

### 2018-07-10 revision/response.tex
edits

@@
 
-\lhead{Manuscript ID\\G3/2018/200159}
-\chead{\bfseries Response to Reviewer Comments}
+\lhead{Manuscript ID: G3/2018/200159 \\ Title: ``\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping''}
+\chead{\bfseries Response to Reviewer Comments \\ \mbox{}\\ }
 \rhead{\today\\William Valdar}
@@
   See detailed response to Reviewer 1 comment 3. 
+  Note that this has lengthened the paper somewhat, but we can move plots to supplemental if that would be preferred.
 }
@@
     Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of three full sets of real data using the R package, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
-Nonetheless, we note that we have added use of a full, real dataset is used in the newly added performance benchmark, as described below.
+    Nonetheless, we note that a full, real dataset is used in the newly added performance benchmark, as described below.
 }
@@
 \Response{
-  First, we demonstrate the performance of the package according to number of computer cores used and number of permutations conducted.
-  For this set of benchmarks, we used the intercross from Kumar et. al 2013 which we reanalyzed in \CortyReanalysisPaper.
+  First, we demonstrate the performance of the package according to the number of computer cores used and the number of permutations conducted.
+  For this set of benchmarks, we used the intercross from Kumar et. al 2013, which we reanalyzed in \CortyReanalysisPaper.
   This intercross population has 244 mice and was genotyped at 93 markers.

### 2018-07-10 Package_vqt_Manuscript.tex
edits

--- 2018-05-06 Package_vqt_Manuscript.tex
@@
 
-\author[$\ast$,1]{Robert W. Corty}
-\author[$\ast, 1$]{William Valdar}
+\author[$\ast,\dagger$]{Robert W. Corty}
+\author[$\ast,\ddagger,1$]{William Valdar}
 
-\affil[$\ast$]{University of North Carolina at Chapel Hill, Department of Genetics}
+\affil[$\ast$]{Department of Genetics}
+\affil[$\dagger$]{Bioinformatics and Computational Biology Curriculum}
+\affil[$\ddagger$]{and Lineberger Comprehensive Cancer Center, University of North Carolina,
+    Chapel Hill, NC}
 
-\keywords{QTL mapping, vQTL, mvQTL, variance heterogeneity, DGLM, R}
+\keywords{heteroskedastic regression, mvQTL, variance heterogeneity, DGLM}
 
@@
 
-QTL mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
-Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.
+Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
+Recent work, however, has challenged the assumption of homogeneous residual variance and sought to identify an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
 
-It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
-This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
+We have developed a standardized method for QTL mapping in experimental crosses that simultanously model mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
+Our approach, which we term ``mean-variance QTL mapping'', can currently be applied to F2 intercross and backcross designs.
+In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
 
-However, traditional QTL mappinng analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for each every individual in the mapping population.
-Recent work, however, has challenged the assumption of homogeneous variance and sought to identify an additional type of QTL --- those that
-%, by driving heterogeneity of epistatic or environmental variance, 
-influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
-
-Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, the statistical methods used for this purpose remain heterogeneous.
-% and sometimes idiosyncratic, with connections to methods for mapping mean QTL often lacking or unclear.
-% a clear or intuitive connection to methods used for detecting mean QTL.
-%there remains great heterogeneity among the statistical methods used to conduct these analyses.
-We developed a standardized method for QTL mapping that can detect mQTL, vQTL, and a generalization of the two that we term ``mvQTL''.
-This approach, which we term ``mean-variance QTL mapping'', can be applied to intercross and backcross designs.
-In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper)
-
-Here, we provide a practical guide to using the \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping.
+Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping and is inter-operable with the established \texttt{qtl} package \citep{Broman2003}.
 First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
 On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
-The association statistics of all four tests are then initially plotted in LOD score units, with drawbacks of this plotting unit discussed.
-Permutation scans are used to determine empirically adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more comparable.
-Last, we describe plots to communicate the effects that led to the detection of a QTL, and use the bootstrap to estimate its confidence interval.
+The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
+Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
+Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.
 
-In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
-In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
-The dataset from one of these intercrosses is used to benchmark the performance of package \texttt{vqtl}, as is described in the ``Performance Benchmarks'' section.
-In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).
+%In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
+%~In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
+%In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).
 
-% Said another way, it is typically assumed that nothing -- neither environmental factors nor genetic factors -- influences the residual variance of the trait.
+% \section*{Old intro material -- TO DELETE OR REPURPOSE}
 
-% In fact, gene-gene interactions, gene-environment interactions, genetic factors that influence microenvironmental sensitivity, and environmental factors can influence residual variance.
+% Quantitative Trait Locus (QTL) mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
+% Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.
 
-% This approach allows for QTL mapping in the presence of both genetic and non-genetic effects on phenotype mean and variance.
-% We compare its behavior to other QTL mapping approaches, and illustrate the discovery of a new QTL from an existing study.
-
-% overview of the software or resource
-% In support of other researchers who may be interested in mvQTL mapping, we created \texttt{R} package \texttt{vqtl}, which provides functions for conducting genome scans, assessing the statistical significance of results, and visualizing and interpreting significant findings.
-
-% describe how researchers can access it.
-% \texttt{R} package \texttt{vqtl} uses the same \texttt{cross} data structure as the popular \texttt{qtl} package and is available on \texttt{CRAN}, so it is easy to get started.
-% Here, we demonstrate typical use of the \texttt{vqtl} package to analyze four simulated traits and discover three QTL.
-% a QTL that influence phenotype mean (an mQTL), a QTL that influences phenotype variance (vQTL) and a QTL that influences a combination of mean and variance (an mvQTL).
-% The code used to simulate the phenotypes, calculate the statistics, estimate their significance, and visualize significant results is available at \texttt{github.com/rcorty}.
+% It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
+% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
 
@@
 
-To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using the popular \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
+To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
 This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model.
@@
 The $p$-values described above, however, are nominal: they do not take into account multiple testing across the genome.
-\RC{
     They also rely on asymptotic theory that assumes the underlying phenotype being residually normal; this may not always be the case and when violated will lead to inflated significance.
-}{
-    Could change this to read ``They also rely on the alternative model being exactly true; severe model misspecification is severe this can lead to inflated statistics.''
-    Would be more accurate, but I don't know the rules with regards to things the reviewer didn't mention.
-}
-
 More robust $p$-values that are corrected for genomewide significance via control of the family-wise error rate (FWER) can be obtained empirically, through a permutation procedure described below.
@@
 
-Last, it is important to assess the genetic precision of a discovered QTL for bioinformatic follow-up.
-The function \texttt{scanonevar.boot} implements the non-parametric bootstrap \citep{Visscher1996}.
+Last, to assess the genetic precision of a discovered QTL for bioinformatic follow-up, the function \texttt{scanonevar.boot} estimates confidence intervals via the non-parametric bootstrap \citep{Visscher1996}.
 This function takes, as arguments, a \texttt{scanonevar} object, the type of QTL detected, the name of the chromosome containing the QTL, and \texttt{num.resamples}, the number of bootstrap resamplings desired.
@@
 For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
-For a given number of cores, there is a linear relationship between number of permutations conducted and time required \autoref{fig:benchmark_kumar}.
-The slope the the line indicates time required per permutation and is dependent on the number of cores, ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
+For a given number of cores, the relationship between time and the number of permutations is linear (\autoref{fig:benchmark_kumar}), the slope depending on the number of cores and ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
 
 For the second benchmark, we ran \texttt{scanonevar.perm} on simulated data, always conducting 1000 permutations and using 32 cores, but varying the number of individuals in the mapping population and the number of markers in the genome.
-For a given population size, there is a slightly supra-linear relationship between number of markers and time required \autoref{fig:benchmark_sim_cross}, which reflects a linear increase in the time taken to conduct the permuted genome scans, plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
-The slope of the line indicates the time required per locus and is dependent on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
+For a given population size, there is a slightly curvilinear relationship between number of markers and time required (\autoref{fig:benchmark_sim_cross}), which reflects a linear increase in the time taken to conduct the permuted genome scans plus an increase in the time taken for ``bookkeeping'' tasks like organizing and reshaping genetic data.
+The slope (minutes per locus) depends on the population size, ranging from $\approx$ 1.4 seconds per locus with a population of size 100 to $\approx$ 3.3 seconds per locus with a population of size 800.
 
-Based on these benchmarks, it is clear that the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
-Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci are relevant, could not be practically analyzed with package \texttt{vqtl} in this way.
-However, both statistical and computational steps could be taken to make such a study feasible.
-Statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004}.
-Computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
+Based on these benchmarks, the workflow presented here is practical for QTL mapping F2 intercross and similar populations on modern, multi-core scientific computers.
+Populations with many recombinations, where dense genotyping arrays that interrogate $\textgreater$ 10,000 loci, could not be practically analyzed with package \texttt{vqtl} in this way, although it is likely that statistical and computational steps could be taken to make such studies feasible:
+statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
+computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
 
@@
 \section{Acknowledgments}
-This work was primarily funded by a National Institutes of General Medical Sciences (NIGMS) grant to WV (R01-GM104125) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
+This work was primarily funded by National Institutes of General Medical Sciences (NIGMS) grants to WV (R01-GM104125,R35-GM127000) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
 RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.

### 2018-07-11 Package_vqt_Manuscript.tex
edits

@@
 Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-Recent work, however, has challenged the assumption of homogeneous residual variance and sought to identify an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
+However, it has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
+This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+Recent work, has begun to recognize heterogeneous residual variance in mapping efforts by identifying an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
 

### 2018-07-12 revision/response.tex
edits

@@
   This literature was foundational toward the development of the work presented in this manuscript, and we agree with the editor that it should be cited in the introduction.
-  We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping.
+  We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping (``It has long been recognized\dots'').
 }
@@
 \Response{
-  The reviewer is quite correct and we have edited the introduction to reflect the fact that QTL mapping has been applied productively in plant and animal research.
+  The reviewer is quite correct.
+Rather than add to this sentence, we have removed the explicit reference to health and disease, recognizing that the introduction started too broadly.
+Instead we have included references to heritability of variance that rely heavily on studies of model organisms, livestock and crops.
 }

### 2018-07-12 Package_vqt_Manuscript.tex
edits

--- 2018-07-10 Package_vqt_Manuscript.tex
@@
 
-\keywords{heteroskedastic regression, mvQTL, variance heterogeneity, DGLM}
+\keywords{heteroscedastic regression, mvQTL, variance heterogeneity, DGLM}
 
@@
 
-Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-Recent work, however, has challenged the assumption of homogeneous residual variance and sought to identify an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
+Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
+It has long been recognized, however, that the residual variance is itself heritable \WV{\citep{Falconer1965,Lynch1998}}{Do they definitely talk about this?}.
+This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \WV{\citep{Sorensen2015}}{Seems a bit recent} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+%To help dissect the genetic basis, 
+Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}\WV{\citep{Soave2017,Dumitrascu2018}}{add Englehardt, Soave?}.
 
-Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
-We have developed a standardized method for QTL mapping in experimental crosses that simultanously model mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
-Our approach, which we term ``mean-variance QTL mapping'', can currently be applied to F2 intercross and backcross designs.
-In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
+% Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
+% However, it has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
+% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+% Recent work, has begun to recognize heterogeneous residual variance in mapping efforts by identifying an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
 
-Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping and is inter-operable with the established \texttt{qtl} package \citep{Broman2003}.
+Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
+We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
+Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
+In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL (\CortyMethodsPaper).
+In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL (\CortyReanalysisPaper).
+
+Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
+%
+%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
+%
+%The R package \texttt{vqtl} implements mean-variance QTL mapping
+which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
+%We introduce the \texttt{vqtl} package as follows.
 First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
@@
 \section{Acknowledgments}
-This work was primarily funded by National Institutes of General Medical Sciences (NIGMS) grants to WV (R01-GM104125,R35-GM127000) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
-RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.
+This work was funded by National Institutes of General Medical Sciences grants R01-GM104125 (RWC,WV), R35-GM127000 (RWC,WV), T32-GM067553 (RWC); National Heart, Lung and Blood Institute grant R21 HL126045 (RWC,WV); National Library of Medicine grant T32-LM012420 (RWC); and a National Institute of Mental Health grant F30-MH108265 (RWC).
 

### 2018-07-12 revision/response.tex
edits

@@
 \Response{
-  The reviewer is quite correct.
-Rather than add to this sentence, we have removed the explicit reference to health and disease, recognizing that the introduction started too broadly.
-Instead we have included references to heritability of variance that rely heavily on studies of model organisms, livestock and crops.
+  The reviewer is quite correct that QTL mapping has been productively applied in plant an animal populations in addition to human populations.
+  Rather than add references from plant and animal genetics to the introduction, we recognized that the introduction started unnecessarily broadly and therefore removed the first sentence.
+  To address the reviewer's concern, however, we have added references to the following paragraph (where the heritability of phenotype variance is discussed) that rely heavily on studies of model organisms, livestock and crops.
 }

### 2018-07-12 Package_vqt_Manuscript.tex
edits

@@
 
-Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-It has long been recognized, however, that the residual variance is itself heritable \WV{\citep{Falconer1965,Lynch1998}}{Do they definitely talk about this?}.
-This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \WV{\citep{Sorensen2015}}{Seems a bit recent} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
+It has long been recognized, however, that the residual variance is itself heritable \citep{Falconer1965,Lynch1998}.
+This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
 %To help dissect the genetic basis, 
-Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}\WV{\citep{Soave2017,Dumitrascu2018}}{add Englehardt, Soave?}.
+Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014,Soave2017,Dumitrascu2018}.
 

### 2018-07-24 Package_vqt_Manuscript.tex
edits

@@
 Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-It has long been recognized, however, that the residual variance is itself heritable \citep{Falconer1965,Lynch1998}.
-This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
+It has long been recognized, however, that the residual variance is itself heritable \citep{Falconer1965,Lynch1998}, a
+phenomenon that has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
 %To help dissect the genetic basis, 
 Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014,Soave2017,Dumitrascu2018}.
+However, although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
 
@@
 
-Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
 We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
@@
 Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
+\WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
 %

### 2018-08-04 revision/response.tex
finished round 1 of revisions
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
   This literature was foundational toward the development of the work presented in this manuscript, and we agree with the editor that it should be cited in the introduction.
-  We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping (``It has long been recognized\dots'').
+  We have improved the introduction to describe how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping and used many of the suggested citations in this section.
 }
@@
   Rather than add references from plant and animal genetics to the introduction, we recognized that the introduction started unnecessarily broadly and therefore removed the first sentence.
-  To address the reviewer's concern, however, we have added references to the following paragraph (where the heritability of phenotype variance is discussed) that rely heavily on studies of model organisms, livestock and crops.
+  To address the reviewer's concern, however, we have added references to the next paragraph in the introduction (where the heritability of phenotype variance is discussed) that rely heavily on studies of model organisms, livestock and crops.
 }
@@
 \Response{
-    Our use of an unrealistically-small genome is to most clearly illustrate the standard workflow.
-%    But we are encouraged by the reviewer's enthusiasm to see the package applied to real datasets.
-    Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of three full sets of real data using the R package, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
-    Nonetheless, we note that a full, real dataset is used in the newly added performance benchmark, as described below.
+  The unrealistically-small genome we used to illustrate the standard workflow serves an important purpose --- to allow a reader to replicate the procedure and its ``findings'' on a typical desktop in a matter of minutes.
+  Since this manuscript has been submitted as part of a three-paper companionship where each of the other papers contains analyses of real data using package \texttt{vqtl}, we feel that using a reduced dataset to demonstrate the package functions is reasonable in this instance.
+  Additionally, in response to other reviewer comments, we have added performance benchmarks to this manuscript, which make use of the dataset from Kumar et al 2013, as described below.
 }

### 2018-08-04 Package_vqt_Manuscript.tex
finished round 1 of revisions
Signed-off-by: Robert Corty <rcorty@gmail.com>

--- 2018-07-12 Package_vqt_Manuscript.tex
@@
 Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
+% \WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
 %
@@
 
-To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using \texttt{R/qtl} package \citep{Broman2003}, on which \texttt{vqtl} is based.
-This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{R/qtl}'s hidden Markov model.
+To illustrate the use of the \texttt{vqtl} package, we first simulated an example F2 intercross using the \texttt{R} package {qtl} \citep{Broman2003}, on which \texttt{vqtl} is based.
+This cross consisted of 200 male and 200 female F2 offspring, with 3 chromosomes of length 100 cM, each tagged by 11 equally-spaced markers and estimated genotype probabilities at 2cM intervals with \texttt{qtl}'s hidden Markov model.
 We then generated four phenotypes:
@@
 
-The central function for genetic mapping in package \texttt{R/qtl} is \texttt{scanone} \citep{Broman2003}.
+The central function for genetic mapping in package \texttt{qtl} is \texttt{scanone} \citep{Broman2003}.
 Analogously, the central function for mean-variance QTL mapping in package \texttt{vqtl} is \texttt{scanonevar}, building on an early version of \texttt{scanonevar} in package \texttt{qtl}.
@@
 statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
-computationally, the software could be changed to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
+computationally, the software could be modified to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
 

### 2018-08-04 Package_vqt_Manuscript.tex
update DOI for resubmission
Signed-off-by: Robert Corty <rcorty@gmail.com>

@@
 \section*{Resources}
-The scripts used to simulate genotypes and phenotypes, conduct the genome scans, and plot the results are available as a public, static Zenodo repository at \url{DOI: 10.5281/zenodo.1173799}.
+The scripts used to simulate genotypes and phenotypes, conduct the genome scans, and plot the results are available as a public, static Zenodo repository at \url{DOI: 10.5281/zenodo.1336302}.
 The package \texttt{vqtl} and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.

### 2018-10-10 revision/response.tex
Updated formatting of co-submitted papers and response to match that of BVH

@@
 \newcommand{\reply}{\ensuremath{\hookrightarrow}\xspace}
-
-\newcommand{\CortyRPaper}{\textcolor{purple}{Corty and Valdar 2018+ [r/vqtl]}\xspace}
-\newcommand{\CortyReanalysisPaper}{\textcolor{purple}{Corty \emph{et al} 2018+}\xspace}
-\newcommand{\CortyMethodsPaper}{\textcolor{purple}{Corty and Valdar 2018+ [BVH]}\xspace}
 
@@
 
-\newcommand{\ReviewerQuestion}[1]{
+\newcommand{\ReviewerQuestion}[2]{
   \vspace{5pt}\goodbreak
-  \noindent\fbox{Reviewer:} #1
+  \noindent\fbox{Reviewer:~#1} #2
   \normalfont\par
@@
   \goodbreak
-  \textcolor{blue}{#1}
+  \textcolor{blue!50!black}{#1}
   \normalfont\par
@@
 
+\newcommand{\Line}[1]{\textbf{line #1}\xspace} % to highlight for our use, in case line numbers change significantly and we have to correct.
+\newcommand{\Lines}[1]{\textbf{line #1}\xspace} %
+\newcommand{\CortyRPaper}{Corty and Valdar (2018+) ``vqtl: An R package for mean-variance QTL mapping'' G3/2018/200159\xspace}
+\newcommand{\CortyReanalysisPaper}{Corty \emph{et. al.} (2018+) ``Mean-variance QTL mapping identifies novel QTL for circadian activity and exploratory behavior in mice'' G3/2018/200194\xspace}
+\newcommand{\CortyMethodsPaper}{Corty and Valdar (2018+) ``QTL mapping on a background of variance heterogeneity'' G3/2018/200195\xspace}
+
 \begin{document}
 \reversemarginpar
+
+\section*{Introduction}
+
+This response refers to paper number \textbf{3} of the 3 co-submitted papers:
+\begin{enumerate}
+\item \CortyMethodsPaper
+\item \CortyReanalysisPaper
+\item \textbf{\CortyRPaper}
+\end{enumerate}
 
@@
   We have updated the manuscripts such that each one refers to the other two in the introduction and discussion.
+We have also edited the introductions so that they are less duplicative and each is more clearly focused on its own distinct contributions.
 }
@@
 \Response{
-  We have incorporated this recommendation, using the dataset from Kumar et al.
-2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
+  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the co-submitted manuscript, \CortyReanalysisPaper.
   The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
@@
 
-\ReviewerQuestion{
+\ReviewerQuestion{1.1}{
   The authors mention ``in almost all traits of interest in human health and disease''.
@@
 
-\ReviewerQuestion{
+\ReviewerQuestion{1.2}{
   The authors used a set of simulated data with 3 chromosomes and 11 equally spaced markers.
@@
 
-\ReviewerQuestion{
+\ReviewerQuestion{1.3}{
   The CPU time is indicated for the data set of the example.
@@
 
-\ReviewerQuestion{
+\ReviewerQuestion{1.4}{
   In the simulation, the authors should indicate the magnitude of the genetic variance explained by the QTL.
@@
 
-\ReviewerQuestion{
+\ReviewerQuestion{1.5}{
   I think the authors should also indicate if the software is capable in terms of computing power to analyze the data from SNP genotyping devices or even sequencing.

### 2018-10-10 Package_vqt_Manuscript.tex
Updated formatting of co-submitted papers and response to match that of BVH

@@
 Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
-In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL (\CortyMethodsPaper).
-In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL (\CortyReanalysisPaper).
+In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL \citep{CortyValdar2018bvh-biorxiv}.
+In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL \citep{Corty2018biorxiv}.
 
@@
 %
-%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
+%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\citep{CortyValdar2018bvh-biorxiv}),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
 %
@@
 Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
-Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.
+Last, we benchmark performance, using one of the datasets examined in \citet{Corty2018biorxiv} to report how computation time varies with marker density and number of permutations.
 
@@
 
-To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure [\CortyMethodsPaper].
+To calculate the empirical, FWER-controlled $p$-value of each test at each locus we advocate use of a permutation procedure \citep{CortyValdar2018bvh-biorxiv}.
 Like previous work on permutation-based thresholds for genetic mapping \citep{Churchill1994,Carlborg2002}, this procedure sidesteps the need to explicitly estimate the effective number of tests.
@@
 
-For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \CortyReanalysisPaper, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
+For the first benchmark, we ran \texttt{scanonevar.perm} on the data from \citet{Kumar2013} and \citet{Corty2018biorxiv}, which contains 244 individuals and 582 loci, varying the number of permutations desired and the number of computer cores used.
 For a given number of cores, the relationship between time and the number of permutations is linear (\autoref{fig:benchmark_kumar}), the slope depending on the number of cores and ranging from $\approx$ 6.3 seconds per permutation with 4 cores to $\approx$ 1.2 second per permutation with 32 cores.
@@
 This package is appropriate for crosses and phenotypes where genetic factors or covariates or are known or suspected to influence phenotype variance.
-In the case of genetic factors, they can be mapped, as illustrated in one companion article (\CortyReanalysisPaper).
-In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as illustrated in another companion article (\CortyMethodsPaper).
+In the case of genetic factors, they can be mapped, as illustrated in one companion article \citep{Corty2018biorxiv}.
+In the case of covariates, they can be accommodated, which can increase power and improve false positive rate control, as illustrated in another companion article (\citep{CortyValdar2018bvh-biorxiv}).
 
@@
 
-\bibliography{10_Aim1,wv}
+\bibliography{vqtl}
 

### 2018-10-16 Package_vqt_Manuscript.tex
minor fix
for re-upload to bioarxiv

@@
 
-\runningtitle{Package vqtl} % For use in the footer
+\runningtitle{R Package vqtl} % For use in the footer
 
@@
 
-\setboolean{displaycopyright}{true}
+\setboolean{displaycopyright}{false}
 
@@
 \thispagestyle{firststyle}
-\logomark
+% \logomark
 \articletypemark
@@
 %To help dissect the genetic basis, 
-Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014,Soave2017,Dumitrascu2018}.
+Correspondingly, several groups have proposed statistical methods for mapping QTL controlling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014,Soave2017,Dumitrascu2018}.
 However, although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
@@
 
-We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
+We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultaneously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
 Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
@@
 
-\bibliography{vqtl}
+\bibliography{10_Aim1,wv}
 

### 2018-10-30 Package_vqt_Manuscript.tex
delete comments

@@
 
-\articletype{inv} % article type
+\articletype{inv}
 
@@
 
-\runningtitle{R Package vqtl} % For use in the footer
+\runningtitle{R Package vqtl}
 
@@
 Because it is inter-operable with the popular \texttt{qtl} package and uses many of the same data structures and input patterns, it will be straightforward for geneticists to analyze future experiments with \texttt{vqtl} as well as re-analyze past experiments, possibly discovering new QTL.
-% % Mean-variance QTL mapping, a relatively novel approach to QTL mapping, elevates phenotype variance to the same level of consideration as phenotype mean.
-% Most existing methods for QTL mapping in experimental crosses assume that the residual variance is constant across all individuals.
-% But common situations violate this assumption.
-% For many phenotypes, one sex is more variable than the other, some experimenters make more precise measurements than others, and specific genetic factors influence environmental sensitivity.
-% In these cases, mean-variance QTL mapping provides higher power and better protection against false positives.
-% It also allows for detection of QTL that influence phenotype variance, termed vQTL and QTL that influence some mixture of phenotype mean and variance, termed mvQTL.
-% We present \texttt{R} package \texttt{vqtl}.
-% This package makes it easy for geneticists to apply the mean-variance QTL mapping approach, control family-wide error rate (FWER), and visualize and interpret their results.
 \end{abstract}
@@
 \setboolean{displaycopyright}{false}
-
-% \linespread{2}
 
@@
 \correspondingauthoraffiliation{Correspondence e-mail: william.valdar@unc.edu}
-\vspace{-24pt}%
+\vspace{-24pt}
 
@@
 
-% Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-% However, it has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
-% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
-% Recent work, has begun to recognize heterogeneous residual variance in mapping efforts by identifying an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.
-
 We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultaneously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
@@
 Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
-% \WV{}{Should we be using ``R/vqtl'' and ``R/qtl'' terminology?}
-%
-%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\citep{CortyValdar2018bvh-biorxiv}),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
-%
-%The R package \texttt{vqtl} implements mean-variance QTL mapping
 which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
-%We introduce the \texttt{vqtl} package as follows.
 First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
@@
 Last, we benchmark performance, using one of the datasets examined in \citet{Corty2018biorxiv} to report how computation time varies with marker density and number of permutations.
-
-%In two companion articles, we apply mean-variance QTL mapping with package \texttt{vqtl} to discover new QTL.
-%~In the first, we describe a novel vQTL for exploratory behavior and a novel mQTL for a circadian behavior trait, both in mouse F2 intercrosses retrieved from the Mouse Phenome Database (\CortyReanalysisPaper) \citep{Bogue2015}.
-%In the second, we describe a unique strength of mean-variance QTL mapping, its ability to accommodate variance heterogeneity arising from background factors, such as sex, batch, and housing, and demonstrate the value of such accommodation with both simulation studies and a report of a novel QTL (\CortyMethodsPaper).
-
-% \section*{Old intro material -- TO DELETE OR REPURPOSE}
-
-% Quantitative Trait Locus (QTL) mapping studies in experimental crosses have provided important insights on nearly every trait of interest in human health, as well as livestock and crop improvement programs.
-% Advances in model organism genotyping \citep{Williams1990} and phenotyping \citep{Yang2014a} as well as in statistical methods \citep{Lander1989a,Martinez1992} and software tools (\eg, \citealt{Broman2003,Mulligan2017}) have supported these discoveries.
-
-% It has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
-% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
 
@@
 \end{enumerate}
-% Optional argument \texttt{chrs} is used to specify a subset of chromosomes to be scanned, defaulting to all chromosomes.
-% Optional argument \texttt{return.covar.effects} is used to specify whether or not fitted effects of all covariates should be returned as part of the scan result, defaulting to \texttt{FALSE}.
-% This option can be useful in assessing the trends in the various genetic components of the phenotype across the genome or in assessing the trends in covariate effects across the genome.
-% \RC{}{Doesn't fit.  It's already in the online code. Delete?}
 For example, to scan a phenotype named \texttt{p1}, we run:
@@
 The $p$-value, which is comparable between different types of tests, transforms the LOD score to take account of the number of parameters being fit: it is calculated from the asymptotic distribution of $2\log_e\left(\text{LR}\right)$ under the null model, namely the $\chi^2$ distribution with degrees of freedom equal to the difference in the number of parameters between the alternative and null models.
-% With this example dataset, it takes five seconds to run one genome scan on a Intel Core i5.
 
@@
 These FWER-corrected $p$-values are straightforwardly interpretable: $p = 0.05$ for a specific test at a specific locus implies that in 5\% of similar experiments where there is no true genotype-phenotype association, we would expect to observe some locus with this much or more evidence of association in this test.
-% Additionally, the returned object contains a list of the per-genome-scan maximum observed LOD for each test and each chromosome type.
 
@@
 Calling \texttt{summary} on the output of \texttt{scanonevar} produces a summary of how the scan was conducted and what the results were.
-
-% The details of the null and alternative models used in each of the three tests can be found in the companion article [CompanionC].
-
-% The object returned by the \texttt{scanonevar} function has class \texttt{scanonevar}.
-% Calling \texttt{plot} on this object produces a publication-quality plot that shows the three association statistics at each locus.
-
-% \subsection*{The LOD Score --- Problems and an Alternative}
-% \WV{}{Edits to here so far}
-% QTL mapping in model organisms is traditionally reported via the LOD score.
-However, for reporting the results of mean-variance QTL mapping, the LOD score is cumbersome because it is not comparable across tests with different degrees of freedom; in this case, the mvQTL test has twice as many degrees of freedom as the mQTL or vQTL tests, meaning that each type of test needs to be presented with its own threshold.
-
-% (2) Another problem with the LOD score is that it relates only to a single locus.
-% Geneticists must take into account the many hypotheses that are tested.
-% When a significance threshold is established, \eg by permutation, to control family-wise error rate, it divides
-
-% (1) It must be interpreted differently in autosomes and sex chromosomes.
-% The sex chromosomes are typically fit with fewer parameters, depressing the expected value of the LOD scores under the null, and thus increasing the significance of any observed LOD score.
-% (2) It's interpretation depends on which test was conducted -- mQTLLOD scores from different tests must be interpreted differently.
-% For example, the LOD score of the mvQTL test is always higher than the LOD score of the mQTL and vQTL tests.
-% This relationship is due to the nested nature of the mvQTL test and the other two tests, where they all share an alternative model and the mvQTL test uses a null model that combines the restrictions of the mQTL null and vQTL null.
-% (3) It relates only to a single locus.
-
-% All three tests use the same alternative model, but the null model in the mvQTL test imposes all the constraints of the null of the mQTL test and all the constraints of the null of the vQTL test, so the LOD score of the mvQTL test must, by definition, be greater than the LOD score of the other tests.
-
-% These difficulties in interpreting the LOD score are evident in \autoref{fig:lod_score_scans}.
-% It seems that there are no important signals in the genome scan of \texttt{phenotype1} and it is visually clear that the most interesting signals for \texttt{phenotype2}, \texttt{phenotype3}, and \texttt{phenotype4} are on chromosomes one, two, and three, respectively.
-% But important questions remain:
-% (1) If we didn't have the genome scans from \texttt{phenotype2} -- \texttt{phenotype4} available for comparison, would we be confident there are no statistically significant signals related to \texttt{phenotype1}?
-% (2) How can we compare the results of the mvQTL test to the results of the other tests?
-% (3) How could we compare the results of tests on autosomes to tests on sex chromosomes (if they were present)?
-% (4) How often do we expect to observe results of this magnitude or greater when there is no true association, due simply to sampling variation and the multiplicity of tests conducted?
-
-% To put all genetic loci and all three tests on a level playing field, we consider two types of $p$-values:
-% (1) Asymptotic $p$-values are calculated by \texttt{scanonevar} using the $\chi^2$ distribution with the appropriate degrees of freedom for each locus.
-% Though these $p$-values overcome the questions 1 through 3 of working with LOD scores described above, they leave question 4 unresolved.
-% (2) Empirical, family-wide error rate (FWER)-corrected $p$-values resolve all the above questions with interpreting LOD scores so they are our preferred association statistics in most cases.
-% % thus they are the recommended means of assessing the significance of QTL mapping results.
-% % Statistical methods for empirical FWER control in mean-variance QTL mapping are described in \CortyMethodsPaper.
-
-%\section*{Assess the Significance of Results}
-
-% The $p$-value assigned to each locus by calculating the position of its calculated LOD score in its null distribution answers the question, ``How probable is it to observe this much of a deviation from the null at this locus, given that there is no true effect?''
-% This question is not entirely relevant for genetic mapping, however, because we typically test for association at many loci, with no special interest in any individual locus \textit{a priori}.
-% Thus, a more appropriate $p$-value would answer the question, ``How probable is it to observe this much or more deviation from the null at \textit{some} locus in the genome, given that there are no true effects?''
-% This is precisely the reasoning behind the family-wide-error-rate (FWER) controlling procedure we illustrate below.
-
-% To calculate a FWER-controlling $p$-value, one typically must estimate the effective number of statistical tests conducted.
-% The effective number of tests in a genome scan, however, is difficult to estimate.
-% One lower bound is the number of chromosomes.
-% Due to the randomization in meiosis no two non-syntenic loci are correlated in an experimental cross and therefore tests on different chromosomes are always independent.
-% But there are many non-identical tests conducted on each chromosome, so the number of chromosomes is an under-estimate.
-% One upper bound on the effective number of tests is the total number of loci.
-% But, loci on the same chromosome are often in linkage disequilibrium, therefore the tests are not independent and the total number of loci is an over-estimate \citep{Lander1989a}.
 
@@
 
-% The identification of an mQTL is consistent, roughly speaking, with horizontal distances between genotype groups that exceed the horizontal standard error bars.
-% The identification of a vQTL is consistent with vertical distances between genotype groups that exceed the vertical standard error bars.
-% And the identification of an mvQTL is consistent with euclidean distance between genotype groups that exceed the standard error density implied by both the horizontal and vertical standard errors.
-
 Additional plotting utilities, \texttt{phenotype\_plot}, \texttt{effects\_plot} and \texttt{mean\_var\_plot\_model\_free} are described in the online documentation, available on CRAN.
@@
 With the datasets simulated here, it takes 20 minutes to run 1000 bootstrap resamples on an Intel core i5.
-
-% \section*{Mathematical Details}
-
-% [todo]
-
-% mean-var plot what is on x what is on y and how are SE's calculated
-
-% DGLM with other links and distribs
 
@@
 
-% Performance benchmarks of package \texttt{vqtl} are presented in \autoref{fig:benchmark_kumar} and \autoref{fig:benchmark_sim_cross}.
-
-% \begin{table}
-%     \centering
-%     \begin{tabular}{r|rrrr}
-%         \multirow{2}{1in}{\raggedleft \textbf{number of permutations}}& \multicolumn{4}{c}{\textbf{number of cores}}\\
-%         \cline{2-5}
-%         & 4 & 8 & 16 & 32\\
-%         \hline
-%         100  & 11 & 6 & 4 & 4\\
-%         1000 & -  & 53 & 34 & 20\\
-%         \hline
-%     \end{tabular}
-%     \caption{
-%         Time in minutes required to conduct permuations for all three tests on the dataset from \citet{Kumar2013}, which has 244 individuals and 582 loci, according to number of permutations desired and number of computer cores available.
-%         We recommend 1000 permutations for publication-quality precision.
-%         We include the time taken to conduct 100 and 500 permutations to illustrate the linear relation between number of permutations and time required.
-%         This table also illustrates the linear relationship between number of computer cores used and time taken to complete permutations.
-%     }
-% \end{table}
-%
-% \begin{table}
-%     \centering
-%     \begin{tabular}{r|rrrr}
-%         \multirow{2}{1in}{\raggedleft \textbf{number of markers}}& \multicolumn{4}{c}{\textbf{number of individuals}}\\
-%         \cline{2-5}
-%         & 100 & 200 & 400 & 800\\
-%         \hline
-%         100 & 2 & 3 & 4 & 4\\
-%         200 & 5 & 5 & 7 & 9\\
-%         400 & 9 & 11 & 13 & 19\\
-%         800 & 19 & 23 & 31 & 46\\
-%         \hline
-%     \end{tabular}
-%     \caption{Time in minutes required to conduct 1000 permuations on 32 cores for all three tests on a simulated F2 intercross, according to number of individuals in the mapping population and number of markers in the genome.}
-% \end{table}
-
 \section*{Conclusion}
@@
 The package \texttt{vqtl} and its documentation are freely available on \texttt{CRAN} at \url{https://CRAN.R-project.org/package=vqtl}.
-
-% The central function of this package, \texttt{scanonevar}, carries out the initial genome scan.
-% The permutation function, \texttt{scanonevar.perm}, executes a set of permutations to empirically estimate a FWER-controlling $p$-value for each observed LOD score.
-% A suite of plotting functions, e.g. the \texttt{mean\_var\_plot}, allows geneticists to evaluate the allelic and covariate effects that underlie detected signals.
-% The bootstrap function, \texttt{scanonevar.boot}, executes bootstrap resampling to assess the genetic precision of a QTL and extablish a confidence interval.
 

### 2018-10-30 Package_vqt_Manuscript.tex
fixed bib

@@
 Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
-It has long been recognized, however, that the residual variance is itself heritable \citep{Falconer1965,Lynch1998}, a
-phenomenon that has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
-%To help dissect the genetic basis, 
-Correspondingly, several groups have proposed statistical methods for mapping QTL controlling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014,Soave2017,Dumitrascu2018}.
-However, although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
+It has long been recognized, however, that the residual variance is itself heritable \citep{Falconer1965,Lynch1998a}, a
+phenomenon that has been described theoretically \citep{Hill2004,Hill2010}, demonstrated in inbred model organisms \citep{Sorensen2015} and crops \citep{Yang2012}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008a}.
+Correspondingly, several groups have proposed statistical methods for mapping QTL controlling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012a,Cao2014,Soave2017,Dumitrascu2018}.
+However, although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016,Wang2017,Wei2017}, statistical tools for this purpose remain heterogeneous.
 
@@
 statistically, techniques could be used that allow for large-scale analysis without permutation testing \citep{Efron2004};
-computationally, the software could be modified to run on a computer cluster, rather than on a single computer \citep{Jette2003a,Marchand2017}.
+computationally, the software could be modified to run on a computer cluster, rather than on a single computer \citep{Jette2003,Marchand2017}.
 
@@
 
-\bibliography{10_Aim1,wv}
+\bibliography{10_Aim1}
 
